---
title: "adding_chlorophyll_a"
author: "Candela Ferrer Diez"
date: 2025-12-01
format: 
  html:
    toc: true
    toc-depth: 2
    code-fold: false
    theme: cosmo
    embed-resources: true
editor: visual
execute:
  warning: false
  message: false
---

## Overview

This document processes NASA Ocean Color Level-3 binned (L3b) chlorophyll-a data for the Western Indian Ocean (WIO) region. The script:

-   Reads HDF5/NetCDF files containing binned satellite ocean color data
-   Converts NASA bin numbers to geographic coordinates (lat/lon)
-   Filters data to the WIO Area of Interest (37°E - 47°E, 15°S - 1°N)
-   Exports combined data to CSV and Parquet formats

## R Set-up

*Important note: Comprehensive instructions, including all prerequisites and detailed guidelines for running this analysis and relevant code, are available in the README file of my [GitHub repository](...) for this project. Please refer to the README to ensure proper setup and usage.*

### Load Required Packages

```{r}
#| label: load-packages
#| echo: false
#| include: false

#Delete for the final version of the code
library(here)
library(hdf5r)
library(data.table)
library(arrow)
library(sf)
library(data.table)
```

### Restore Package Environment

```{r}
#| label: renv-restore
#| eval: false

renv::restore()
```

## Configuration

```{r}
#| label: config

# Western Indian Ocean Area of Interest (Kenya, Tanzania, Comoros)
coords_list <- list(
  W = 37.0,   # West longitude
  E = 47.0,   # East longitude
  S = -15.0,  # South latitude
  N = 1.0     # North latitude
)
```

### Source Functions

Functions for data processing are sourced from a separate script to keep this document clean and focused on the analysis workflow.
```{r}
#| label: source-functions
source(here::here("functions", "extracting.R"))
```


## File Processing

### Locate Input Files

```{r}
#| label: find-files

# Define path to all the raw chlor a files 
# - returns a vector of the full paths for each chlor a file
filenames <- list.files(
  here::here("raw_data", "chlor_a_raw"),
  full.names = TRUE)

```

```{r}
### Single File Processing Function
#| label: process-file-function

#libraries

#functions 
source(here::here("functions", "extracting.R"))

#| label: process-data

# 1. Apply processing function to all files
# We use lapply to loop over the filenames vector
data_list <- lapply(filenames, function(f) {
  process_single_file(filepath = f, coords_list)
})

# 2. Combine into a single data table
all_data <- data.table::rbindlist(data_list, use.names = TRUE, fill = TRUE)

message(sprintf("Processing complete. Total rows extracted: %d", nrow(all_data)))
```

## Results and Export

```{r}
#| label: save-data


#libraries
library(data.table)

#save csv
fwrite(
  all_data,
  here::here("processed_data", "all_chlor_a_combined.csv")
)


```


# Part 2: Merging with Location Data

## Load Libraries

```{r}

# load location data

location_data <- read.csv(here("processed_data", "clean_location_2018.csv"))

```

#Merge Nearest-bin, same month -\> assigns each station the single closest satellite bin from the same year-month.

```{r}
# 1) convert to sf (lon/lat are numeric columns in your clean file)
sat_sf <- st_as_sf(all_data, coords = c("lon", "lat"), crs = 4326, remove = FALSE)
sta_sf <- st_as_sf(clean_location_2018, coords = c("longitude", "latitude"), crs = 4326, remove = FALSE)

# 2) add year-month for temporal matching
sat_sf$ym <- format(as.Date(sat_sf$date), "%Y-%m")
sta_sf$ym <- format(as.Date(sta_sf$date), "%Y-%m")

# 3) find nearest *within same month*, falling back to global nearest if none
nearest_matches <- integer(nrow(sta_sf))
for (i in seq_len(nrow(sta_sf))) {
  ym_i <- sta_sf$ym[i]
  candidates <- which(sat_sf$ym == ym_i)
  if (length(candidates) == 0) {
    nearest_matches[i] <- st_nearest_feature(sta_sf[i, ], sat_sf)
  } else {
    idx_local <- st_nearest_feature(sta_sf[i, ], sat_sf[candidates, ])
    nearest_matches[i] <- if (!is.na(idx_local)) candidates[idx_local] else st_nearest_feature(sta_sf[i, ], sat_sf)
  }
}

# 4) build merged table
merged_nearest <- cbind(
  as.data.table(clean_location_2018),
  as.data.table(sat_sf)[nearest_matches, .(sat_date = date, sat_chlor_a = chlor_a, sat_nobs = nobs, sat_lon = lon, sat_lat = lat)]
)

# 5) inspect
head(merged_nearest)

```

## Visualization

### Chlorophyll-a Distribution

```{r}
#| label: plot-chlor-distribution
#| fig-width: 10
#| fig-height: 6

ggplot(merged_nearest %>% filter(!is.na(sat_chlor_a)), 
       aes(x = sat_chlor_a)) +
  geom_histogram(bins = 50, fill = "steelblue", alpha = 0.7) +
  scale_x_log10() +
  labs(
    title = "Distribution of Chlorophyll-a at Survey Locations",
    subtitle = "Monthly averages matched by location and month",
    x = "Chlorophyll-a (mg/m³, log scale)",
    y = "Count"
  ) +
  theme_minimal()
```

### Spatial Distribution

```{r}
#| label: plot-spatial
#| fig-width: 10
#| fig-height: 8

ggplot(merged_nearest %>% filter(!is.na(sat_chlor_a)), 
       aes(x = longitude, y = latitude, color = sat_chlor_a)) +
  geom_point(alpha = 0.6, size = 2) +
  scale_color_viridis_c(trans = "log10", option = "viridis") +
  coord_fixed(ratio = 1) +
  labs(
    title = "Spatial Distribution of Chlorophyll-a Values",
    subtitle = "Survey locations with matched monthly chlorophyll-a data",
    x = "Longitude",
    y = "Latitude",
    color = "Chlorophyll-a\n(mg/m³)"
  ) +
  theme_minimal()
```

### Match Quality

```{r}
#| label: plot-match-quality
#| fig-width: 10
#| fig-height: 6

location_with_chlor %>%
  filter(!is.na(chlor_a_distance_km)) %>%
  ggplot(aes(x = chlor_a_distance_km)) +
  geom_histogram(bins = 30, fill = "coral", alpha = 0.7) +
  labs(
    title = "Distribution of Spatial Match Distances",
    subtitle = "Distance between survey location and nearest chlorophyll-a grid cell",
    x = "Distance (km)",
    y = "Count"
  ) +
  theme_minimal()
```

## Final Summary

```{r}
#| label: final-summary

cat("\n=== FINAL SUMMARY ===\n\n")

cat("Chlorophyll-a Processing:\n")
cat(sprintf("  Files processed: %d\n", length(unique(chlor_df$source_file))))
cat(sprintf("  Total chlorophyll observations: %s\n", 
            format(nrow(chlor_df), big.mark = ",")))
cat(sprintf("  Months covered: %d\n", n_distinct(chlor_df$year_month)))

cat("\nSpatial Matching (by month):\n")
cat(sprintf("  Survey locations: %s\n", 
            format(nrow(location_with_chlor), big.mark = ",")))
cat(sprintf("  Successful matches: %s (%.1f%%)\n",
            format(sum(!is.na(location_with_chlor$chlor_a)), big.mark = ","),
            100 * mean(!is.na(location_with_chlor$chlor_a))))

if (sum(!is.na(location_with_chlor$chlor_a)) > 0) {
  cat("\nChlorophyll-a Statistics (matched points):\n")
  cat(sprintf("  Mean: %.4f mg/m³\n", 
              mean(location_with_chlor$chlor_a, na.rm = TRUE)))
  cat(sprintf("  Median: %.4f mg/m³\n", 
              median(location_with_chlor$chlor_a, na.rm = TRUE)))
  cat(sprintf("  Range: %.4f - %.4f mg/m³\n",
              min(location_with_chlor$chlor_a, na.rm = TRUE),
              max(location_with_chlor$chlor_a, na.rm = TRUE)))
  
  cat("\nMatch Quality:\n")
  cat(sprintf("  Mean spatial distance: %.2f km\n",
              mean(location_with_chlor$chlor_a_distance_km, na.rm = TRUE)))
  cat(sprintf("  Median spatial distance: %.2f km\n",
              median(location_with_chlor$chlor_a_distance_km, na.rm = TRUE)))
}
```
