---
title: "additiong_gravity_data"
author: "Candela Ferrer Diez"
date: 2025-11-06
format: 
  html:
    toc: true
    toc-depth: 2
    code-fold: false
    theme: cosmo
    embed-resources: true
editor: visual
execute:
  warning: false
  message: false
---
## Overview


## R Set-up

Load the necessary packages (these can be found in ...)

*Important note: Comprehensive instructions, including all prerequisites and detailed guidelines for running this analysis and relevant code, are available in the README file of my [GitHub repository](...) for this project. Please refer to the README ... to ensure proper setup and usage.*

Use the renv::restore() to load up the necessary package versions along with the necessary dependencies.

```{r}
#| label: r set-up
#| echo: false
#| include: false

#Load packages -> BAD CODE (delete once renv snapshot)
library(sf)           # For spatial data handling
library(dplyr)        # For data manipulation
library(terra)        # Alternative for raster operations (if needed)
library(here)
library(dplyr)
library(ggplot2)
```

```{r}
#| label: renv-restore
renv::restore()
```

# Adding Human Gravity

## Location data

This will added to a separte file including lcoation, station, site,

### Load cleaned data

...

```{r}
#| label: load location data
#| echo: false
#| warning: false

#load libraries
library(readr)

##load location data
clean_location_2018 <- read.csv(here::here("processed_data","clean_location_2018.csv"))

```

### Convert to Spatial Format

Transform the location data into a spatial object (simple features) to enable spatial operations and joins with the gravity data.

```{r}
#| label: convert-to-spatial
#| echo: false
#| warning: false

#load libraries
library(sf)


# Convert to sf object with WGS84 coordinate system
sites_sf <- st_as_sf(clean_location_2018, 
                     coords = c("longitude", "latitude"),
                     crs = 4326,  # WGS84
                     remove = FALSE)  # Keep original lon/lat columns

```

## Gravity Data

### Load Gravity Data

Load the human gravity shapefile which contains modeled human pressure on coral reefs based on population density and accessibility within a 500 km radius. ... (more about this)

```{r}
#| label: load-gravity-data
#| echo: false
#| warning: false

#load libraries
library(sf)
library(dplyr)

# Load the shapefile
gravity_data <- st_read(here::here("raw_data", "Total Gravity of Coral Reefs 1.0.shp"))

# Check the structure
glimpse(gravity_data)

```

### Clean Gravity Data

Standardise column names to match the cleaned data

```{r}
#| label: clean-gravity-data
#| echo: false
#| warning: false

#load libraries 
library(janitor)

# Source data cleaning function
source(here::here("functions", "cleaning.R"))

# Clean gravity data
clean_gravity_data <- clean_column_names(gravity_data)

```

### Extract Gravity Values

Match each survey site to the nearest gravity grid cell and extract the human pressure values. I will be doing this through a spatial join by finding the nearest gravity grid cell for each cell - For each site in sites_sf, find the nearest polygon in gravity_data and copy all its attributes (reef_id, grav_tot) to that site's row

```{r}
#| label: extract-gravity
#| echo: false
#| warning: false

library(sf)
library(dplyr)

sf_use_s2(FALSE)

sites_with_gravity <- sites_sf %>%
  
  # compute index of the nearest gravity cell for each site
  mutate(nearest_idx = st_nearest_feature(., clean_gravity_data)) %>%
  
  # bind the gravity columns by indexing the gravity table with nearest_idx
  {
    # grab grav columns from gravity table as a plain data.frame/tibble
    grav_df <- st_drop_geometry(clean_gravity_data) %>% dplyr::select(contains("grav"))
    # bind the matching rows (use drop = FALSE to preserve columns when single column)
    bind_cols(., grav_df[.$nearest_idx, , drop = FALSE])
  } %>%
  
  # compute distance once using the stored index
  mutate(distance_to_grid_km = as.numeric(
    st_distance(., clean_gravity_data[nearest_idx, ], by_element = TRUE)
  ) / 1000) %>%
  
  # drop geometry and keep only the desired columns (nearest_idx will be removed)
  st_drop_geometry() %>%
  dplyr::select(
    date, station, site, location, country,
    latitude, longitude,
    contains("grav"),
    distance_to_grid_km
  )

```

### Checking results

Convert the spatial object back to a regular dataframe and organize columns for easier interpretation.

```{r}
#| label: check-results
#| echo: false
#| warning: false

library(dplyr)
library(sf)

local({
  # local scope - prevents unnecessary objects that clutter the global environment
  missing_grav <- sites_with_gravity %>% filter(is.na(grav_tot))

  if (nrow(missing_grav) > 0) {
    cat("⚠ Warning:", nrow(missing_grav), "sites missing gravity values\n")
    # drop geometry for nicer printing and only show the key columns
    print(missing_grav %>% st_drop_geometry() %>% select(station, site, location))
  } else {
    cat("All sites successfully matched with gravity values\n")
  }

  invisible(NULL)  # nothing returned to the parent environment
})

```

# Save data

...

```{r}

#| label: export-results
#| echo: false
#| warning: false

#load libraries
library(dplyr)
library(sf)

# Export results
write.csv(sites_with_gravity, 
          "sites_with_gravity_2018.csv", 
          row.names = FALSE)

# Export summaries
# write.csv(country_summary, 
#         "gravity_summary_by_country.csv", 
#          row.names = FALSE)

# write.csv(location_summary, 
#          "gravity_summary_by_location.csv", 
#          row.rows = FALSE)

# Save as spatial file if needed for GIS
st_write(sites_with_gravity, 
         "sites_with_gravity.shp", 
         delete_dsn = TRUE)  # Overwrite if exists

print("\n✓ All results exported successfully!")
print("Files created:")
print("  - clean_location_2018_with_gravity.csv")
print("  - gravity_summary_by_country.csv")
print("  - gravity_summary_by_location.csv")
print("  - sites_with_gravity.shp (and associated files)")


```
