---
title: "adding_chlorophyll_a"
author: "Candela Ferrer Diez"
date: 2025-12-01
format: html
editor: visual
---

## Overview

This document processes NASA Ocean Color Level-3 binned (L3b) chlorophyll-a data.. (and sst data) for the Western Indian Ocean (WIO) region. The script:

-   Reads HDF5/NetCDF files containing binned satellite ocean color data
-   Converts NASA bin numbers to geographic coordinates (lat/lon)
-   Filters data to the WIO Area of Interest (37°E - 47°E, 15°S - 1°N)
-   Exports combined data to CSV and Parquet formats

## R Set-up

*Important note: Comprehensive instructions, including all prerequisites and detailed guidelines for running this analysis and relevant code, are available in the README file of my [GitHub repository](...) for this project. Please refer to the README to ensure proper setup and usage.*

### Load Required Packages (delete when clean)

```{r}
#| label: load-packages
#| echo: false
#| include: false

#Delete for the final version of the code
library(here)
library(hdf5r)
library(data.table)
library(arrow)
library(sf)
library(tidyverse) 
library(dplyr)      
library(readr)

```

renv::restore() loads the necessary package versions along with the necessary dependencies. Information about this can be found in the lockfile (\[where...\] in the Github repository).

```{r}
#| label: renv-restore
renv::restore()
```

### Set WIO box

This defines the region of interest to filter the extracted environmental data.

```{r}
#| label: define-WIO-box

# Western Indian Ocean Area of Interest (Kenya, Tanzania, Comoros)
coords_list <- list(
  W = 37.0,   # West longitude
  E = 47.0,   # East longitude
  S = -15.0,  # South latitude
  N = 1.0     # North latitude
)
```

## Chlor a data for 2009

### Define path to raw chlor a files for 2009

```{r}
#| label: find-files-2009

# Define path to all the raw chlor a files 
filenames_2009 <- list.files(
  here::here("raw_data", "chlor_a_raw_2009"),
  full.names = TRUE)
```

### Extract raw chlor a data for 2009 from NASA L3b files

This extracts the relevant data from each of the NASA L3b files and combines them into a single data table of raw chlorophyll data for ease of use. It extracts: - chlorophyll a monthly mean values - number of observations - latitude and longitude values for each satellite cell - each satellite cell id

*See extracting.r document in functions folder to find the data extraction functions*

```{r}
#| label: process-chlor-a-files-2009

# Source structure exploration function
source(here::here("functions", "extracting.R"))

# Libraries
# ...

# Begin by processing the first file only - will use later this to get unique cell ids
  # but is this really the way to do it|? 
  # may need to change the matching process - think about this...
first_measurements_2009 <- extract_raw_data(filenames_2009[1], coords_list)

# Process all raw files using the processing function
data_list_2009 <- lapply(filenames_2009, function(f) {
  extract_raw_data(filepath = f, coords_list)
})

# Combine into a single data table
raw_chlor_a_2009 <- data.table::rbindlist(data_list_2009, use.names = TRUE, fill = TRUE)

message(sprintf("Processing complete. Total rows extracted: %d", nrow(raw_chlor_a_2009)))
```

#### Save raw chlor-a data for 2009

```{r}
#| label: save-chlor-a-data-2009

# Libraries
library(data.table)

# Save csv
fwrite(raw_chlor_a_2009, here::here("raw_data", "raw_chlor_a_2009.csv"))
```

### Match raw chlor-a data to each station in 2009-2016 data set

#### Load 2009-2016 location data and empty columsn to add chor-a data values

```{r}
#| label: load-2009-location-data

# Libraries 
library(readr)

# Load location data and add columns to add relevant chlor-a data
clean_location_2009 <- read_rds(
  here::here("processed_data", "clean_location_2009.rds")
) %>% 
  dplyr::mutate(closest_cell = 0, cell_distance_m = 0, chlor_a_annual_mean = 0)

# maybe also remove unnecessary columns from location data set... 

```

#### Find nearest satellite cells of chlor-a values to 2009 stations

\[explain what this chuck is doing exactly...\]

... make this into a function...

```{r}
#| label: find-nearest-satellite-cells-2009
 
# Libraries
library(sf)
library(dplyr)
library(units)

# convert to sf (assumes columns as specified)
stations_as_sf_2009  <- st_as_sf(clean_location_2009, coords = c("longitude", "latitude"), crs = 4326, remove = FALSE)
first_measurements_as_sf_2009  <- st_as_sf(first_measurements_2009, coords = c("lon", "lat"),            crs = 4326, remove = FALSE)

# find index of nearest feature in first_sf for each point in clean_sf
nearest_indices_2009 <- st_nearest_feature(stations_as_sf_2009, first_measurements_as_sf_2009)
# nearest_idx[i] is the row index in first_sf closest to clean_sf[i,]

# extract corresponding bin id
stations_as_sf_2009$closest_cell <- first_measurements_2009$bin_num[nearest_indices_2009]

# compute accurate geodetic distances (in meters) for those pairs
# st_distance on geographic coordinates uses s2/geodetic when available; specify by_element=TRUE
dists_2009 <- st_distance(stations_as_sf_2009, first_measurements_as_sf_2009[nearest_indices_2009,], by_element = TRUE)

# convert to numeric (in meters)
stations_as_sf_2009$cell_distance_m <- as.numeric(set_units(dists_2009, "m"))

# put back into the original dataframe if needed
clean_location_2009$closest_cell   <- stations_as_sf_2009$closest_cell
clean_location_2009$cell_distance_m  <- stations_as_sf_2009$cell_distance_m

```

### Adding annual chlorophyll a means

```{r}
# Adding annual chlorophyll a means
for (i in 1:length(clean_location_2009$station)){
  #get month and year and id
  date <- clean_location_2009$date[i]
  closest_cell = clean_location_2009$closest_cell[i]

  # start = first day of the month 11 months before the reference month
  start_month <- floor_date(date, "month") %m-% months(11)
  
  # end = last day of the reference month
  end_month <- ceiling_date(date, "month") - days(1)
  
  relevant_cells <- raw_chlor_a_2009 %>%
    filter(
      bin_num == closest_cell,
      date >= start_month, # Sort to get relevant date (previous year)
      date <= end_month    # ^
    ) 
  
  #find their chlorophyll average
  weighted_mean <- with(relevant_cells, sum(chlor_a_mean * nobs, na.rm = TRUE) / sum(nobs, na.rm = TRUE))

  
  #add it to the table
  clean_location_2009$chlor_a_annual_mean[i] = weighted_mean

}

```

Things to do still: 
- Need to rewrite part about matching to the cell 
- Separate the necessary functions (if needed)
- inspect the data

## Chlor a data for 2018

### Define path to raw chlor a files for 2018

```{r}
#| label: find-files-2018

# Define path to all the raw chlor a files 
filenames_aqua_2018 <- list.files(
  here::here("raw_data", "chlor_a_raw_aqua_2018"),
  full.names = TRUE)

#ocnet
filenames_ocnet_2018 <- list.files(
  here::here("raw_data", "ocnet_chla_data"),
  full.names = TRUE)
```

### Extract raw chlor a data for 2018 from NASA L3b files

*See extracting.r document in functions folder to find the data extraction functions*

```{r}
#| label: process-chlor-a-files-2018

# Source structure exploration function
source(here::here("functions", "extracting.R"))

# Libraries
# ...

# Process all raw files using the processing function
data_list_2018 <- lapply(filenames_aqua_2018, function(f) {
  extract_raw_data(filepath = f, coords_list)
})

coords_list_smaller <- list(
  W = 35.0,   # West longitude
  E = 44.0,   # East longitude
  S = -15.0,  # South latitude
  N = 1.0     # North latitude
)
path <- "C:/KenyaSSFProject/raw_data/ocnet_chla_data/OCNET_chla_2015.nc"

data_list_2018_ocnet <- extract_latlon_ocnet(path, coords_list_smaller)



# Combine into a single data table
raw_chlor_a_2018 <- data.table::rbindlist(data_list_2018, use.names = TRUE, fill = TRUE)


message(sprintf("Processing complete. Total rows extracted: %d", nrow(raw_chlor_a_2018)))
```

#### Save raw chlor-a data for 2018

```{r}
#| label: save-chlor-a-data-2018

# Libraries
library(data.table)

# Save csv
fwrite(raw_chlor_a_2018, here::here("raw_data", "raw_chlor_a_2018.csv"))
```

### Match raw chlor-a data to each station in 2018-2016 data set

#### Load 2018-2016 location data and empty columsn to add chor-a data values

```{r}
#| label: load-2018-location-data

# Libraries 
library(readr)

# Load location data and add columns to add relevant chlor-a data
clean_location_2018 <- read_rds(
  here::here("processed_data", "clean_location_2018.rds")
) %>% 
  dplyr::select(date, station, site, country, latitude, longitude)%>% 
  dplyr::mutate(closest_cell = 0, cell_distance_m = 0, chlor_a_annual_mean = 0) 

# maybe also remove unnecessary columns from location data set... 

```

#### Find nearest satellite cells of chlor-a values to 2018 stations

\[explain what this chunk is doing exactly...\]

... make this into a function...

```{r}
#| label: find-nearest-satellite-cells-2018

# Source structure exploration function
source(here::here("functions", "extracting.R"))

# Libraries
library(sf)
library(dplyr)
library(units)

station_with_sat_2018 <- assign_nearest_cell(raw_chlor_a_2018, clean_location_2018)

```

# Adding annual chlorophyll a means

```{r}

# Source structure exploration function
source(here::here("functions", "extracting.R"))


# Adding annual chlorophyll a means
stations_with_annual_chla <- compute_annual_mean(station_with_sat_2018, raw_chlor_a_2018, 'chlor_a_mean', 'chlor_a_annual_mean')

```

## Visualise on a map

```{r}

library(sf)
library(ggplot2)
library(dplyr)
library(rnaturalearth)
library(rnaturalearthdata)


world <- ne_countries(scale = "medium", returnclass = "sf")
focus_region <- world %>%
  filter(admin %in% c("Kenya", "Tanzania", "Comoros"))

# Create the map with a tighter zoom
map_plot<- ggplot() +
  
  # Plot all countries in Africa as the base layer
  geom_sf(data = world[world$continent == "Africa",], 
          fill = "gray95", color = "gray50", size = 0.3) + 
  
  geom_sf(data = focus_region, fill = "lightblue", color = "gray30", size = 0.3) +
  
  geom_point(data = clean_location_2018,
             aes(x = longitude, y = latitude, ),
             alpha = 0.6, color = "#440154") +
            # each country gets a different 
  geom_point(data = data_list_2018_ocnet, 
             aes(x = lon, y = lat),
             alpha = 0.6, color = "pink")+
  geom_point(data = raw_chlor_a_2018, 
             aes(x = lon, y = lat),
             alpha = 0.6, color = "darkblue")+
  
  
  scale_size_continuous(name = "Number of\nObservations", 
                        range = c(0.5, 5),
                        breaks = c(10, 25, 50, 75, 100, 150, 200)) +
  
  # Specify region of focus - Kenya in frame
  coord_sf(xlim = c(37, 46.5), ylim = c(1, -15)) + 
  
  labs(title = "Marine Survey Sites in Kenya",
       subtitle = "Point size represents number of fish observations per site",
       x = "Longitude", y = "Latitude") +
  
  theme_minimal() +
  theme(panel.grid.major = element_line(color = "gray90", size = 0.2),
        panel.background = element_rect(fill = "aliceblue"),
        plot.title = element_text(face = "bold", size = 14),
        legend.position = "right")

print(map_plot)

```

### Match Quality

```{r}
#| label: plot-match-quality
#| fig-width: 10
#| fig-height: 6

location_with_chlor %>%
  filter(!is.na(chlor_a_distance_km)) %>%
  ggplot(aes(x = chlor_a_distance_km)) +
  geom_histogram(bins = 30, fill = "coral", alpha = 0.7) +
  labs(
    title = "Distribution of Spatial Match Distances",
    subtitle = "Distance between survey location and nearest chlorophyll-a grid cell",
    x = "Distance (km)",
    y = "Count"
  ) +
  theme_minimal()
```

## General summary
