---
title: "adding_gravity_data"
author: "Candela Ferrer Diez"
date: 2025-11-06
format: 
  html:
    toc: true
    toc-depth: 2
    code-fold: false
    theme: cosmo
    embed-resources: true
editor: visual
execute:
  warning: false
  message: false
---

# Overview

...

# R Set-up

*Important note: Comprehensive instructions, including all prerequisites and detailed guidelines for running this analysis and relevant code, are available in the README file of my [GitHub repository](...) for this project. Please refer to the README ... to ensure proper setup and usage.*

## Restore ...

Use the renv::restore() to load up the necessary package versions along with the necessary dependencies.

```{r}
#| label: renv-restore
renv::restore()
```

## Load the necessary packages (these can be found in ...)

```{r}
#| label: r set-up
#| echo: false
#| include: false

#Load required libraries
library(sf)           # For spatial data handling
library(dplyr)        # For data manipulation
library(terra)        # Alternative for raster operations (if needed)
library(here)
library(dplyr)
library(ggplot2)
library(readr)
```

## Import required functions

I wonder if from the extracting documents I can only extract the necessary functions (not the whole doc) Or maybe I should create a separate extraction doc for human gravity

```{r}
#| label: load-functions
#| echo: false
#| include: false

# Import data extraction functions
source(here::here("functions", "extracting.R"))

```

# Adding Human Gravity

Human gravity is ...

## 2009 dataset

### Location data

...

```{r}
#| label: load-2009-location-data

# Load location data and add columns to add relevant chlor-a data
locations_2009 <- read_rds(
  here::here("processed_data", "clean_location_2009.rds")) %>% 
  dplyr::select(date, station, site, country, latitude, longitude) 
#%>% 
 # dplyr::mutate(closest_cell_lat =0, closest_cell_lon=0, cell_distance_km = 0, chla_annual_mean = 0)
```

### Convert to Spatial Format

Transform the location data into a spatial object (simple features) to enable spatial operations and joins with the gravity data.

```{r}
#| label: locations-to-sf-2009
#| echo: false
#| warning: false

# Convert to sf object with WGS84 coordinate system
locations_sf <- st_as_sf(locations_2009, 
                     coords = c("longitude", "latitude"),
                     crs = 4326,  # WGS84
                     remove = FALSE)  # Keep original lon/lat columns

```

## Gravity Data

### Load Gravity Data

Load the human gravity shapefile which contains modeled human pressure on coral reefs based on population density and accessibility within a 500 km radius. ... (more about this)

```{r}
#| label: load-gravity-data
#| echo: false
#| warning: false

# Load the human gravity shapefile
gravity_data <- st_read(here::here("raw_data", "Total Gravity of Coral Reefs 1.0.shp"))

```

### Extract Gravity Values

Match each survey site to the nearest gravity grid cell and extract the human pressure values. I will be doing this through a spatial join by finding the nearest gravity grid cell for each cell - For each site in sites_sf, find the nearest polygon in gravity_data and copy all its attributes (reef_id, grav_tot) to that site's row

```{r}
#| label: extract-gravity
#| echo: false
#| warning: false

sf_use_s2(FALSE)

locations_with_gravity <- locations_sf %>%
  
  # compute index of the nearest gravity cell for each site
  mutate(nearest_idx = st_nearest_feature(., gravity_data)) %>%
  
  # bind the gravity columns by indexing the gravity table with nearest_idx
  {
    # grab grav columns from gravity table as a plain data.frame/tibble
    grav_df <- st_drop_geometry(gravity_data) %>% dplyr::select(contains("Grav"))
    # bind the matching rows (use drop = FALSE to preserve columns when single column)
    bind_cols(., grav_df[.$nearest_idx, , drop = FALSE])
  } %>%
  
  # compute distance once using the stored index
  mutate(distance_to_grid_km = as.numeric(
    st_distance(., gravity_data[nearest_idx, ], by_element = TRUE)
  ) / 1000) %>%
  
  # drop geometry and keep only the desired columns (nearest_idx will be removed)
  st_drop_geometry() %>%
  dplyr::select(
    date, station, site, country,
    latitude, longitude,
    contains("Grav"),
    distance_to_grid_km
  )

```

### Checking results

Convert the spatial object back to a regular dataframe and organize columns for easier interpretation.

```{r}
#| label: check-results
#| echo: false
#| warning: false


local({
  # local scope - prevents unnecessary objects that clutter the global environment
  missing_grav <- locations_with_gravity %>% filter(is.na(Grav_tot))

  if (nrow(missing_grav) > 0) {
    cat("⚠ Warning:", nrow(missing_grav), "sites missing gravity values\n")
    # drop geometry for nicer printing and only show the key columns
    print(missing_grav %>% st_drop_geometry() %>% select(station, site, location))
  } else {
    cat("All sites successfully matched with gravity values\n")
  }

  invisible(NULL)  # nothing returned to the parent environment
})

```

# Save data

...

```{r}

#| label: export-results
#| echo: false
#| warning: false

#load libraries
library(dplyr)
library(sf)


# Save csv
write.csv(locations_with_gravity , here::here("processed_data", "locations_with_gravity.csv"))


# Export summaries
# write.csv(country_summary, 
#         "gravity_summary_by_country.csv", 
#          row.names = FALSE)

# write.csv(location_summary, 
#          "gravity_summary_by_location.csv", 
#          row.rows = FALSE)

# Save as spatial file if needed for GIS
# st_write(sites_with_gravity, 
#          "sites_with_gravity.shp", 
#         delete_dsn = TRUE)  # Overwrite if exists
```

## 2018 dataset

### Location data

...

```{r}
#| label: load-2009-location-data

# Load location data and add columns to add relevant chlor-a data
locations_2009 <- read_rds(
  here::here("processed_data", "clean_location_2009.rds")) %>% 
  dplyr::select(date, station, site, country, latitude, longitude) 
#%>% 
 # dplyr::mutate(closest_cell_lat =0, closest_cell_lon=0, cell_distance_km = 0, chla_annual_mean = 0)
```

### Convert to Spatial Format

Transform the location data into a spatial object (simple features) to enable spatial operations and joins with the gravity data.

```{r}
#| label: locations-to-sf-2009
#| echo: false
#| warning: false

# Convert to sf object with WGS84 coordinate system
locations_sf <- st_as_sf(locations_2009, 
                     coords = c("longitude", "latitude"),
                     crs = 4326,  # WGS84
                     remove = FALSE)  # Keep original lon/lat columns

```

## Gravity Data

### Load Gravity Data

Load the human gravity shapefile which contains modeled human pressure on coral reefs based on population density and accessibility within a 500 km radius. ... (more about this)

```{r}
#| label: load-gravity-data
#| echo: false
#| warning: false

# Load the human gravity shapefile
gravity_data <- st_read(here::here("raw_data", "Total Gravity of Coral Reefs 1.0.shp"))

```

### Extract Gravity Values

Match each survey site to the nearest gravity grid cell and extract the human pressure values. I will be doing this through a spatial join by finding the nearest gravity grid cell for each cell - For each site in sites_sf, find the nearest polygon in gravity_data and copy all its attributes (reef_id, grav_tot) to that site's row

```{r}
#| label: extract-gravity
#| echo: false
#| warning: false

sf_use_s2(FALSE)

locations_with_gravity <- locations_sf %>%
  
  # compute index of the nearest gravity cell for each site
  mutate(nearest_idx = st_nearest_feature(., gravity_data)) %>%
  
  # bind the gravity columns by indexing the gravity table with nearest_idx
  {
    # grab grav columns from gravity table as a plain data.frame/tibble
    grav_df <- st_drop_geometry(gravity_data) %>% dplyr::select(contains("Grav"))
    # bind the matching rows (use drop = FALSE to preserve columns when single column)
    bind_cols(., grav_df[.$nearest_idx, , drop = FALSE])
  } %>%
  
  # compute distance once using the stored index
  mutate(distance_to_grid_km = as.numeric(
    st_distance(., gravity_data[nearest_idx, ], by_element = TRUE)
  ) / 1000) %>%
  
  # drop geometry and keep only the desired columns (nearest_idx will be removed)
  st_drop_geometry() %>%
  dplyr::select(
    date, station, site, country,
    latitude, longitude,
    contains("Grav"),
    distance_to_grid_km
  )

```

### Checking results

Convert the spatial object back to a regular dataframe and organize columns for easier interpretation.

```{r}
#| label: check-results
#| echo: false
#| warning: false


local({
  # local scope - prevents unnecessary objects that clutter the global environment
  missing_grav <- locations_with_gravity %>% filter(is.na(Grav_tot))

  if (nrow(missing_grav) > 0) {
    cat("⚠ Warning:", nrow(missing_grav), "sites missing gravity values\n")
    # drop geometry for nicer printing and only show the key columns
    print(missing_grav %>% st_drop_geometry() %>% select(station, site, location))
  } else {
    cat("All sites successfully matched with gravity values\n")
  }

  invisible(NULL)  # nothing returned to the parent environment
})

```

# Save data

...

```{r}

#| label: export-results
#| echo: false
#| warning: false

#load libraries
library(dplyr)
library(sf)

# Export results
write.csv(sites_with_gravity, 
          "sites_with_gravity_2018.csv", 
          row.names = FALSE)

# Export summaries
# write.csv(country_summary, 
#         "gravity_summary_by_country.csv", 
#          row.names = FALSE)

# write.csv(location_summary, 
#          "gravity_summary_by_location.csv", 
#          row.rows = FALSE)

# Save as spatial file if needed for GIS
# st_write(sites_with_gravity, 
#          "sites_with_gravity.shp", 
#         delete_dsn = TRUE)  # Overwrite if exists
```
