---
title: "adding_sst"
format: html
editor: visual
---

# Overview

This document processes NASA Ocean Color Level-3 binned (L3b) chlorophyll-a data.. and sea-surface temperature data (SST) for the Western Indian Ocean (WIO) region. The script:

-   Reads HDF5/NetCDF files containing binned satellite ocean color data
-   Converts NASA bin numbers to geographic coordinates (lat/lon)
-   Filters data to the WIO Area of Interest (37°E - 47°E, 15°S - 1°N)
-   Exports combined data to CSV and Parquet formats

# R Set-up

*Important note: Comprehensive instructions, including all prerequisites and detailed guidelines for running this analysis and relevant code, are available in the README file of my [GitHub repository](...) for this project. Please refer to the README to ensure proper setup and usage.*

## Load required libraries

```{r}
#| label: load-packages
#| echo: false
#| include: false

# Load required libraries
library(here)
library(ncdf4)
library(data.table)
library(arrow)
library(sf)
library(tidyverse) 
library(dplyr)      
library(readr)

source(here::here("functions", "extracting.R"))
```

```{r}

example_dhw <- extract_sat_data(filepath =here::here("raw_data", "dhw_raw_2009", "ct5km_dhw_v3.1_20080101.nc"), metric_name = "degree_heating_week", filter = TRUE)

example_dhw <- example_dhw %>% filter(!is.na(degree_heating_week))
```

```{r}
# Load location data and add columns to add relevant chlor-a data
locations_2009 <- read_rds(
  here::here("processed_data", "clean_location_2009.rds")) %>% 
  dplyr::select(date, station, site, country, latitude, longitude) %>% 
  dplyr::mutate(closest_cell_lat =0, closest_cell_lon=0, cell_distance_km = 0)

locations_with_dhw_sat_2009 <- assign_nearest_cell(example_dhw, locations_2009) %>% 
  dplyr::mutate(dhw_sum = 0)
```

```{r}
nc <- nc_open(here::here("raw_data", "dhw_raw_2009", "ct5km_dhw_v3.1_20080101.nc"))
```

```{r}
# Extract coordinates for the selected indices
lons <- ncvar_get(nc, "lon")
lats <- ncvar_get(nc, "lat")
length(lons)


i <- match(locations_with_dhw_sat_2009$closest_cell_lat, lats)
j <- match(locations_with_dhw_sat_2009$closest_cell_lon, lons)
keep_idxs <- cbind(col = j, row = i)
keep_idxs <- unique(keep_idxs)

lons[keep_idxs[1,1]] #  lon is index 2
lats[keep_idxs[1,2]] # lat is index 1


```

```{r}
#Testing keep_idxs is correct

test_lon_idx = which(lons < 43.3 & lons > 43.2)
test_lon_idx
test_lat_idx = which(lats <= -11.6 & lats >= -11.7)
test_lat_idx
lats

#43.275 #4466
#-11.6 #2034
#
m <- ncvar_get(nc, "degree_heating_week")
m[4466,2034]

lons[4466 ]
lats[2034]
```

```{r}
# Define path to all the raw chla files 
dhw_filenames_2009 <- list.files(
  here::here("raw_data", "dhw_raw_2009"),
  full.names = TRUE)

dhw_filenames_2009 <- dhw_filenames_2009

dhw_2009 <- 
lapply(dhw_filenames_2009, function(f) {
  extract_sat_data(filepath =  f, metric_name = "degree_heating_week", filter = TRUE, keep_idxs = keep_idxs)
})

# Combine into a single data table
raw_dhw_2009 <- data.table::rbindlist(dhw_2009, use.names = TRUE, fill = TRUE)

message(sprintf("Processing complete. Total rows extracted: %d", nrow(raw_dhw_2009)))


```

```{r}
readr::write_rds(raw_dhw_2009,here::here("raw_data","raw_dhw_2009.rds"))
```

```{r}
raw_dhw_2009 <- readr::read_rds(here::here("raw_data","raw_dhw_2009.rds"))

```

```{r}

for (i in 1:length(locations_with_dhw_sat_2009$site)){
  end_date <- locations_with_dhw_sat_2009$date[i]
  start_date <- end_date - days(364)
  
  closest_cell_lat = locations_with_dhw_sat_2009$closest_cell_lat[i]
  closest_cell_lon = locations_with_dhw_sat_2009$closest_cell_lon[i]
  
  
  relevant_cells <- raw_dhw_2009 %>%
      filter(
        lat == closest_cell_lat,
        lon == closest_cell_lon,
        date >= start_date,
        date <= end_date
      ) 
  
  sum <- sum(relevant_cells$degree_heating_week)
  
  locations_with_dhw_sat_2009$dhw_sum[i] = sum
}
```

```{r}
readr::write_rds(locations_with_dhw_sat_2009,here::here("processed_data","locations_with_dhw_2009"))

```

```{r}
library(ggplot2)
library(dplyr)
library(rnaturalearth)
library(rnaturalearthdata)

# World map
world <- ne_countries(scale = "medium", returnclass = "sf")

plot <- ggplot() +
  geom_sf(data = world, fill = "grey95", colour = "grey70") +
  geom_point(
    data = example_dhw,
    aes(x = lon, y = lat, colour = degree_heating_week),
    size = 1.5
  ) +
  scale_colour_gradient(
    low = "#56B1F7",
    high = "red3",
    na.value = "grey78",
    name = "Biomass"
  ) +
  coord_sf(
    xlim = c(37, 55),
    ylim = c(-19, 1),
    expand = FALSE
  ) +
  theme_classic() +   theme(aspect.ratio = 1)   #wider map

plot
```
