---
title: "basic_analyses"
format: html
editor: visual
---

# 

# Overview

This document explores the 2009 dataset. This data ...

# R Set-up

*Important note: Comprehensive instructions, including all prerequisites and detailed guidelines for running this analysis and relevant code, are available in the README file of my [GitHub repository](...) for this project. Please refer to the README ... to ensure proper setup and usage.*

Load the necessary packages (these can be found in ...) \## Packages

```{r}
#| label: r set-up
#| echo: false
#| include: false

#Load require libraries
library(tidyverse)
library(dplyr)
library(tidyr)

```

## Set working space

Use the renv::restore() to load up the necessary package versions along with the necessary dependencies.

```{r}
#| label: renv-restore
renv::restore()
```

## Load data

```{r}
#| label: load-data

fish_data_2009 <- readr::read_rds(here::here("processed_data", "clean_fish_2009.rds"))
location_2009 <- readr::read_rds(here::here("processed_data", "clean_location_2009.rds"))
sst_2009 <- read.csv(here::here("processed_data", "locations_with_sst_2009.csv"))
chla_2009 <- read.csv(here::here("processed_data", "locations_with_chla_2009.csv"))
h_gravity_2009 <- read.csv(here::here("processed_data", "locations_with_gravity.csv"))
```

## Functions

```{r}
descriptive_stats <- function(x) {
  x <- x[!is.na(x)]
  tibble(
    n        = length(x),
    mean     = mean(x),
    median   = median(x),
    sd       = sd(x),
    cv       = ifelse(mean(x) > 0, sd(x) / mean(x), NA_real_),
    min      = min(x),
    max      = max(x),
    prop0    = mean(x == 0)
  )
}

```

# Descriptive stats

# Biomass per site - assuming independent variables

```{r}
transect_biomass <- fish_data_2009 %>%
  group_by(location, site, ts_no) %>%
  summarise(
    tot_biomass = sum(tot_wt_g, na.rm = TRUE),
    .groups = "drop"
  )

(site_variation <- transect_biomass %>%
  group_by(site) %>%
  summarise(descriptive_stats(tot_biomass), .groups = "drop"))

(overall_site_summary <- site_variation %>%
  summarise(
    mean_n_transects= mean(n),
    mean_of_means   = mean(mean, na.rm = TRUE),
    median_of_means = median(mean, na.rm = TRUE),
    sd_of_means     = sd(mean, na.rm = TRUE),
    cv_of_means     = sd_of_means / mean_of_means,

    mean_cv         = mean(cv, na.rm = TRUE),
    median_cv       = median(cv, na.rm = TRUE),
    min_cv          = min(cv, na.rm = TRUE),
    max_cv          = max(cv, na.rm = TRUE)
  ))

```

# By location

```{r}
site_biomass <- fish_data_2009 %>%
  group_by(country, location, site) %>%
  summarise(
    tot_biomass = sum(tot_wt_g, na.rm = TRUE),
    .groups = "drop"
  )

(total_biomass_by_region <- site_biomass %>%
  group_by(location) %>%
  summarise(descriptive_stats(tot_biomass), .groups = "drop"))
```

# By reef geomorphology

```{r}
fish_2009 <- fish_data_2009 %>%
  left_join(
    location_2009 %>%
      dplyr::select(station, reef_geomorphology) %>%
      distinct(),
    by = "station"
  )

site_biomass_geo <- fish_2009 %>%
  group_by(reef_geomorphology, site) %>%
  summarise(
    tot_biomass = sum(tot_wt_g, na.rm = TRUE),
    .groups = "drop"
  )


(total_biomass_by_region <- site_biomass %>%
  group_by(reef_geomorphology) %>%
  summarise(descriptive_stats(tot_biomass), .groups = "drop"))
```

As show by the table, mean biomass differs somewhat among geomoprhologies but not greatly. i.e. geomorphology explains some but not the dominant variation.

In terms of the variance: Some geomorphologies (patch reef complexes) are internally very coherent Others (especially lagoon-exposed fringing reefs) are highly variable This suggests that geomorphology is not the driver of variation

If you compare with location, geomorphology explians far less variation - so not a niche enough group to include as a random effect.

# By country

```{r}
(total_biomass_by_country <- site_biomass %>%
  group_by(country) %>%
  summarise(descriptive_stats(tot_biomass), .groups = "drop"))
```

# Spatial auto-correlation checks

## stations

```{r}
fish_2009_2 <- fish_data_2009 %>%
  left_join(
    location_2009 %>%
      dplyr::select(station, latitude, longitude) %>%
      distinct(),
    by = "station"
  )

station_means <- fish_2009_2 %>%
  group_by(station, latitude, longitude) %>%
  summarise(
    biomass_station = mean(tot_wt_g, na.rm = TRUE),
    .groups = "drop"
  )
```

```{r}
# convert to s[atial object
library(sf)

station_sf <- st_as_sf(
  station_means,
  coords = c("longitude", "latitude"),
  crs = 4326
)

```

```{r}
# Define spatial neighbours

library(spdep)

coords <- st_coordinates(station_sf)

# define neighbours within 100 km
nb_station <- dnearneigh(
  coords,
  d1 = 0,
  d2 = 100000  # metres
)

lw_station <- nb2listw(nb_station, style = "W", zero.policy = TRUE)

```
