---
title: "prelim_analyses_temp_2009"
format: html
editor: visual
---

# Overview

This document explores the relationship between chlorophyll a and fish biomass.

# R Set-up

*Important note: Comprehensive instructions, including all prerequisites and detailed guidelines for running this analysis and relevant code, are available in the README file of my [GitHub repository](...) for this project. Please refer to the README ... to ensure proper setup and usage.*

Load the necessary packages (these can be found in ...)

Use the renv::restore() to load up the necessary package versions along with the necessary dependencies.

```{r}
#| label: r set-up
#| echo: false
#| include: false

#Load require libraries
library(tidyverse)
library(sf)
library(arm)
library(here)
library(dplyr)
library(tidyverse)
library(ggplot2)
library(ggfortify) # extends ggplot2 for some more ease with stats
library(MASS) # fits regression models etc.
library(MuMIn) # multi-model inference (AIC etc.)
library(AICcmodavg)
library(DHARMa)
library(mgcv)
library(tweedie)
library(glmmTMB)
```

```{r}
#| label: renv-restore
# renv::restore()
```

## Load data

```{r}
#| label: load-data

fish_2009 <- readr::read_rds(here::here("processed_data", "clean_fish_2009.rds"))
location_2009 <- readr::read_rds(here::here("processed_data", "clean_location_2009.rds"))
dhw_2009 <- readr::read_rds(here::here("processed_data", "locations_with_dhw_2009.rds"))

```

## Total Biomass

This refers to an aggregate of all species.

### Calculate total biomass per station, site, and country

Biomass was quantified at the transect level, with each transect representing a replicate sample within a site. Because transects within a site are spatially proximate and not statistically independent, sites were treated as the independent units in all analyses. Need to explain the stations as well as many metrics are taken using such coordinates!!

Environmental predictors were extracted at station coordinates and then averaged to the site level; stations were grouped into sites if they fell within 2 km of each other. Because environmental rasters are at \~4 km resolution, centroid-based extraction produced effectively identical values in our data (see diagnostic), so either approach is acceptable — we report station-mean values to preserve the sampled footprint.

Total biomass was first calculated for each transect by summing biomass across species, and site-level biomass was then estimated as the mean of its 3–5 constituent transects. This approach ensures equal weighting of transects within sites and avoids pseudo-replication in comparisons among sites.

```{r}
#| label: calculate-total-data

# Aggregate biomass at transect level
transect_biomass <- fish_2009 %>%
  group_by(site, station, ts_no) %>%
  summarise(
    transect_biomass = sum(tot_wt_g, na.rm = TRUE),
    country = first(country),
    .groups = "drop")

# Mean biomass per site
biomass_per_site <- transect_biomass %>%
  group_by(site) %>%
  summarise(
    mean_biomass = mean(transect_biomass),
    country = first(country),
    .groups = "drop")

# Mean biomass per country
biomass_per_country <- biomass_per_site %>%
  group_by(country) %>%
  summarise(
    mean_biomass = mean(mean_biomass, na.rm = TRUE))

```

### check number of transects per site

There are some that have fewer than 3 - either group or remove. There are also some with more than 5 - check these (maybe split if needs be)

```{r}
# Number of transects per site
transect_biomass %>%
  group_by(site) %>%
  summarise(ts_no = n())

# Number of transects per station
transect_biomass %>% 
    group_by(station) %>%
  summarise(ts_no = n())
```

### Biomass

Check the structure of biomass data

```{r}
# Overview of the data
head(biomass_per_site,3) # see first three rows
str(biomass_per_site) #structure of data
summary(biomass_per_site) #summary of data

# Histogram 
( biomass_hist <- ggplot(biomass_per_site, aes(x = mean_biomass)) +
  geom_histogram(bins = 30) +
  labs(
    x = expression("Mean site biomass "),
    y = "Frequency"
  ) +
  theme_bw() )

# Boxcox 
boxcox(lm(mean_biomass ~ 1, data = biomass_per_site),
       lambda = seq(-2, 2, 0.1))
```

### Calculate env. variable means

```{r}
#| label: calculate-env-var-means

# Calculate mean dhw per site
dhw_2009_sites <- dhw_2009 %>% 
  group_by(site) %>% 
  summarise(mean_max_dhw = mean(max_dhw))

# alterantive:
# dhw_2009_sites2 <- dhw_2009 %>% 
#   group_by(site) %>% 
#   summarise(
#     mean_dhw_all = mean(dhw_sum, na.rm = TRUE),
#     mean_dhw_when_present = mean(dhw_sum[dhw_sum > 0], na.rm = TRUE),
#     proportion_zero = mean(dhw_sum == 0, na.rm = TRUE)
#   )

```

# Biomass vs DHW

## DHW

```{r}

# Overview of the data
head(dhw_2009_sites,3) # see first three rows
str(dhw_2009_sites) #structure of data
summary(dhw_2009_sites) #summary of data

# Histogram 
( dhw_hist <- ggplot(dhw_2009_sites, aes(x = mean_max_dhw_sum)) +
  geom_histogram(bins = 20) +
  labs(
    x = expression("Mean dhw sum"),
    y = "Frequency"
  ) +
  theme_bw() )

```

Keep zeros... maybe don't transform jsut yet...

## Exploratory Plot

```{r}
#| label: biomass-vs-dhw-plot

# Graph the data - X-Y scatter plot
biomass_dhw_fig1 <- merge(x = biomass_per_site, y = dhw_2009_sites, by = "site", all = TRUE) %>% 
  ggplot(aes(x=mean_dhw_sum, y=mean_biomass)) + 
  geom_point() +
  geom_smooth(method=lm)

biomass_dhw_fig1

# Correlation
with(merge(x = biomass_per_site, y = dhw_2009_sites, by = "site", all = TRUE),
    cor(mean_dhw_sum, mean_biomass))

```

Chlorophyll a and biomass seem to be weakly positively related, with a correlation of 0.16, but the are large confidence intervals, and a lot of uncertainty in the data.

## Simple ls model

Note that chlorophyll-a, as a continuous predictors, was mean-centred prior to analysis to improve interpretability of model parameters. Centering ensures that the intercept represents the expected response at the average value of the predictor, rather than at zero, which would be biologically unrealistic.

```{r}
#| label: biomass-vs-dhw-lm

# Merge needed data
dhw_model_data <- merge(x = biomass_per_site, y = dhw_2009_sites, by = "site", all = TRUE)

# Centre dhw
dhw_model_data$mean_dhw_centred <- as.numeric(scale(
  dhw_model_data$mean_dhw_sum,
  center = TRUE,
  scale = FALSE
))

# Build lm 
lm_biomass_dhw <- lm(mean_biomass~mean_dhw_centred,data = dhw_model_data) 
```

Analyse model:

```{r}
# Analyse the model 
coef(lm_biomass_dhw)
#residuals(lm_biomass_dhw)
summary(lm_biomass_dhw) 
# or arm::display(lm_biomass_dhw)

confint(lm_biomass_dhw) 

```

Outputs:

1.  Intercept: Because chlorophyll was centred, the intercept (21,397) represents the expected biomass at average chlorophyll concentration. The 95% CI (16,982 to 25,811) gives the plausible range for that mean biomass level.

2.  Slopes The slope estimate (3,934) represents the expected increase in biomass for each one-unit increase in chlorophyll above the mean. However, this effect is not statistically significant (p = 0.206), and the confidence interval includes zero (−2,218 to 10,085). This means the data do not provide evidence of a linear relationship between chlorophyll and biomass on the raw scale.

3.  Correlation: 0.162924 (weakly positive)

4.  Effect size The model explains very little variation (R² ≈ 0.027), indicating that chlorophyll alone accounts for only about 3% of the variability in biomass.

Overall, there is no strong support for a simple linear biomass–chlorophyll relationship in this specification.

### Residual checking of simple ls model

Checking model assumptions and violations:

```{r}
par(mfrow= c(2,2)) # divide the page into 2 x 2 frame
plot( lm_biomass_dhw, which= c(1:4))

```

Diagnostic plots: - Slight downward trend in residual smooth line - Increasing spread at higher fitted values - Extreme positive outliers (obs 16, 21, 26)

This suggests possible non-linearity (the curve is not flat) and heteroscedasticity (variance increases as fitted values increase) - i.e. residuals are not normally distributed

## Tranformations

To improve model fit, we can transform the raw data to better conform to the regression assumptions, thereby amplifying predictive power and increasing the overall quality of the model.

We also check how the distribution of the data changes after each transformation to ensure that everything is working as it should.

```{r}

# Transforming variables
dhw_model_data$log_biomass <- log10(dhw_model_data$mean_biomass)
dhw_model_data$log_dhw <- log10(dhw_model_data$mean_dhw_sum)

# Check transformations 
( log_biomass_hist <- ggplot(dhw_model_data, aes(x = log_biomass)) +
  geom_histogram(bins = 20) +
  labs(
    x = expression("Log-Transformed Biomass"),
    y = "Frequency"
  ) +
  theme_bw() )

# ( log_dhw_hist <- ggplot(dhw_model_data, aes(x = log_dhw)) +
#   geom_histogram(bins = 20) +
#   labs(
#     x = expression("Log-Transformed Chlorophyll-a"),
#     y = "Frequency"
#   ) +
#   theme_bw() )
```

The data is now looking much more normal!! Great :)

We can now fit models with the transformed data. Ensure you are also centering the log_dhw!

```{r}

# # Centre logged dhw
# dhw_model_data$log_dhw_centred <- as.numeric(
#   scale(dhw_model_data$log_dhw, center = TRUE, scale = FALSE)
# )


# Transformed models 
log_biomass_dhw <- lm(log_biomass~mean_dhw_centred,data = dhw_model_data) 
# biomass_log_dhw <- lm(mean_biomass~log_dhw_centred,data = dhw_model_data) 
# log_biomass_log_dhw <- lm(log_biomass~log_dhw_centred,data = dhw_model_data) 

# Summaries of transformed models
summary(log_biomass_dhw) 
# summary(biomass_log_dhw) 
# summary(log_biomass_log_dhw) 

# or arm::display(lm_biomass_dhw)
#
```

The pattern is very consistent: none of the simple bivariate models show evidence of a relationship between chlorophyll and biomass.

First model: log_biomass \~ mean_annual_dhw

The slope is −0.016 (p = 0.79), with R² ≈ 0.001. This indicates essentially no relationship between raw chlorophyll and log-transformed biomass. Chlorophyll explains virtually none of the variation in log biomass.

Second model: mean_biomass \~ log_dhw_centred

The slope is 6,048 (p = 0.34), R² ≈ 0.015. This suggests that a one-unit increase in log10 chlorophyll (i.e., a tenfold increase in chlorophyll) would be associated with an increase of about 6,000 biomass units, but this effect is not statistically supported (the confidence interval spans zero).

Because log_dhw is centred, the intercept (21,397) represents mean biomass at average log chlorophyll. That makes the intercept interpretable, but it does not change inference.

Third model: log_biomass \~ log_dhw_centred (log–log model)

The slope is −0.007 (p = 0.95), R² ≈ 0.00006. This is effectively zero. In scaling terms, the estimated exponent is essentially zero, meaning biomass does not scale with chlorophyll in a simple global power-law relationship.

### Residual checking of transformations

```{r}

# biomass_log_dhw
par(mfrow= c(2,2)) # divide the page into 2 x 2 frame
plot(biomass_log_dhw, which= c(1:4))

# log_biomass_dhw
par(mfrow= c(2,2)) # divide the page into 2 x 2 frame
plot(log_biomass_dhw, which= c(1:4))

# log_biomass_log_dhw
par(mfrow= c(2,2)) # divide the page into 2 x 2 frame
plot(log_biomass_log_dhw, which= c(1:4))
```

Even though neither model shows a significant slope, the log–log form better satisfies linear model assumptions slightly, as it reduces The residuals look the best for the log_log model, with the assumptions now satisfied -\> The log-transformed predictor slightly improves linearity.

### Check AICs

Note that you cannot compare AIC across different response scales, so, given that log-biomass better fits the residuals, we are interested to know which model - log_biomass_dhw or log-log - provides the best fit

```{r}
models <- list(log_biomass_dhw, log_biomass_log_dhw)
aictab(models)
```

When we compare these models we see that they are essentially identical:

ΔAICc = 0.00/0.07 → essentially identical. AIC weights ≈ 0.5 / 0.5 → no meaningful support difference. Log-likelihoods almost identical.

Statistically: The data cannot distinguish between these two models. R²: 0.0012 vs 0.00006s Both slopes non-significant (≈ 0)

So, model comparison based on AICc indicated no meaningful difference between the log–linear and log–log formulations (ΔAICc = 0.07), and both models explained negligible variation in biomass (R² ≈ 0).

However, diagnostic plots suggested slightly improved homoscedasticity and linearity under the log–log specification. Given the equivalent statistical support and the expectation of power-law scaling in ecological productivity relationships, the log–log model was retained for subsequent analyses.

\< Competing models were indistinguishable based on AICc (ΔAICc \< 0.1); however, the log–log specification exhibited improved residual structure and provides a biologically interpretable scaling relationship. We therefore retained the log-transformed predictor. \>

## Taking into account eco-region (country)

### Fixed-effects model (separate)

We will first separate this into each different eco-region (country for now)

```{r}
# plot of biomass vs chlorophyll for a 'no pooling analysis'
( separate_lm_plot <- ggplot(dhw_model_data, aes(x = log_dhw_centred,
           y = log_biomass)) + geom_point() + geom_smooth(method = "lm", formula = y ~ x) +
  facet_grid(. ~ country) + theme_bw() + labs( x = "Chlorophyll a",
    y = "Log-Biomass", title = "No pooling analysis") )

```

Great differences across countries!!

This analysis treats the relationship at each site as unique and independent of the others.

### Combined fixed-effects model

At the other extreme, we could simply pool all the data together in a single simple regression analysis: (same as above).

```{r}
# Graph the data - X-Y scatter plot
combined_lm_plot <- dhw_model_data %>% 
  ggplot(aes(x=log_dhw_centred, y=log_biomass)) + 
  geom_point() +
  geom_smooth(method=lm)

combined_lm_plot
```

This model helps us to understand the analysis we want to build from these initial simple extremes.

### Country as a fixed effect

Note: - from above, we will carry on with the log_log model - we centre dhw. In models including interactions, centering also makes main effects interpretable as effects evaluated at the mean of the interacting variable and can improve numerical stability in mixed-effects models. - We used sum-to-zero contrasts (effect coding) for country so that parameter estimates represent deviations from the overall mean rather than differences relative to an arbitrarily chosen reference country. Under effect coding, the intercept corresponds to the grand mean of the response (averaged across all observations), and each country coefficient represents that country’s deviation from this grand mean. In other words, coefficients quantify how much higher or lower a given country’s mean is compared to the overall mean.

#### Additive model

First we add it in an additive simple least squares regression. This additive model assumes that the biomass–chlorophyll slope is the same across all countries, but allows countries to differ in their mean biomass levels:

```{r}
# Set sum-to-zero contrasts for country (the grand mean is the baseline - not an arbitrary country level)
contrasts(dhw_model_data$country) <- contr.sum(length(levels(dhw_model_data$country)))


dhw_country_additive <- lm(log_biomass ~ log_dhw_centred + country, data = dhw_model_data)
summary(dhw_country_additive) 
```

Output interpretation:

1.  Intercept Because log chlorophyll is centred, the intercept (4.208) represents the grand mean log biomass at average chlorophyll concentration. Under effect coding, each country coefficient represents that country’s deviation from this overall mean (evaluated at mean chlorophyll), not from a specific reference country.

2.  Slope The slope for log chlorophyll is estiamted to be −0.364 (p = 0.019). This indicates that, when a common slope is imposed across countries, biomass declines with increasing chlorophyll on the log–log scale. In scaling terms, this corresponds to a negative exponent: at higher chlorophyll, biomass is lower, on average. However, this interpretation must be treated cautiously because earlier analyses showed evidence of slope heterogeneity across countries. If slopes differ, this “average” negative slope may reflect opposing patterns cancelling out. So beware with this interpretation!

Country1 has significantly lower biomass than the grand mean (−0.418), while country2 has significantly higher biomass (+0.323), both evaluated at mean chlorophyll. Countries 3 and 4 do not differ significantly from the overall mean.

The model explains about 19% of the variance (R² = 0.19), and the overall model is statistically significant (p = 0.033), indicating that adding country improves fit relative to a simple log–log model without country.

In summary, this additive model suggests that biomass differs among countries and shows a negative average biomass–chlorophyll relationship.

However, given your earlier interaction results, the assumption of a common slope across countries is likely too restrictive, and the interaction model is probably the more appropriate representation of the data.

#### Interaction model

This interaction model allows the biomass–chlorophyll slope to differ among countries rather than forcing a common slope.

```{r}

# Log-log interaction
dhw_country_interaction <- lm(log_biomass ~ log_dhw_centred * country,
  data = dhw_model_data)
summary(dhw_country_interaction)

anova(dhw_country_interaction)

```

Interpration of outputs:

1.  Intercept Because log chlorophyll is centred and effect coding is used, the intercept (4.276) represents the grand mean log biomass at average chlorophyll across all countries.

2.  Slopes The main effect of log chlorophyll (0.058, p = 0.81) represents the average slope across countries, but in the presence of an interaction this average is not biologically meaningful on its own. The country main effects now represent deviations from the grand mean biomass at mean chlorophyll, conditional on the interaction being included.

3.  Interaction coefficients These are the key terms of the model. These quantify how each country’s slope differs from the overall (grand) slope. Most interaction terms are not statistically different from zero, but the interaction for country4 is strongly negative (−0.899, p = 0.004), indicating that the biomass–chlorophyll scaling relationship in that country differs substantially from the overall trend. Country3 also shows a positive deviation from the mean slope (0.757), with marginal evidence (p = 0.078).

Importantly, the overall model fit improves considerably relative to the additive model. The R² increases from 0.19 to 0.36, and the overall F-test is highly significant (p = 0.003). This indicates that allowing slopes to vary by country explains substantially more variation in biomass.

In substantive terms, there is no single universal biomass–chlorophyll scaling relationship across the region. Instead, the strength and direction of the relationship depend on country, with at least one country exhibiting a markedly different (negative) scaling exponent. This supports the conclusion that the biomass–productivity relationship is spatially heterogeneous rather than globally consistent.

ANOVa - ancova...

#### Test global, additive and interaction models

```{r}
#AIC
interaction_test_models <- list(log_biomass_log_dhw, dhw_country_additive, dhw_country_interaction)
aictab(interaction_test_models) 

# Classical F-test
anova(log_biomass_log_dhw, dhw_country_additive, dhw_country_interaction)
```

The three models represent increasing structural complexity: a global log–log relationship only, an additive model including country-level differences in mean biomass, and a full interaction model allowing country-specific slopes.

The AICc comparison clearly favours the interaction model. It has the lowest AICc (38.91) and carries 83% of the model weight. The additive model is moderately supported (ΔAICc = 3.45, weight = 0.15), while the simple global log–log model has very little support (ΔAICc = 6.88, weight = 0.03). A ΔAICc greater than 6 indicates substantially less support, so the global model is effectively ruled out.

The nested F-tests lead to the same conclusion. Adding country to the global log–log model significantly improves fit (p = 0.008), indicating that mean biomass differs among countries. Adding the interaction further improves fit (p = 0.014), demonstrating that slopes differ among countries. Thus, both country-level differences in mean biomass and heterogeneity in scaling exponents are statistically supported.

Taken together, both information-theoretic (AICc) and classical hypothesis testing approaches converge on the same conclusion: the interaction model provides the best representation of the data. There is no single biomass–chlorophyll relationship across the region; instead, both intercepts and slopes vary among countries.

#### Try country as a random effect

Because sites are nested within countries and sample sizes are uneven, we additionally explored a mixed-effects formulation in which country-specific intercepts and slopes were treated as random deviations rather than fixed parameters.

This specification allows each country to have its own biomass–chlorophyll slope, while assuming that these slopes arise from a common underlying distribution. Conceptually, this induces partial pooling: countries with fewer observations are shrunk toward the overall mean slope, reducing the influence of small sample sizes:

Randome slopes only: This removes the intercept variance parameter, avoiding the ... boundary issue. It directly models heterogeneity in slopes, which appears to be the dominant signal in your data.

```{r}

# Random slope only model
lmer(log_biomass ~ log_dhw_centred + 
       (0 + log_dhw_centred | country),
     data = dhw_model_data)

```

Output interpretation: The fixed intercept (4.23) represents the mean log biomass at average chlorophyll across the region. The fixed slope (0.043) represents the average scaling exponent across countries. That average effect is small and close to zero.

The key result is the random slope standard deviation of 0.598. This is large relative to the fixed slope. It implies substantial heterogeneity in biomass–chlorophyll scaling among countries. Roughly speaking, country-specific slopes are distributed around 0.043 with a standard deviation of about 0.60. That means some countries likely have strongly positive exponents while others have negative ones.

This aligns closely with your fixed interaction model, which also showed slope heterogeneity and a particularly strong deviation for one country. The mixed model expresses the same signal in variance terms rather than individual coefficients.

However, the interpretational limitation remains: the slope variance is estimated from only five countries. So while the magnitude suggests real heterogeneity, its precision is limited. The model is statistically coherent, but inference about the variance component should be cautious.

Random intercept only model: This tests whether there is overall between-country variation in mean biomass without attempting to estimate slope variance. With five groups, this model is much more stable.

```{r}

# Random intercept only model
lmer(log_biomass ~ log_dhw_centred + (1 | country),
     data = dhw_model_data)
```

this random-intercept-only model is statistically stable and easier to interpret than the random-slope formulation.

Here, the fixed intercept (4.21) represents the mean log biomass at average chlorophyll across the WIO. The fixed slope (−0.224) represents the average biomass–chlorophyll scaling exponent across countries, assuming that slope is common to all countries.

The random intercept standard deviation (0.188) indicates that countries differ in their baseline biomass levels (at mean chlorophyll). This confirms between-country variation in mean biomass. Because you are not estimating slope variance, the model avoids the instability seen in the random-slope model.

However, this model assumes that the biomass–chlorophyll relationship is identical across countries. That assumption conflicts with your earlier fixed-interaction and random-slope results, which showed substantial slope heterogeneity. In other words, this model is statistically stable but structurally restrictive.

Substantively, this model says: countries differ in mean biomass, but they share a common biomass–chlorophyll slope. Your earlier analyses suggest that this is unlikely to be true.

Note that random-effects models rely on sufficient numbers of higher-level units to reliably estimate variance components. With so few groups, the estimation of slope and intercept variances becomes unstable and may lead to boundary estimates or singular fits. In practical terms, this means that the random-effects parameters are weakly identified and sensitive to small changes in the data.

### Try a GLM

Given that biomass is positive and right-skewed, we additionally fitted generalized linear models using a Gamma error distribution with a log link, which models multiplicative effects directly on the original biomass scale and accommodates mean–variance relationships typical of ecological biomass data.

First we begin with an additive model:

```{r}
# Gaussian
glm_biomass_dhw_gaussian <- glm(
  mean_biomass ~ mean_dhw_centred,
  family = gaussian(link = "identity"),
  data = dhw_model_data
)

summary(glm_biomass_dhw_gaussian)

# GLM
glm_biomass_dhw_gamma <- glm(
  mean_biomass ~ mean_dhw_centred,
  family = Gamma(link = "log"),
  data = dhw_model_data
)

summary(glm_biomass_dhw_gamma)


```

```{r}
models <- list(glm_biomass_dhw_gaussian, glm_biomass_dhw_gamma)
aictab(models)
```

```{r}

# Additive
glm_gamma_additive <- glm(
  mean_biomass ~ mean_dhw_centred + country,
  family = Gamma(link = "log"),
  data = dhw_model_data
)

summary(glm_gamma_additive)

# Interaction
glm_gamma_interaction <- glm(
  mean_biomass ~ mean_dhw_centred*country,
  family = Gamma(link = "log"),
  data = dhw_model_data
)

summary(glm_gamma_interaction)
```

Note that the Gamma GLM results closely mirror your Gaussian log–log results in terms of structure.

In this additive Gamma model, the chlorophyll effect is negative but not statistically significant (p = 0.14), and country1 and country2 differ significantly from the regional mean. That is broadly consistent with your additive log–log model.

Now let's check residuals:

```{r}
plot(glm_gamma_additive)
```

...

Then we try the interaction, multiplicative model:

```{r}
glm_gamma_interaction <- glm(
  mean_biomass ~ log_dhw_centred * country,
  family = Gamma(link = "log"),
  data = dhw_model_data
)

summary(glm_gamma_interaction)
```

In this interaction Gamma model, as with the lm model, the interaction term for country4 is strongly negative and significant (p = 0.004), and country3 shows marginal evidence of slope deviation. This is almost identical to what you saw in the Gaussian log–log interaction model. So the mean-structure conclusions are robust to distributional assumptions.

```{r}
plot(glm_gamma_interaction)
```

```{r}
glm_test_models <- list(glm_biomass_dhw_gaussian, glm_biomass_dhw_gamma, glm_gamma_additive, glm_gamma_interaction)
aictab(glm_test_models)
```

The interaction Gamma model has substantially lower AICc (ΔAICc = 5.84) and carries about 95% of the model weight. That is strong support for slope heterogeneity under a Gamma error structure as well.

Given that the linear model is simpler I will keep this :)

### Other groupings! ...

```{r}
dhw_gamma2 <- glm(mean_biomass ~ mean_annual_dhw + location, data = dhw_model_data,
                  family = Gamma(link = "log"))

summary(dhw_gamma2)
(AIC_location <- AIC(dhw_gamma2))
(AICc_location <- AICc(dhw_gamma2))
```

```{r}
AIC(dhw_gamma, dhw_gamma2)
AICc(dhw_gamma, dhw_gamma2)

```

##### Plot

```{r}
#| label: biomass-vs-sst-plot

# Graph the data - X-Y scatterplot
biomass_sst_fig1 <- merge(x = biomass_per_site, y = sst_2009_sites, by = "site", all = TRUE) %>% 
  ggplot(aes(x=mean_annual_sst, y=mean_biomass)) + 
  geom_point() +
  geom_smooth(method=lm) 

biomass_sst_fig1


biomass_sst_fig2 <- merge(x = biomass_per_site, y = sst_2009_sites, by = "site", all = TRUE) %>% 
ggplot(aes(mean_annual_sst, log(mean_biomass), colour = country)) +
  geom_point(alpha = 0.6) +
  geom_smooth(method = lm, se = FALSE)
biomass_sst_fig2
```

##log

```{r}
#| label: biomass-vs-sst-plot

# Graph the data - X-Y scatterplot
biomass_sst_fig2 <- merge(x = biomass_per_site, y = sst_2009_sites, by = "site", all = TRUE) %>% 
  ggplot(aes(x=mean_annual_sst, y=log(mean_biomass))) + 
  geom_point() +
  geom_smooth(method=lm) 

biomass_sst_fig2
```

```{r}
par(mfrow= c(1,2)) # divide the page into 1 x 2 frame
plot( lm_biomass_dhw, which= c(1:2))
```

##### Model

```{r}
#| label: biomass-vs-sst-lm

# Merge needed data
sst_x_and_y_site <- merge(x = biomass_per_site, y = sst_2009_sites, by = "site", all = TRUE)

# Build lm 
lm_biomass_sst <- lm(mean_biomass~mean_annual_sst,data = sst_x_and_y_site) 

# Analyse the model 
summary(lm_biomass_sst) 
```

## Log

```{r}

# Merge needed data
sst_x_and_y_site <- merge(x = biomass_per_site, y = sst_2009_sites, by = "site", all = TRUE)

# Build lm 
lm_biomass_sst <- lm(log(mean_biomass)~mean_annual_sst,data = sst_x_and_y_site) 

# Analyse the model 
summary(lm_biomass_sst)


library(lme4)

model_sst <- lmer(
  log(mean_biomass) ~ mean_annual_sst + (1 | country),
  data = merge(biomass_per_site, sst_2009_sites, by = "site")
)

summary(model_sst)
```

## Comparing differnet SST metrics

# Biomass vs Connectivity

## Plot

```{r}
#| label: biomass-vs-gravity-plot

# Graph the data - X-Y scatterplot
biomass_gravity_fig1 <- merge(x = biomass_per_site, y = h_gravity_2009_sites, by = "site", all = TRUE) %>% 
  ggplot(aes(x=mean_gravity, y=mean_biomass)) + 
  geom_point() +
  geom_smooth(method=lm) 

biomass_gravity_fig1

biomass_gravity_fig2 <- merge(x = biomass_per_site, y = h_gravity_2009_sites, by = "site", all = TRUE) %>% 
ggplot(aes(mean_gravity, log(mean_biomass), colour = country)) +
  geom_point(alpha = 0.6) +
  geom_smooth(method = lm, se = FALSE)
biomass_gravity_fig2
```

## log

```{r}
#| label: biomass-vs-gravity-plot

# Graph the data - X-Y scatterplot
biomass_gravity_fig2 <- merge(x = biomass_per_site, y = h_gravity_2009_sites, by = "site", all = TRUE) %>% 
  ggplot(aes(x=mean_gravity, y=log(mean_biomass))) + 
  geom_point() +
  geom_smooth(method=lm) 

biomass_gravity_fig2
```

##### Model

```{r}
#| label: biomass-vs-gravity-lm

# Merge needed data
gravity_x_and_y_site <- merge(x = biomass_per_site, y = h_gravity_2009_sites, by = "site", all = TRUE)

# Build lm 
lm_biomass_gravity <- lm(mean_biomass~mean_gravity,data = gravity_x_and_y_site) 
# Analyse the model 
summary(lm_biomass_gravity) 

```

### Log

```{r}
#| label: biomass-vs-gravity-lm

# Merge needed data
gravity_x_and_y_site <- merge(x = biomass_per_site, y = h_gravity_2009_sites, by = "site", all = TRUE)

# Build lm 
lm_biomass_gravity <- lm(log(mean_biomass)~mean_gravity,data = gravity_x_and_y_site) 
# Analyse the model 
summary(lm_biomass_gravity) 

```
