---
title: "perliminary_analyses"
format: html
editor: visual
---

# Overview

This document explores the 2009 dataset. This data ...

# R Set-up

*Important note: Comprehensive instructions, including all prerequisites and detailed guidelines for running this analysis and relevant code, are available in the README file of my [GitHub repository](...) for this project. Please refer to the README ... to ensure proper setup and usage.*

Load the necessary packages (these can be found in ...)

Use the renv::restore() to load up the necessary package versions along with the necessary dependencies.

```{r}
#| label: r set-up
#| echo: false
#| include: false

#...
library(tidyverse)
library(here)
library(dplyr)
library(tidyverse)
library(ggplot2)
library(ggfortify) # extends ggplot2 for some more ease with stats
library(MASS) # fits regression models etc.

```

```{r}
#| label: renv-restore
renv::restore()
```



#...

## Load data 2009

```{r}
#| label: load data
#| echo: false
#| warning: false
#| message: false

fish_data_2009 <- readr::read_rds(here::here("processed_data", "clean_fish_2009.rds"))
location_2009 <- readr::read_rds(here::here("processed_data", "clean_location_2009.rds"))
sst_2009 <- read.csv(here::here("processed_data", "locations_with_sst_2009.csv"))
chla_2009 <- read.csv(here::here("processed_data", "locations_with_chla_2009.csv"))
h_gravity_2009 <- read.csv(here::here("processed_data", "locations_with_gravity.csv"))
```


### Calculate total biomass per station

```{r}

biomass_per_station <- fish_data_2009 %>%
  group_by(station) %>%
  summarise(
    tot_biomass = sum(tot_wt_g, na.rm = TRUE),
    site = site,
    country = country) #%>% 
#  summarise(
#    mean_biomass = sum(tot_wt_g, na.rm = TRUE),
#    site = site)

biomass_per_site <- biomass_per_station %>%
  group_by(site) %>%
  summarise(
    av_biomass = mean(tot_biomass, na.rm = TRUE),
    country = country)

```

```{r}

chla_2009_sites <- chla_2009 %>% 
  group_by(site) %>% 
  summarise(
    av_annual_chla = mean(chla_annual_mean)
  )

sst_2009_sites <- sst_2009 %>% 
  group_by(site) %>% 
  summarise(
    av_annual_sst = mean(sst_annual_mean)
  )

h_gravity_2009_sites <- h_gravity_2009 %>% 
  group_by(site) %>% 
  summarise(
    av_annual_gravity = mean(Grav_tot)
  )



```



### Linear regression


```{r}
#Graph the data - X-Y Scatterplot
library(ggplot2)
fig1 <- merge(x = biomass_per_site, y = chla_2009_sites, by = "site", all = TRUE) %>% 
  ggplot(aes(x=av_annual_chla, y=av_biomass)) + geom_point()
fig1
```

```{r}
#Graph the data - X-Y Scatterplot
library(ggplot2)
fig1 <- merge(x = biomass_per_site, y = sst_2009_sites, by = "site", all = TRUE) %>% 
  ggplot(aes(x=av_annual_sst, y=av_biomass)) + geom_point()
fig1
```





```{r}
#Graph the data - X-Y Scatterplot
library(ggplot2)
fig1 <- merge(x = biomass_per_station, y = chla_2009, by = "station", all = TRUE) %>% 
  ggplot(aes(x=chla_annual_mean, y=tot_biomass)) + geom_point()
fig1
```
```{r}
#Graph the data - X-Y Scatterplot
library(ggplot2)
fig1 <- merge(x = biomass_per_site, y = h_gravity_2009_sites, by = "site", all = TRUE) %>% 
  ggplot(aes(x=av_annual_gravity, y=av_biomass)) + geom_point()
fig1
```



```{r}
library(ggplot2)
fig2 <- merge(x = tot_biomass_per_station, y = sst_2009, by = "station", all = TRUE) %>% 
  ggplot(aes(x=sst_annual_mean, y=tot_biomass)) + geom_point()
fig2
```

```{r}
library(ggplot2)
fig3 <- merge(x = tot_biomass_per_station, y = h_gravity_2009, by = "station", all = TRUE) %>% 
  ggplot(aes(x=Grav_tot, y=tot_biomass)) + geom_point()
fig3
```

# Linear Model ----
we want to fit a linear model (.;. lm() fucntion), with hardness as the response and density as the explanatory, using the janka data set

```{r}
chla_x_and_y_site <- merge(x = biomass_per_site, y = chla_2009_sites, by = "site", all = TRUE)

model1 <- lm(av_biomass~av_annual_chla,data = chla_x_and_y_site) 
# Analyse the model using extractor functions ----
coef(model1) # extract coefficients
?coef() # to get details about the function

confint(model1) # extract confidence intervals
?confint() # to get details about the function

summary(model1) # summary of model, including significance values

library(arm) #load package
display(model1) #load function

fig1 <- chla_x_and_y_site %>% 
  ggplot(aes(x=av_annual_chla, y=av_biomass)) + geom_point() + geom_smooth(method=lm) #this ensures that the line is straight (as per linear regression)
fig1
?geom_smooth
```
```{r}
sst_x_and_y_site <- merge(x = biomass_per_site, y = sst_2009_sites, by = "site", all = TRUE)

model1 <- lm(av_biomass~av_annual_sst,data = sst_x_and_y_site) 
# Analyse the model using extractor functions ----
coef(model1) # extract coefficients
?coef() # to get details about the function

confint(model1) # extract confidence intervals
?confint() # to get details about the function

summary(model1) # summary of model, including significance values

library(arm) #load package
display(model1) #load function

fig3 <- sst_x_and_y_site %>% 
  ggplot(aes(x=av_annual_sst, y=av_biomass)) + geom_point() + geom_smooth(method=lm) #this ensures that the line is straight (as per linear regression)
fig3
?geom_smooth
```


```{r}
gravity_x_and_y_site <- merge(x = biomass_per_site, y = h_gravity_2009_sites, by = "site", all = TRUE)

model1 <- lm(av_biomass~av_annual_gravity,data = gravity_x_and_y_site) 

# Analyse the model using extractor functions ----
#coef(model1) # extract coefficients
#?coef() # to get details about the function

#confint(model1) # extract confidence intervals
#?confint() # to get details about the function

#summary(model1) # summary of model, including significance values

library(arm) #load package
display(model1) #load function

fig2 <- gravity_x_and_y_site %>% 
  ggplot(aes(x=av_annual_gravity, y=av_biomass)) + geom_point() + geom_smooth(method=lm) #this ensures that the line is straight (as per linear regression)
fig2
?geom_smooth
```
##By country
```{r}
biomass_per_country <- biomass_per_site %>%
  group_by(country) %>%
  summarise(
    av_biomass = mean(av_biomass, na.rm = TRUE),)
```

```{r}
fig1 <- biomass_per_country %>% 
  ggplot(aes(x=country, y=av_biomass))+geom_bar(stat = "identity")
fig1

```

```{r}
model2 <- lm(av_biomass~country,data = biomass_per_country) 
#analysing variables
anova(model2)
summary(model2)
```

## Load data

Using 2018 dataset

```{r}
#| label: load data
#| echo: false
#| warning: false
#| message: false

fish_data_2018 <- readr::read_rds(here::here("processed_data", "clean_fish_2018.rds"))
location_2018 <- readr::read_rds(here::here("processed_data", "clean_location_2018.rds"))


```

## Exploring the data

```{r}

#| label: explore data
#| echo: false
#| warning: false
#| message: false
# Overview of the data
head(fish_data_2018) # see first three rows
str(fish_data_2018) #structure of data
summary(fish_data_2018) #summary of data


```

## Biomass

This is ...

## Mean Biomass per site

This is the average of the weight of all individuals per unit area

To do this we must find the mean of across the station transects

```{r}
#libraries
library(dplyr)

# find the mean per station first
station_mean <- fish_data_2018 %>%
  group_by(station) %>%
  summarise(
    station_mean = mean(total_wt_g, na.rm = TRUE),
    site = first(site),
    country = first(country),
    .groups = "drop"
  ) %>% 
  left_join(
    location_2018 %>%
      dplyr::select(station, latitude, longitude),
    by = "station"
  )

#then find the site mean
site_mean <- station_mean %>%
  group_by(site) %>%
  summarise(
    site_mean = mean(station_mean, na.rm = TRUE),
    latitude = mean(latitude, na.rm = TRUE),
    longitude = mean(longitude, na.rm = TRUE),
    country = first(country),
    .groups = "drop"
  )

```

## Plot Mean Site Biomass

```{r}
library(ggplot2)
library(dplyr)
library(sf)
library(rnaturalearth)
library(rnaturalearthdata)

# 3. Build a named color vector (guaranteed to match the data)
country_levels <- sort(unique(site_mean$country))
# provide preferred base colours for known countries, fallback to palette if more levels appear
base_colours <- c("#1b9e77", "#d95f02", "#7570b3")
if (length(country_levels) <= length(base_colours)) {
  fill_vals <- setNames(base_colours[seq_along(country_levels)], country_levels)
} else {
  fill_vals <- setNames(
    grDevices::colorRampPalette(base_colours)(length(country_levels)),
    country_levels
  )
}

# map data (assume world/region already created)
world <- ne_countries(scale = "medium", returnclass = "sf")
region <- world %>% filter(admin %in% c("Kenya", "Tanzania", "Comoros"))


# 4. Plot with halo (larger black points underneath + colored points on top)
map_plot_2009 <- ggplot() +
  geom_sf(
    data = world[world$continent == "Africa", ],
    fill = "grey95", colour = "grey70", linewidth = 0.25
  ) +
  geom_sf(data = region, fill = "grey88", colour = "grey40", linewidth = 0.35) +

  # halo layer: slightly larger than colored points
  geom_point(
    data = site_mean,
    aes(x = longitude, y = latitude, size = site_mean),
    shape = 21, fill = "black", colour = NA, alpha = 0.9, show.legend = FALSE
  ) +

  # colored layer
  geom_point(
    data = site_mean,
    aes(x = longitude, y = latitude, size = site_mean, fill = country),
    shape = 21, colour = "black", stroke = 0.25, alpha = 0.95
  ) +

  scale_fill_manual(values = fill_vals, na.value = "grey50", name = "Country") +
  scale_size_area(name = "Mean site biomass (g)", max_size = 5) +

  coord_sf(xlim = c(37.0, 47.8), ylim = c(-15.5, 1.5)) +
  labs(title = "Mean Fish Biomass per Survey Site",
       subtitle = "Point area = mean biomass; fill = country",
       x = "Longitude", y = "Latitude") +
  theme_minimal(base_size = 1) +
  theme(panel.background = element_rect(fill = "aliceblue", colour = NA),
        panel.grid.major = element_line(colour = "grey90", linewidth = 0.2),
        legend.position = "right")

# draw
map_plot_2009

```

## Mean Biomass per country

```{r}

#then find the country mean
country_mean <- site_mean %>%
  group_by(country) %>%
  summarise(
    country_mean = mean(site_mean, na.rm = TRUE),
    latitude = mean(latitude, na.rm = TRUE),
    longitude = mean(longitude, na.rm = TRUE),
    .groups = "drop"
    )

```

### Plot Mean Country Biomass

...

## Total Biomass per country

```{r}
#then find the country mean
country_totals <- site_mean %>%
  group_by(country) %>%
  summarise(country_mean = sum(site_mean, na.rm = TRUE))
```
