---
title: "perliminary_analyses"
format: html
editor: visual
---

# Overview

This document explores the 2009 dataset. This data ...

# R Set-up

*Important note: Comprehensive instructions, including all prerequisites and detailed guidelines for running this analysis and relevant code, are available in the README file of my [GitHub repository](...) for this project. Please refer to the README ... to ensure proper setup and usage.*

Load the necessary packages (these can be found in ...)

Use the renv::restore() to load up the necessary package versions along with the necessary dependencies.

```{r}
#| label: r set-up
#| echo: false
#| include: false

#Load require libraries
library(tidyverse)
library(arm)
library(here)
library(dplyr)
library(tidyverse)
library(ggplot2)
library(ggfortify) # extends ggplot2 for some more ease with stats
library(MASS) # fits regression models etc.

```

```{r}
#| label: renv-restore
renv::restore()
```

# 2009-2015 data

## Load data for 2009-2015

```{r}
#| label: load-data

fish_data_2009 <- readr::read_rds(here::here("processed_data", "clean_fish_2009.rds"))
location_2009 <- readr::read_rds(here::here("processed_data", "clean_location_2009.rds"))
sst_2009 <- read.csv(here::here("processed_data", "locations_with_sst_2009.csv"))
chla_2009 <- read.csv(here::here("processed_data", "locations_with_chla_2009.csv"))
h_gravity_2009 <- read.csv(here::here("processed_data", "locations_with_gravity.csv"))
```

##join reef geomoirphology

```{r}
location_subset <- location_2009[, c("station", "reef_geomorphology")]
fish_2009 <- fish_data_2009 %>%
  left_join(location_subset, by = "station")
```

## Total Biomass

This refers to an aggregate of all species.

### Calculate total biomass per station, site, and country

```{r}
#| label: calculate-total-data

# Mean biomass per station
biomass_per_station <- fish_2009 %>%
  group_by(station) %>%
  summarise(
    mean_tot_biomass = mean(tot_wt_g, na.rm = TRUE),
    site = first(site),
    location = first(location),
    geomorphology = first(reef_geomorphology),
    country = first(country), 
    .groups = "drop")

# Mean biomass per site
biomass_per_site <- biomass_per_station %>%
  group_by(site) %>%
  summarise(
    mean_biomass = mean(mean_tot_biomass),
    location = first(location),
    geomorphology = first(geomorphology),
    country = first(country))

# Mean biomass per country
biomass_per_country <- biomass_per_site %>%
  group_by(country) %>%
  summarise(
    mean_biomass = mean(mean_biomass, na.rm = TRUE))

# Mean biomass per geomorphology
biomass_per_geomorphology <- biomass_per_site %>%
  group_by(geomorphology) %>%
  summarise(
    mean_biomass = mean(mean_biomass, na.rm = TRUE))

```

### Calculate env. variable means

```{r}
#| label: calculate-env-var-means

# Calculate mean chla per site
chla_2009_sites <- chla_2009 %>% 
  group_by(site) %>% 
  summarise(
    mean_annual_chla = mean(chla_annual_mean))

# Calculate mean SST per site
sst_2009_sites <- sst_2009 %>% 
  group_by(site) %>% 
  summarise(
    mean_annual_sst = mean(sst_annual_mean))

# Calculate mean human gravity per site
h_gravity_2009_sites <- h_gravity_2009 %>% 
  group_by(site) %>% 
  summarise(
    mean_gravity = mean(Grav_tot))
```

### Linear regressions

#### Biomass vs Chlorophyll-a

```{r}

# Overview of the data
head(chla_2009_sites,3) # see first three rows
str(chla_2009_sites) #structure of data
summary(chla_2009_sites) #summary of data
```
Outputs:
- 30 sites
- right-skewed distribution:
  mean>median and upper tail extends to 3.24 (which is then ~4x median)

This suggetss that chlorophyll-a follows a log-normal or gamma-like distribution

Visualise:
```{r}
library(ggplot2)

ggplot(chla_2009_sites, aes(x = mean_annual_chla)) +
  geom_histogram(bins = 20) +
  labs(
    x = expression("Mean annual chlorophyll-a (mg m"^{-3}*")"),
    y = "Number of sites"
  ) +
  theme_bw()

```
Looks like log-noraml distibuton maybe gamma

To do for chlor-a:
plot these...
plot the spaital scale 
test fir spatial autocorrekation


##### See auto-correlation
...

##### Plot

```{r}
#| label: biomass-vs-chla-plot

# Graph the data - X-Y scatter plot
biomass_chla_fig1 <- merge(x = biomass_per_site, y = chla_2009_sites, by = "site", all = TRUE) %>% 
  ggplot(aes(x=mean_annual_chla, y=log(mean_biomass))) + 
  geom_point() +
  geom_smooth(method=lm)

biomass_chla_fig1


```

##### Simple linear regression

```{r}
#| label: biomass-vs-chla-lm

# Merge needed data
chla_x_and_y_site <- merge(x = biomass_per_site, y = chla_2009_sites, by = "site", all = TRUE)

# Build lm 
lm_biomass_chla <- lm(mean_biomass~mean_annual_chla,data = chla_x_and_y_site) 
```
Analyse model:

```{r}
# Analyse the model 
coef(lm_biomass_chla)
#residuals(lm_biomass_chla)
summary(lm_biomass_chla) 
# or arm::display(lm_biomass_chla)

confint(lm_biomass_chla) 

```
Doesn't include 0 so chlr a seems to have an effect on fish bioamss
But... 

model assumptiosna dn vioaltions:
```{r}
par(mfrow= c(1,2)) # divide the page into 1 x 2 frame
plot( lm_biomass_chla, which= c(1:2))
```
Really not very good...
The residuals see, to be going down 
and the vairacne increasses with mean...

##### Log transformation
```{r}
# lm with log transformation
lm_biomass_chla_log <- lm(log(mean_biomass)~mean_annual_chla, data = chla_x_and_y_site) 

# plot model
(lm_biomass_chla_log_plot <- qplot(mean_annual_chla, log(mean_biomass), data= chla_x_and_y_site, geom= c("point", "smooth"), method= "lm", formula= y~x, main= "Linear regression, log data") +
    theme_bw())

```
```{r}
par(mfrow= c(1,2)) # divide the page into 1 x 2 frame
plot( lm_biomass_chla_log, which= c(1:2))
```
Residuals are definitely better but they still follow a strange down shape...

### Lme

model 1: locations
```{r}
lmer_model_1 <- lmer(mean_biomass ~ mean_annual_chla +
    (1 | location), data = chla_x_and_y_site)

summary(lmer_model_1) 
```
- Too few locations or very unbalanced data
- Little between-location variation in mean biomass after aggregation
- Biomass already averaged at the location level → no replication left
- Location effect is absorbed by chl-a (collinearity)

```{r}

```


model 2: geomorphologies
```{r}
lmer_model_2 <- lmer(mean_biomass ~ mean_annual_chla +
    (1 | geomorphology), data = chla_x_and_y_site)
summary(lmer_model_2) 
```

model 3: country 
```{r}
lmer_model_3 <- lmer(mean_biomass ~ mean_annual_chla +
    (1 | country), data = chla_x_and_y_site)
summary(lmer_model_3)
```
comparison:


### 


library(lme4)

model_chla <- lmer(
  log(mean_biomass) ~ mean_annual_chla + (1 | country),
  data = merge(biomass_per_site, chla_2009_sites, by = "site")
)

summary(model_chla)

#### Biomass vs SST

##### Plot

```{r}
#| label: biomass-vs-sst-plot

# Graph the data - X-Y scatterplot
biomass_sst_fig1 <- merge(x = biomass_per_site, y = sst_2009_sites, by = "site", all = TRUE) %>% 
  ggplot(aes(x=mean_annual_sst, y=mean_biomass)) + 
  geom_point() +
  geom_smooth(method=lm) 

biomass_sst_fig1


biomass_sst_fig2 <- merge(x = biomass_per_site, y = sst_2009_sites, by = "site", all = TRUE) %>% 
ggplot(aes(mean_annual_sst, log(mean_biomass), colour = country)) +
  geom_point(alpha = 0.6) +
  geom_smooth(method = lm, se = FALSE)
biomass_sst_fig2
```

##log

```{r}
#| label: biomass-vs-sst-plot

# Graph the data - X-Y scatterplot
biomass_sst_fig2 <- merge(x = biomass_per_site, y = sst_2009_sites, by = "site", all = TRUE) %>% 
  ggplot(aes(x=mean_annual_sst, y=log(mean_biomass))) + 
  geom_point() +
  geom_smooth(method=lm) 

biomass_sst_fig2
```

```{r}
par(mfrow= c(1,2)) # divide the page into 1 x 2 frame
plot( lm_biomass_chla, which= c(1:2))
```



##### Model

```{r}
#| label: biomass-vs-sst-lm

# Merge needed data
sst_x_and_y_site <- merge(x = biomass_per_site, y = sst_2009_sites, by = "site", all = TRUE)

# Build lm 
lm_biomass_sst <- lm(mean_biomass~mean_annual_sst,data = sst_x_and_y_site) 

# Analyse the model 
summary(lm_biomass_sst) 
```

## Log

```{r}

# Merge needed data
sst_x_and_y_site <- merge(x = biomass_per_site, y = sst_2009_sites, by = "site", all = TRUE)

# Build lm 
lm_biomass_sst <- lm(log(mean_biomass)~mean_annual_sst,data = sst_x_and_y_site) 

# Analyse the model 
summary(lm_biomass_sst)


library(lme4)

model_sst <- lmer(
  log(mean_biomass) ~ mean_annual_sst + (1 | country),
  data = merge(biomass_per_site, sst_2009_sites, by = "site")
)

summary(model_sst)
```

#### Biomass vs Gravity

##### Plot

```{r}
#| label: biomass-vs-gravity-plot

# Graph the data - X-Y scatterplot
biomass_gravity_fig1 <- merge(x = biomass_per_site, y = h_gravity_2009_sites, by = "site", all = TRUE) %>% 
  ggplot(aes(x=mean_gravity, y=mean_biomass)) + 
  geom_point() +
  geom_smooth(method=lm) 

biomass_gravity_fig1

biomass_gravity_fig2 <- merge(x = biomass_per_site, y = h_gravity_2009_sites, by = "site", all = TRUE) %>% 
ggplot(aes(mean_gravity, log(mean_biomass), colour = country)) +
  geom_point(alpha = 0.6) +
  geom_smooth(method = lm, se = FALSE)
biomass_gravity_fig2
```

## log

```{r}
#| label: biomass-vs-gravity-plot

# Graph the data - X-Y scatterplot
biomass_gravity_fig2 <- merge(x = biomass_per_site, y = h_gravity_2009_sites, by = "site", all = TRUE) %>% 
  ggplot(aes(x=mean_gravity, y=log(mean_biomass))) + 
  geom_point() +
  geom_smooth(method=lm) 

biomass_gravity_fig2
```

##### Model

```{r}
#| label: biomass-vs-gravity-lm

# Merge needed data
gravity_x_and_y_site <- merge(x = biomass_per_site, y = h_gravity_2009_sites, by = "site", all = TRUE)

# Build lm 
lm_biomass_gravity <- lm(mean_biomass~mean_gravity,data = gravity_x_and_y_site) 
# Analyse the model 
summary(lm_biomass_gravity) 

```

### Log

```{r}
#| label: biomass-vs-gravity-lm

# Merge needed data
gravity_x_and_y_site <- merge(x = biomass_per_site, y = h_gravity_2009_sites, by = "site", all = TRUE)

# Build lm 
lm_biomass_gravity <- lm(log(mean_biomass)~mean_gravity,data = gravity_x_and_y_site) 
# Analyse the model 
summary(lm_biomass_gravity) 

```

### accountign for country

```{r}

library(lme4)

model_sst <- lmer(
  log(mean_biomass) ~ mean_gravity + (1 | country),
  data = gravity_x_and_y_site)

summary(model_sst)
```

```{r}
# fix
analysis_2009 <- biomass_per_site %>%
  left_join(location_subset, by = "site") %>%
  left_join(chla_x_and_y_site, by = "site") %>%
  left_join(sst_x_and_y_site, by = "site") %>%
  left_join(gravity_x_and_y_site,  by = "site")


m_fixed <- lm(log(biomass) ~ sst + chla + gravity, data = data)

#vs
m_mixed <- lmer(log(biomass) ~ sst + chla + gravity + (1 | country), data = data)


```

# 2018 data (to do)

## Load data

Using 2018 dataset

```{r}
#| label: load data
#| echo: false
#| warning: false
#| message: false

fish_data_2018 <- readr::read_rds(here::here("processed_data", "clean_fish_2018.rds"))
location_2018 <- readr::read_rds(here::here("processed_data", "clean_location_2018.rds"))


```

## Exploring the data

```{r}

#| label: explore data
#| echo: false
#| warning: false
#| message: false
# Overview of the data
head(fish_data_2018) # see first three rows
str(fish_data_2018) #structure of data
summary(fish_data_2018) #summary of data


```

## Biomass

This is ...

## Mean Biomass per site

This is the average of the weight of all individuals per unit area

To do this we must find the mean of across the station transects

```{r}
#libraries
library(dplyr)

# find the mean per station first
station_mean <- fish_data_2018 %>%
  group_by(station) %>%
  summarise(
    station_mean = mean(total_wt_g, na.rm = TRUE),
    site = first(site),
    country = first(country),
    .groups = "drop"
  ) %>% 
  left_join(
    location_2018 %>%
      dplyr::select(station, latitude, longitude),
    by = "station"
  )

#then find the site mean
site_mean <- station_mean %>%
  group_by(site) %>%
  summarise(
    site_mean = mean(station_mean, na.rm = TRUE),
    latitude = mean(latitude, na.rm = TRUE),
    longitude = mean(longitude, na.rm = TRUE),
    country = first(country),
    .groups = "drop"
  )

```

## Plot Mean Site Biomass

```{r}
library(ggplot2)
library(dplyr)
library(sf)
library(rnaturalearth)
library(rnaturalearthdata)

# 3. Build a named color vector (guaranteed to match the data)
country_levels <- sort(unique(site_mean$country))
# provide preferred base colours for known countries, fallback to palette if more levels appear
base_colours <- c("#1b9e77", "#d95f02", "#7570b3")
if (length(country_levels) <= length(base_colours)) {
  fill_vals <- setNames(base_colours[seq_along(country_levels)], country_levels)
} else {
  fill_vals <- setNames(
    grDevices::colorRampPalette(base_colours)(length(country_levels)),
    country_levels
  )
}

# map data (assume world/region already created)
world <- ne_countries(scale = "medium", returnclass = "sf")
region <- world %>% filter(admin %in% c("Kenya", "Tanzania", "Comoros"))


# 4. Plot with halo (larger black points underneath + colored points on top)
map_plot_2009 <- ggplot() +
  geom_sf(
    data = world[world$continent == "Africa", ],
    fill = "grey95", colour = "grey70", linewidth = 0.25
  ) +
  geom_sf(data = region, fill = "grey88", colour = "grey40", linewidth = 0.35) +

  # halo layer: slightly larger than colored points
  geom_point(
    data = site_mean,
    aes(x = longitude, y = latitude, size = site_mean),
    shape = 21, fill = "black", colour = NA, alpha = 0.9, show.legend = FALSE
  ) +

  # colored layer
  geom_point(
    data = site_mean,
    aes(x = longitude, y = latitude, size = site_mean, fill = country),
    shape = 21, colour = "black", stroke = 0.25, alpha = 0.95
  ) +

  scale_fill_manual(values = fill_vals, na.value = "grey50", name = "Country") +
  scale_size_area(name = "Mean site biomass (g)", max_size = 5) +

  coord_sf(xlim = c(37.0, 47.8), ylim = c(-15.5, 1.5)) +
  labs(title = "Mean Fish Biomass per Survey Site",
       subtitle = "Point area = mean biomass; fill = country",
       x = "Longitude", y = "Latitude") +
  theme_minimal(base_size = 1) +
  theme(panel.background = element_rect(fill = "aliceblue", colour = NA),
        panel.grid.major = element_line(colour = "grey90", linewidth = 0.2),
        legend.position = "right")

# draw
map_plot_2009

```

## Mean Biomass per country

```{r}

#then find the country mean
country_mean <- site_mean %>%
  group_by(country) %>%
  summarise(
    country_mean = mean(site_mean, na.rm = TRUE),
    latitude = mean(latitude, na.rm = TRUE),
    longitude = mean(longitude, na.rm = TRUE),
    .groups = "drop"
    )

```

### Plot Mean Country Biomass

...

## Total Biomass per country

```{r}
#then find the country mean
country_totals <- site_mean %>%
  group_by(country) %>%
  summarise(country_mean = sum(site_mean, na.rm = TRUE))
```
