---
title: "Extracting Environmental Data from ERDDAP"
author: "Candela Ferrer Diez"
date: 2025-11-06
format: 
  html:
    toc: true
    toc-depth: 2
    code-fold: false
    theme: cosmo
    embed-resources: true
editor: visual
execute:
  warning: false
  message: false
---

## R Set-up

Load the necessary packages for extracting satellite data from ERDDAP servers.

*Important note: Comprehensive instructions, including all prerequisites and detailed guidelines for running this analysis and relevant code, are available in the README file of my [GitHub repository](...) for this project. Please refer to the README to ensure proper setup and usage.*

```{r}
#| label: r-setup
#| echo: false
#| include: false

library(rerddap)
library(dplyr)
library(here)
library(ggplot2)
```

...

```{r}
#| label: renv-restore
#| eval: false

renv::restore()
```

## Load Location Data

Load the cleaned location data containing survey sites with coordinates and dates.

```{r}
#| label: load-location-data

# Load location data
clean_location_2018 <- read.csv(here::here("processed_data", "clean_location_2018.csv"))

```

## Get Dataset Information

Connect to ERDDAP and retrieve information about the MODIS chlorophyll dataset.

```{r}
#| label: dataset-info

# Get information about the MODIS chlorophyll dataset
chla_info <- info('erdMH1chlamday_Lon0360')

# View dataset details
print(chla_info)
```

## Determine Spatial and Temporal Bounds

Calculate the bounding box and time range from your survey data to understand what data needs to be extracted.

```{r}
#| label: calculate-bounds

# Calculate spatial bounds from your data
lat_range <- range(clean_location_2018$latitude, na.rm = TRUE)
lon_range <- range(clean_location_2018$longitude, na.rm = TRUE)
time_range <- range(clean_location_2018$date, na.rm = TRUE)

print(paste("Latitude range:", lat_range[1], "to", lat_range[2]))
print(paste("Longitude range:", lon_range[1], "to", lon_range[2]))
print(paste("Time range:", time_range[1], "to", time_range[2]))


# Note: This dataset uses Lon0360 (0-360°), so convert if needed
# Western Indian Ocean: roughly 40-80°E becomes 40-80 in Lon0360 system

# Check if your longitudes need conversion
if(any(clean_location_2018$longitude < 0)) {
  print("⚠ Warning: Some longitudes are negative. Converting to 0-360 system...")
  clean_location_2018$longitude_360 <- ifelse(
    clean_location_2018$longitude < 0,
    clean_location_2018$longitude + 360,
    clean_location_2018$longitude
  )
} else {
  clean_location_2018$longitude_360 <- clean_location_2018$longitude
}
```

## Extract Chlorophyll Data

Extract monthly chlorophyll-a values for each survey site. This dataset provides monthly composites, which helps reduce cloud gaps in the data.

```{r}
#| label: extract-chlorophyll

# Initialise column to store chlorophyll values
clean_location_2018$chla_monthly <- NA_real_

# Clear cache to ensure fresh data
cache_delete_all()

# Extract chlorophyll for each station
for(i in 1:nrow(clean_location_2018)) {
  
  # Get coordinates and date for this station
  station_lat <- clean_location_2018$latitude[i]
  station_lon <- clean_location_2018$longitude_360[i]
  station_date <- as.character(clean_location_2018$date[i])
  
  # Extract chlorophyll with error handling
  tryCatch({
    chla_extract <- griddap(
      chla_info,
      latitude = c(station_lat, station_lat),
      longitude = c(station_lon, station_lon),
      time = c(station_date, station_date),
      fields = 'chlorophyll'
    )
    
    # Store the extracted value
    if(nrow(chla_extract$data) > 0) {
      clean_location_2018$chla_monthly[i] <- chla_extract$data$chlorophyll[1]
    }
    
  }, error = function(e) {
    warning(paste("Chlorophyll extraction failed for row", i, 
                  "(", station_date, station_lat, station_lon, "):", e$message))
  })
  
  # Progress indicator every 50 sites
  if(i %% 50 == 0) {
    print(paste("Processed", i, "of", nrow(clean_location_2018), "stations"))
  }
}

print("✓ Chlorophyll extraction complete")
```

## Check Extraction Results

Validate the extracted chlorophyll data and identify any issues.

```{r}
#| label: check-results

# Summary statistics
print("Chlorophyll-a Summary:")
summary(clean_location_2018$chla_monthly)

# Count missing values
missing_count <- sum(is.na(clean_location_2018$chla_monthly))
print(paste("Missing chlorophyll values:", missing_count, 
            "out of", nrow(clean_location_2018), 
            "(", round(100*missing_count/nrow(clean_location_2018), 1), "%)"))

# Check distribution by location
chla_by_site <- clean_location_2018 %>%
  group_by(site) %>%
  summarise(
    n_sites = n(),
    mean_chla = mean(chla_monthly, na.rm = TRUE),
    median_chla = median(chla_monthly, na.rm = TRUE),
    missing = sum(is.na(chla_monthly)),
    .groups = "drop"
  )

print("Chlorophyll by Location:")
print(chla_by_site)

# Identify sites with missing data
missing_sites <- clean_location_2018 %>%
  filter(is.na(chla_monthly)) %>%
  select(station, site, location, date, latitude, longitude)

if(nrow(missing_sites) > 0) {
  print("Sites with missing chlorophyll data:")
  print(head(missing_sites, 10))
}
```

## Visualisze Chlorophyll Distribution

Create maps and plots to visualize the extracted chlorophyll data.

```{r}
#| label: visualise-chlorophyll
#| fig-width: 10
#| fig-height: 8

# Map of chlorophyll values
library(ggplot2)
library(mapdata)

# Get appropriate bounds for WIO
lon_bounds <- range(clean_location_2018$longitude, na.rm = TRUE)
lat_bounds <- range(clean_location_2018$latitude, na.rm = TRUE)

# Get coastline
coast <- map_data("worldHires", 
                  xlim = lon_bounds, 
                  ylim = lat_bounds)

# Map of chlorophyll
ggplot() +
  geom_polygon(data = coast, 
               aes(x = long, y = lat, group = group), 
               fill = "grey80") +
  geom_point(data = filter(clean_location_2018, !is.na(chla_monthly)),
             aes(x = longitude, y = latitude, color = log10(chla_monthly)),
             size = 2, alpha = 0.7) +
  scale_color_viridis_c(name = "log10(Chl-a)\n(mg/m³)") +
  facet_wrap(~country) +
  theme_bw() +
  coord_fixed(1.3, xlim = lon_bounds, ylim = lat_bounds) +
  labs(title = "Monthly Chlorophyll-a at Survey Sites",
       subtitle = "MODIS Aqua 4km Monthly Composite",
       x = "Longitude", y = "Latitude")

```

## Analyze Data Coverage by Country

Examine which sites successfully extracted chlorophyll data and which did not, broken down by country.

```{r}
#| label: analyze-coverage-by-country

# Create a flag for data availability
clean_location_2018 <- clean_location_2018 %>%
  mutate(has_chla = ifelse(is.na(chla_monthly), "No Data", "Has Data"))

# Summary by country
coverage_by_country <- clean_location_2018 %>%
  group_by(country, has_chla) %>%
  summarise(
    n_stations = n(),
    n_sites = n_distinct(site),
    .groups = "drop"
  ) %>%
  pivot_wider(
    names_from = has_chla,
    values_from = c(n_stations, n_sites),
    values_fill = 0
  ) %>%
  mutate(
    total_stations = `n_stations_Has Data` + `n_stations_No Data`,
    total_sites = `n_sites_Has Data` + `n_sites_No Data`,
    pct_stations_with_data = round(100 * `n_stations_Has Data` / total_stations, 1),
    pct_sites_with_data = round(100 * `n_sites_Has Data` / total_sites, 1)
  )

print("Coverage Summary by Country:")
print(coverage_by_country)

# Detailed breakdown by country and location
coverage_by_location <- clean_location_2018 %>%
  group_by(country, location, has_chla) %>%
  summarise(
    n_stations = n(),
    .groups = "drop"
  ) %>%
  pivot_wider(
    names_from = has_chla,
    values_from = n_stations,
    values_fill = 0
  ) %>%
  mutate(
    total = `Has Data` + `No Data`,
    pct_with_data = round(100 * `Has Data` / total, 1)
  ) %>%
  arrange(country, desc(`No Data`))

print("\nCoverage by Country and Location:")
print(coverage_by_location)

# List all stations with no data
stations_no_data <- clean_location_2018 %>%
  filter(is.na(chla_monthly)) %>%
  select(country, location, site, station, date, latitude, longitude) %>%
  arrange(country, location, site)

print("\nStations with No Chlorophyll Data:")
print(stations_no_data)

# Check temporal patterns in missing data
temporal_coverage <- clean_location_2018 %>%
  mutate(year = format(date, "%Y"),
         month = format(date, "%m")) %>%
  group_by(country, year, month, has_chla) %>%
  summarise(n = n(), .groups = "drop") %>%
  pivot_wider(
    names_from = has_chla,
    values_from = n,
    values_fill = 0
  ) %>%
  mutate(total = `Has Data` + `No Data`,
         pct_with_data = round(100 * `Has Data` / total, 1))

print("\nTemporal Coverage Pattern:")
print(temporal_coverage)
```

## Visualize Data Coverage

Create visual summaries of data availability by country and location.

```{r}
#| label: visualize-coverage
#| fig-width: 12
#| fig-height: 8

library(ggplot2)

# Bar chart: Coverage by country
coverage_plot <- ggplot(clean_location_2018, aes(x = country, fill = has_chla)) +
  geom_bar(position = "stack") +
  scale_fill_manual(values = c("No Data" = "#d95f02", "Has Data" = "#1b9e77"),
                    name = "Data Status") +
  theme_bw() +
  labs(title = "Chlorophyll Data Coverage by Country",
       x = "Country",
       y = "Number of Stations") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

print(coverage_plot)

# Percentage stacked bar chart
pct_coverage_plot <- ggplot(clean_location_2018, aes(x = country, fill = has_chla)) +
  geom_bar(position = "fill") +
  scale_fill_manual(values = c("No Data" = "#d95f02", "Has Data" = "#1b9e77"),
                    name = "Data Status") +
  scale_y_continuous(labels = scales::percent) +
  theme_bw() +
  labs(title = "Chlorophyll Data Coverage by Country (%)",
       x = "Country",
       y = "Percentage of Stations") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

print(pct_coverage_plot)

# Coverage by location within each country
location_plot <- clean_location_2018 %>%
  group_by(country, location) %>%
  mutate(location_label = paste0(location, "\n(n=", n(), ")")) %>%
  ggplot(aes(x = location_label, fill = has_chla)) +
  geom_bar(position = "stack") +
  scale_fill_manual(values = c("No Data" = "#d95f02", "Has Data" = "#1b9e77"),
                    name = "Data Status") +
  facet_wrap(~country, scales = "free_x") +
  theme_bw() +
  labs(title = "Chlorophyll Data Coverage by Location",
       x = "Location",
       y = "Number of Stations") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1, size = 8))

print(location_plot)
```

## Map of Data Coverage

Create a spatial map showing which sites have data and which don't.

```{r}
#| label: map-coverage
#| fig-width: 12
#| fig-height: 8

library(mapdata)

# Get coastline
lon_bounds <- range(clean_location_2018$longitude, na.rm = TRUE)
lat_bounds <- range(clean_location_2018$latitude, na.rm = TRUE)
coast <- map_data("worldHires", xlim = lon_bounds, ylim = lat_bounds)

# Map showing data coverage
coverage_map <- ggplot() +
  geom_polygon(data = coast, 
               aes(x = long, y = lat, group = group), 
               fill = "grey80") +
  geom_point(data = clean_location_2018,
             aes(x = longitude, y = latitude, 
                 color = has_chla, shape = has_chla),
             size = 3, alpha = 0.7) +
  scale_color_manual(values = c("No Data" = "#d95f02", "Has Data" = "#1b9e77"),
                     name = "Data Status") +
  scale_shape_manual(values = c("No Data" = 4, "Has Data" = 16),
                     name = "Data Status") +
  facet_wrap(~country) +
  theme_bw() +
  coord_fixed(1.3, xlim = lon_bounds, ylim = lat_bounds) +
  labs(title = "Spatial Distribution of Chlorophyll Data Coverage",
       subtitle = "X = No data available, • = Data available",
       x = "Longitude", y = "Latitude")

print(coverage_map)

# Single map with all countries
all_coverage_map <- ggplot() +
  geom_polygon(data = coast, 
               aes(x = long, y = lat, group = group), 
               fill = "grey90") +
  geom_point(data = clean_location_2018,
             aes(x = longitude, y = latitude, 
                 color = has_chla, shape = has_chla),
             size = 2.5, alpha = 0.8) +
  scale_color_manual(values = c("No Data" = "#d95f02", "Has Data" = "#1b9e77"),
                     name = "Data Status") +
  scale_shape_manual(values = c("No Data" = 4, "Has Data" = 16),
                     name = "Data Status") +
  theme_bw() +
  coord_fixed(1.3, xlim = lon_bounds, ylim = lat_bounds) +
  labs(title = "Chlorophyll Data Coverage Across All Sites",
       subtitle = paste("Total:", nrow(clean_location_2018), "stations |",
                       sum(!is.na(clean_location_2018$chla_monthly)), "with data |",
                       sum(is.na(clean_location_2018$chla_monthly)), "without data"),
       x = "Longitude", y = "Latitude") +
  theme(legend.position = "bottom")

print(all_coverage_map)
```

## Export Coverage Summary

Save the coverage analysis for reference.

```{r}
#| label: export-coverage-summary

# Export coverage summaries
write.csv(coverage_by_country,
          here::here("processed_data", "chla_coverage_by_country.csv"),
          row.names = FALSE)

write.csv(coverage_by_location,
          here::here("processed_data", "chla_coverage_by_location.csv"),
          row.names = FALSE)

write.csv(stations_no_data,
          here::here("processed_data", "stations_without_chla.csv"),
          row.names = FALSE)

print("✓ Coverage summaries exported successfully!")
```

## Export Results

Save the enriched dataset with chlorophyll data.

```{r}
#| label: export-results

# Export data with chlorophyll values
write.csv(clean_location_2018,
          here::here("processed_data", "clean_location_2018_with_chla.csv"),
          row.names = FALSE)

print("✓ Data exported successfully!")
print(paste("File saved:", here::here("processed_data", "clean_location_2018_with_chla.csv")))
```

## Notes on Data Quality

**Important considerations:**

-   This is a **monthly composite** product, which reduces cloud gaps but may not match your exact survey date
-   The spatial resolution is **4km**, so nearby sites will have similar values
-   Missing values (NA) can occur due to:
    -   Cloud cover
    -   Land contamination near coastlines
    -   Instrument issues
    -   Data falling outside the dataset's temporal coverage (2003-present)

**To handle missing data**, consider: 1. Using temporal windows (e.g., ±15 days around survey date) 2. Using 8-day composites instead of monthly for better temporal matching 3. Spatial averaging over a small window rather than single pixel extraction


## Try a different method.
