---
title: "tot_biomass_model_testing"
format: html
editor: visual
---

# Overview

This document explores the relationship between human gravity and fish biomass

# R Set-up

*Important note: Comprehensive instructions, including all prerequisites and detailed guidelines for running this analysis and relevant code, are available in the README file of my [GitHub repository](...) for this project. Please refer to the README ... to ensure proper setup and usage.*

```{r}
#| label: renv-restore
# renv::restore()
```

Use the renv::restore() to load up the necessary package versions along with the necessary dependencies.

```{r}
#| label: r set-up
#| echo: false
#| include: false

#Load require libraries
library(tidyverse)
library(sf)
library(arm)
library(here)
library(dplyr)
library(tidyr)
library(ggplot2)
library(ggfortify) # extends ggplot2 for some more ease with stats
library(MASS) # fits regression models etc.
library(MuMIn) # multi-model inference (AIC etc.)
library(AICcmodavg)
library(DHARMa)
library(tweedie)
library(glmmTMB)
library(statmod)
```

## Load data

```{r}
#| label: load-data

fish_2009 <- readr::read_rds(here::here("processed_data", "clean_fish_2009.rds"))
location_2009 <- readr::read_rds(here::here("processed_data", "clean_location_2009.rds"))
gravity_2009 <- readr::read_rds(here::here("city_data", "locations_with_grav_combined.rds"))
gravity_2009_2<- readr::read_rds(here::here("city_data", "locations_with_grav.rds"))
old_grav_2009 <- read.csv(here::here("processed_data", "locations_with_gravity.csv"))
chla_2009 <- read.csv(here::here("processed_data", "locations_with_chla_2009.csv"))
dhw_2009 <- readr::read_rds(here::here("processed_data", "locations_with_dhw_2009.rds"))
```

# Total Biomass

This refers to an aggregate of all species.

### Calculate total biomass per station, site, and country

Biomass was quantified at the transect level, with each transect representing a replicate sample within a site. Because transects within a site are spatially proximate and not statistically independent, sites were treated as the independent units in all analyses. Need to explain the stations as well as many metrics are taken using such coordinates!!

Environmental predictors were extracted at station coordinates and then averaged to the site level; stations were grouped into sites if they fell within 2 km of each other. Because environmental rasters are at \~4 km resolution, centroid-based extraction produced effectively identical values in our data (see diagnostic), so either approach is acceptable — we report station-mean values to preserve the sampled footprint.

Total biomass was first calculated for each transect by summing biomass across species, and site-level biomass was then estimated as the mean of its 3–5 constituent transects. This approach ensures equal weighting of transects within sites and avoids pseudo-replication in comparisons among sites.

Note that I drop those sites that have fewer than 3 replicates - which are ch_mvudeni, lalane, and mongo_shariff

```{r}
#| label: calculate-total-data

# Aggregate biomass at transect level
transect_biomass <- fish_2009 %>%
  group_by(site, station, ts_no) %>%
  summarise(
    transect_biomass = sum(tot_wt_g, na.rm = TRUE),
    country = first(country),
    .groups = "drop")

# Mean biomass per site
biomass_per_site <- transect_biomass %>%
  group_by(site) %>%
  filter(n() >= 3) %>% #remove sites with fewer than 3 replicates
  summarise(
    mean_biomass = mean(transect_biomass, na.rm = TRUE),
    country = first(country),
    .groups = "drop"
  )

biomass_per_site_laura <- transect_biomass %>%
  filter(
    country != "kenya", # remove kenya
    !site %in% c("fernauvloso", "paradise") # remove specific s
    ) %>%   # remove Kenya
  group_by(site) %>%
  filter(n() >= 3) %>%             # remove sites with <3 replicates
  summarise(
    mean_biomass = mean(transect_biomass, na.rm = TRUE),
    country = first(country),
    .groups = "drop"
  )

# Mean biomass per country
biomass_per_country <- biomass_per_site %>%
  group_by(country) %>%
  summarise(
    mean_biomass = mean(mean_biomass, na.rm = TRUE))

```

### Check number of transects per site

There are some that have fewer than 3 - either group or remove. There are also some with more than 5 - check these (maybe split if needs be)

Then check number of sites per country:

```{r}
# Number of transects per site
biomass_per_site %>%
  group_by(country) %>%
  summarise(site = n())

# Number of transects per site
biomass_per_site_laura %>%
  group_by(country) %>%
  summarise(site = n())

```

Note that both of these show that the design is unbalanced. Maybe I should balance this by ensuring an equal number of transects per site.

Also very few sites in Kenya

### Check the structure of biomass data

```{r}
# Overview of the data
head(biomass_per_site,3) # see first three rows
str(biomass_per_site) #structure of data
summary(biomass_per_site) #summary of data

# Histogram 
( biomass_hist <- ggplot(biomass_per_site, aes(x = mean_biomass)) +
  geom_histogram(bins = 20) +
  labs(
    x = expression("Mean site biomass "),
    y = "Frequency"
  ) +
  theme_bw() )

# Boxcox 
boxcox(lm(mean_biomass ~ 1, data = biomass_per_site),
       lambda = seq(-2, 2, 0.1))
```

The biomass seems to follow a moderately right-skewed distribution. Boxcox suggests the data might need to be log transformed.

```{r}
# Overview of the data
head(biomass_per_site_laura,3) # see first three rows
str(biomass_per_site_laura) #structure of data
summary(biomass_per_site_laura) #summary of data

# Histogram 
( biomass_hist <- ggplot(biomass_per_site_laura, aes(x = mean_biomass)) +
  geom_histogram(bins = 20) +
  labs(
    x = expression("Mean site biomass "),
    y = "Frequency"
  ) +
  theme_bw() )

# Boxcox 
boxcox(lm(mean_biomass ~ 1, data = biomass_per_site_laura),
       lambda = seq(-2, 2, 0.1))
```

### Calculate gravity means per site

The gravity values were calculated for each station - as these were the reef sampling locations that were geo-referenced in the data. This code aggregates human‐gravity metrics to the site level by calculating mean values of the stations (excluding missing data) for each variable within sites. Note that the gravity metrics are the updated metrics from 2009, and the old gravity refers to the market gravity extracted from \[site\]. All the metrics are then combined into a table for ease of use.

```{r}
#| label: calculate-env-var-means

# Calculate mean human gravity per site
gravity_2009_sites <- gravity_2009 %>%
  st_drop_geometry() %>% # remove geometry before summarise
  group_by(site) %>%
  summarise(
    market_gravity   = mean(market_grav, na.rm = TRUE),
    settlement_tot_pop = mean(settlement_tot_pop, na.rm = TRUE),
    nearest_pop_grav = mean(nearest_pop75_grav, na.rm = TRUE),
    .groups = "drop"
  )

# Calculate mean human gravity per site
old_grav_2009_sites <- old_grav_2009 %>% 
  group_by(site) %>% 
  summarise(
    old_gravity = mean(Grav_tot))

# Combine gravity metrics
gravity_metrics_2009 <- gravity_2009_sites %>%
  left_join(old_grav_2009_sites, by = "site") %>%
  semi_join(biomass_per_site, by = "site")

# Do the same for laura's site selection
gravity_metrics_2009_laura <- gravity_2009_sites %>%
  left_join(old_grav_2009_sites, by = "site") %>%
  semi_join(biomass_per_site_laura, by = "site")

# Overview of the data
head(gravity_metrics_2009,3) # see first three rows
str(gravity_metrics_2009) #structure of data
summary(gravity_metrics_2009) #summary of data
```

# Total Biomass vs Human gravity

## Old market gravity

Firstly we explore the relationship between biomass and the market gravity extracted from \[site\] While this was calculated for \[year\], it has been used in other papers \[such as samoilys, 2025\] and therefore provides a good reference for comparison.

### Explore the structure of the data

We explore the distribution of the gravity variable to assess its suitability for parametric statistical modelling. Visualising the histogram allows us to evaluate distributional properties such as skewness, dispersion, and the presence of outliers, which inform assumptions of normality and homoscedasticity. The Box–Cox analysis formally evaluates whether a power transformation would improve normality and stabilize variance, thereby guiding appropriate transformation of the predictor prior to inclusion in subsequent models.

```{r}

# Histogram market gravity 
( gravity_hist <- ggplot(gravity_metrics_2009, aes(x = old_gravity)) +
  geom_histogram(bins = 10) +
  labs(
    x = expression(" Old market gravity"),
    y = "Number of sites"
  ) +
  theme_bw() )

# Boxcox 
boxcox(lm(old_gravity ~ 1, data = gravity_metrics_2009),
       lambda = seq(-2, 2, 0.1))

```

The histogram shows that the market gravity metric is highly right-skewed, with most sites clustered at low values and a small number of extreme high values (strong positive skew and potential outliers). This indicates a clear deviation from normality and suggests that the raw variable is unlikely to meet linear modelling assumptions.

The Box–Cox analysis indicates an optimal negative λ (≈ −0.5), confirming that a decreasing power transformation is required and that the untransformed variable is inappropriate for parametric modelling. Because λ values near the optimum yield similar likelihoods and log transformations are more interpretable, it is reasonable to compare a log transformation with a reciprocal-type transformation and select the most appropriate form based on subsequent model diagnostics.

### Exploratory Plot

Although prior diagnostics indicated that old_gravity is strongly right-skewed and likely requires transformation, we first fit a simple linear regression using the untransformed predictor as a baseline model. This provides a reference against which transformed models can be formally compared, allowing us to evaluate whether transformation improves model fit, residual behaviour, and inferential robustness.

```{r}
#| label: biomass-vs-gravity-plot

# Graph the data - X-Y scatter plot
biomass_gravity_fig1 <- merge(x = biomass_per_site, y = gravity_metrics_2009, by = "site", all = TRUE) %>%
  ggplot(aes(x=old_gravity, y=mean_biomass)) + 
  geom_point() +
  geom_smooth(method=lm)

biomass_gravity_fig1

# Correlation
with(merge(x = biomass_per_site, y = gravity_metrics_2009, by = "site", all = TRUE),
    cor(old_gravity, mean_biomass))

```

The scatterplot and regression line suggest a weak negative relationship between human gravity and mean biomass, with biomass tending to decline slightly as gravity increases. However, variability is high, confidence intervals are wide—particularly at higher gravity values—and the correlation is small (r = −0.136), indicating that gravity explains little of the observed variation in biomass in the untransformed linear model.

### Untransformed ls model

We fitted a simple linear regression model with mean biomass as the response variable and centred old market gravity as the sole predictor. This represents the most parsimonious, untransformed baseline model against which subsequent transformed models can be compared.

Note that the gravity predictor was mean-centred by subtracting its sample mean from each observation, so that a value of zero corresponds to the average gravity across sites. This reparameterisation does not change the estimated slope or its statistical significance, but it shifts the intercept to represent the expected biomass at average gravity rather than at gravity = 0, thereby improving the interpretability of the results.

```{r}
#| label: biomass-vs-chla-lm

# Merge needed data
og_model_data <- merge(biomass_per_site, gravity_metrics_2009[, c("site", "old_gravity")], by = "site", all = TRUE)

# Centre old market gravity
og_model_data$og_centred <- as.numeric(scale(
  og_model_data$old_gravity,
  center = TRUE,
  scale = FALSE ))

# Build lm 
base_model_og <- lm(mean_biomass~og_centred,data = og_model_data) 
```

Analyse model:

```{r}
# Analyse the model 
coef(base_model_og)
summary(base_model_og) 
confint(base_model_og) 
```

The model indicates a negative but non-significant association between gravity and biomass (β = −3.81 ± 3.59 SE, p = 0.292), with the 95% confidence interval spanning zero. The model explains very little variation in biomass (R² = 0.018), and the overall regression is not statistically significant (F₁,₆₀ = 1.13, p = 0.292). Thus, in its untransformed form, human gravity does not provide meaningful explanatory power for variation in biomass.

#### Checking assumptions

```{r}

# Extracting fitted values and residuals 
og_model_data$predicted <- fitted.values(base_model_og)
og_model_data$residuals <- residuals(base_model_og) 

# Histogram of residuals 
( og_residual_hist <- ggplot(og_model_data, aes(x = residuals)) +
  # Change y to density so the curve matches the scale
  geom_histogram(aes(y = after_stat(density)), bins = 10, fill = "lightgray", color = "white") +
  geom_density(color = "black", size = 1) +
  labs(
    x = expression("Residuals"),
    y = "Number of sites"
  ) +
  theme_bw() )
```

Residual checking of simple ls model To evaluate whether the assumptions of the linear model are satisfied, we examined standard diagnostic plots: residuals versus fitted values, a normal Q–Q plot, a scale–location plot, and Cook’s distance. These diagnostics assess linearity, homoscedasticity, normality of residuals, and the presence of influential observations.

But model assumptions and violations:

```{r}
par(mfrow= c(2,2)) # divide the page into 2 x 2 frame
# par(mar=c(1,1,1,1))
plot( base_model_og, which= c(1:4))
```

The residuals versus fitted plot indicates increasing dispersion at higher fitted values, suggesting heteroscedasticity rather than constant error variance. This pattern is reinforced by the scale–location plot, where residual spread clearly increases with fitted biomass. The Q–Q plot further reveals departures from normality, particularly in the upper tail, consistent with right-skewed residuals and the presence of extreme values. Cook’s distance highlights several moderately influential observations, likely associated with sites exhibiting high gravity values and leverage.

Taken together, these patterns indicate that key assumptions of normality and homoscedasticity are not satisfied under the untransformed specification, supporting the need for transformation before proceeding with formal inference.

### Transformations

Transformations were applied to improve adherence to linear regression assumptions, particularly normality and homoscedasticity, and to reduce the influence of extreme values. After each transformation, the distribution of the variable was re-examined to verify that skewness was reduced and that the transformation behaved as expected. These transformations amplifying predictive power and increasing the overall quality of the model

```{r}

# Transforming variables
og_model_data$log_biomass <- log10(og_model_data$mean_biomass)
og_model_data$log_og <- log10(og_model_data$old_gravity)

# Check transformations 

( log_market_grav_hist <- ggplot(og_model_data, aes(x = log_og)) +
  geom_histogram(bins = 20) +
  labs(
    x = expression("Log-Transformed Old Gravity"),
    y = "Frequency"
  ) +
  theme_bw() )

# Reciprocal square-root transformation (λ ≈ -0.5)
og_model_data$og_recipr_sqrt <- 1 / sqrt(og_model_data$old_gravity)

# Check distribution

( recip_sqrt_market_grav_hist <- ggplot(og_model_data, 
                                        aes(x = og_recipr_sqrt)) +
  geom_histogram(bins = 20) +
  labs(
    x = expression("Reciprocal Square-Root Old Gravity (1 / sqrt(Old Gravity))"),
    y = "Frequency"
  ) +
  theme_bw() )
```

The log-transformed gravity variable exhibits a substantially improved distribution relative to the raw scale. The strong right skew observed in the untransformed data is markedly reduced, and the values are more evenly distributed across the range. Although some mild skewness persists, the transformation compresses extreme high values and reduces leverage from previously influential observations. This suggests that a log transformation is effective in stabilising variance and improving conformity with linear modelling assumptions.

The reciprocal square-root transformation also substantially alters the distribution relative to the raw scale. By strongly compressing large gravity values and stretching differences among smaller values, it reduces the influence of extreme high-gravity observations. The resulting distribution appears more symmetric than the original variable, although some skewness and clustering remain. Compared to the log transformation, the reciprocal square-root places greater weight on variation at the lower end of the gravity gradient, which may reduce leverage from high values but can also concentrate variability among low-gravity sites. Overall, this transformation is effective in mitigating extreme right-tail influence, but it should be noted that its interpretability is less straightforward than the log scale.

#### Transformed models: log

Model specification To evaluate whether transformation improves model performance, we fitted three alternative linear models: (i) log-transformed biomass regressed on untransformed gravity, (ii) untransformed biomass regressed on log-transformed gravity, and (iii) a fully log–log model with both variables log-transformed. The gravity predictor was mean-centred prior to modelling to ensure that intercepts represent expected biomass at average gravity and to maintain numerical stability:

```{r}

# Centre logged market gravity
og_model_data$log_og_centred <- as.numeric(
  scale(og_model_data$log_og, center = TRUE, scale = FALSE)
)

# Transformed models 
log_biomass_og <- lm(log_biomass~og_centred,data = og_model_data) 
biomass_log_og <- lm(mean_biomass~log_og_centred,data = og_model_data) 
log_biomass_log_og <- lm(log_biomass~log_og_centred,data = og_model_data) 

# Summaries of transformed models
summary(log_biomass_og) 
summary(biomass_log_og) 
summary(log_biomass_log_og) 

# or arm::display(lm_biomass_chla)
```

Across all transformed specifications, there remains no statistically supported relationship between human gravity and biomass. In the model with log-transformed biomass and untransformed gravity, the slope is negative but non-significant (β = −7.69 × 10⁻⁵, p = 0.268), explaining approximately 2% of the variance (R² = 0.020). When biomass is modelled against log-transformed gravity, explanatory power further declines (R² = 0.003), and the slope remains non-significant (p = 0.687). The fully log–log model similarly shows no evidence of association (β = −0.0196, p = 0.812), with negligible explanatory power (R² \< 0.001).

Thus, although transformations improve distributional properties, they do not materially strengthen the relationship between gravity and biomass. The consistently small effect sizes and very low R² values indicate that human gravity, as measured here, provides minimal explanatory value for variation in site-level biomass.

#### Transformed models: reciprocal sqrt

Model specification To assess whether the Box–Cox–suggested reciprocal square-root transformation (λ ≈ −0.5) improves model performance, we fitted two linear models using the transformed gravity predictor: (i) untransformed biomass regressed on reciprocal square-root gravity, and (ii) log-transformed biomass regressed on reciprocal square-root gravity. The transformed gravity variable was mean-centred prior to analysis so that intercepts represent expected biomass at average transformed gravity and to maintain numerical stability:

```{r}

# Centre reciprocal sqrt market gravity
og_model_data$og_recipr_sqrt_centred <- as.numeric(
  scale(og_model_data$og_recipr_sqrt, center = TRUE, scale = FALSE)
)

# Transformed models 
biomass_rsqrt_og <- lm(mean_biomass~og_recipr_sqrt_centred,data = og_model_data)
log_biomass_rsqrt_og <- lm(log_biomass~og_recipr_sqrt_centred,data = og_model_data) 

# Summaries of transformed models
summary(biomass_rsqrt_og) 
summary(log_biomass_rsqrt_og) 

```

The model with untransformed biomass shows no evidence of association between biomass and reciprocal square-root gravity (β = −2011, SE = 36874, p = 0.957). The explanatory power is effectively zero (R² ≈ 0.00005), indicating that the transformed predictor accounts for none of the variation in biomass.

Similarly, when biomass is log-transformed, the relationship remains non-significant (β = −0.164, SE = 0.707, p = 0.818), with negligible explanatory power (R² ≈ 0.0009).

Thus, although the reciprocal square-root transformation improves the distributional properties of the gravity variable, it does not strengthen the statistical relationship between gravity and biomass. Across specifications, effect sizes remain small, highly uncertain, and biologically negligible.

#### Residual checking of transformations

To determine whether transformation improved model adequacy, we examined the same set of diagnostic plots (residuals vs fitted, Q–Q plots, scale–location plots, and Cook’s distance) for each transformed specification:

```{r}

# log_biomass_old_market_grav
par(mfrow= c(2,2)) # divide the page into 2 x 2 frame
plot(biomass_log_og, which= c(1:4))

# biomass_log_old_market_grav
par(mfrow= c(2,2)) # divide the page into 2 x 2 frame
plot(log_biomass_og, which= c(1:4))

# log_biomass_log_old_market_grav
par(mfrow= c(2,2)) # divide the page into 2 x 2 frame
plot(log_biomass_log_og, which= c(1:4))

# biomass_rsqrt_old_market_grav
par(mfrow= c(2,2)) # divide the page into 2 x 2 frame
plot(biomass_rsqrt_og, which= c(1:4))

# log_biomass_rsqrt_old_market_grav
par(mfrow= c(2,2)) # divide the page into 2 x 2 frame
plot(log_biomass_rsqrt_og, which= c(1:4))
```

Results: In the model with log-transformed gravity only (mean_biomass \~ log_gravity), the residual structure remains broadly similar to the untransformed model. Heteroscedasticity persists, with increasing spread at higher fitted values, and the Q–Q plot still shows deviation in the upper tail. Influential observations remain present.

When biomass alone is log-transformed (log_biomass \~ gravity), residual variance is noticeably more stable across fitted values, and the Q–Q plot shows improved alignment with the normal reference line, although mild tail deviations remain. This indicates partial improvement in normality and homoscedasticity.

The fully log–log model (log_biomass \~ log_gravity) shows the most consistent variance structure, with residuals more evenly distributed and closer to normality, though slight upper-tail deviations and a few influential observations persist.

Models incorporating the reciprocal square-root transformation of gravity show a similar pattern: transforming gravity alone does not resolve heteroscedasticity or tail deviations, whereas models with log-transformed biomass exhibit improved variance stability and closer adherence to normality. However, the reciprocal square-root transformation provides no clear diagnostic advantage over the log transformation of gravity.

Overall, transforming the response variable (log biomass) substantially improves adherence to model assumptions, whereas transforming gravity alone—whether via log or reciprocal square-root—offers limited benefit.

### Check AICs

Because AIC values are only comparable among models fitted to the same response variable on the same scale, model selection was restricted to specifications with log-transformed biomass as the response. We therefore compared the models including untransformed gravity, log-transformed gravity, and reciprocal square-root gravity as predictors:

```{r}
og_base_models <- list(log_biomass_og, log_biomass_log_og, log_biomass_rsqrt_og)
aictab(og_base_models)
```

Model selection based on AICc indicates that the model with untransformed gravity predicting log biomass (Mod1) provides the best relative fit (AICc = 44.51). However, the log–log and reciprocal square-root models (Mod2 and Mod3) are within ΔAICc = 1.22 of the top model. Because models within ΔAICc \< 2 are generally considered to have substantial support, all three models are statistically competitive.

So, to conclude: When comparing functional forms of the gravity predictor (untransformed, log-transformed, and reciprocal square-root), there is no meaningful difference in inference. All three specifications yield very small, non-significant slope estimates, extremely low R² values, and nearly identical AICc scores. The ΔAICc values (\< 2) indicate that the models are statistically indistinguishable in terms of fit.

Residual diagnostics show that improvements in model assumptions are primarily driven by transforming the response variable (log biomass), not by transforming gravity. Changing the functional form of gravity—whether log or reciprocal square-root—does not materially improve variance structure, normality, or explanatory power relative to using gravity on its original scale.

Overall, there is no statistical or ecological justification for preferring a transformed gravity predictor over the untransformed version. The absence of a detectable effect of gravity on biomass is robust to the choice of transformation.

### Taking into account eco-region (country)

#### Fixed-effects model (separate)

We will first separate this into each different eco-region (country for now) We will use log for ease of interpretation

```{r}
# plot of biomass vs chlorophyll for a 'no pooling analysis'
( separate_lm_plot <- ggplot(og_model_data, aes(x = log_og_centred,
           y = log_biomass)) + geom_point() + geom_smooth(method = "lm", formula = y ~ x) +
  facet_grid(. ~ country) + theme_bw() + labs( x = " Logged Market Gravity",
    y = "Log-Biomass", title = "No pooling analysis") )

```

Differences across countries!

This analysis treats the relationship at each site as unique and independent of the others.

#### Combined fixed-effects model

At the other extreme, we could simply pool all the data together in a single simple regression analysis: (same as above).

```{r}
# Graph the data - X-Y scatter plot
og_combined_lm_plot <- og_model_data %>% 
  ggplot(aes(x=log_og_centred, y=log_biomass)) + 
  geom_point() +
  geom_smooth(method=lm)

og_combined_lm_plot
```

This model helps us to understand the analysis we want to build from these initial simple extremes.

#### Country as a fixed effect

Note: - from above, we will carry on with the log_log model - we centre chla. In models including interactions, centering also makes main effects interpretable as effects evaluated at the mean of the interacting variable and can improve numerical stability in mixed-effects models. - We used sum-to-zero contrasts (effect coding) for country so that parameter estimates represent deviations from the overall mean rather than differences relative to an arbitrarily chosen reference country. Under effect coding, the intercept corresponds to the grand mean of the response (averaged across all observations), and each country coefficient represents that country’s deviation from this grand mean. In other words, coefficients quantify how much higher or lower a given country’s mean is compared to the overall mean.

##### Additive model

First we add it in an additive simple least squares regression. This additive model assumes that the biomass–chlorophyll slope is the same across all countries, but allows countries to differ in their mean biomass levels:

```{r}
# Set sum-to-zero contrasts for country (the grand mean is the baseline - not an arbitrary country level)
contrasts(og_model_data$country) <- contr.sum(length(levels(og_model_data$country)))

# Addtitive model
og_country_additive <- lm(log_biomass ~ log_og_centred + country, data = og_model_data)
summary(og_country_additive) 
```

Output interpretation: ...

##### Interaction model

This interaction model allows the biomass–chlorophyll slope to differ among countries rather than forcing a common slope.

```{r}

# Log-log interaction
og_country_interaction <- lm(log_biomass ~ log_og_centred * country, data = og_model_data)
summary(og_country_interaction) 

```

...

#### Test global, additive and interaction models

```{r}
#AIC
interaction_test_models_og <- list(log_biomass_log_og, og_country_additive, og_country_interaction)
aictab(interaction_test_models_og) 

# Classical F-test
anova(log_biomass_log_og, og_country_additive, og_country_interaction)
```

It seems that not interaction, but additive mdoel does not add loads...

## Market gravity

```{r}

# Histogram market gravity 
( gravity_hist <- ggplot(gravity_metrics_2009, aes(x = market_gravity)) +
  geom_histogram(bins = 10) +
  labs(
    x = expression("Market  gravity"),
    y = "Number of sites"
  ) +
  theme_bw() )

# Boxcox 
boxcox(lm(market_gravity ~ 1, data = gravity_metrics_2009),
       lambda = seq(-2, 2, 0.1))

```

Outputs: Market gravity is extremely right skewed. The vast majority of sites have very low (or zero) values of market gravity, with only a small number of sites exhibiting very large values (in the millions). This indicates a heavy-tailed distribution with strong imbalance. Since it is not normally distributed, it will violate Gaussian assumptions if used on the raw scale. Any linear model on the untransformed scale would be dominated by the few extreme values.

Additionally, since the dynamic range spans several orders of magnitude. That strongly suggests that effects, if present, are likely multiplicative rather than additive.

The Box–Cox analysis confirms that market gravity should be log-transformed before modelling. Note that there is no need to use the exact maximum-likelihood λ; log is statistically justified and substantively coherent for a gravity-type variable spanning orders of magnitude.

### Exploratory Plot

```{r}
#| label: biomass-vs-gravity-plot

# Graph the data - X-Y scatter plot
biomass_mg_fig1 <- merge(x = biomass_per_site, y = gravity_metrics_2009, by = "site", all = TRUE) %>%
  ggplot(aes(x=market_gravity, y=mean_biomass)) + 
  geom_point() +
  geom_smooth(method=lm)

biomass_mg_fig1

# Correlation
with(merge(x = biomass_per_site, y = gravity_metrics_2009, by = "site", all = TRUE),
    cor(market_gravity, mean_biomass))

```

Human gravity seems to have a negative effect on biomass, but the are large confidence intervals, and a lot of uncertainty in the data.

### Untransformed ls model

model 1 - the simplest untransforemd regression

```{r}
#| label: biomass-vs-chla-lm

# Merge needed data
mg_model_data <- merge(biomass_per_site, gravity_metrics_2009[, c("site", "market_gravity")], by = "site", all = TRUE)


# Centre market gravity
mg_model_data$mg_centred <- as.numeric(scale(
  mg_model_data$market_gravity,
  center = TRUE,
  scale = FALSE ))

# Build lm 
base_model_mg <- lm(mean_biomass~mg_centred,data = mg_model_data) 
```

Analyse model:

```{r}
# Analyse the model 
coef(base_model_mg)
summary(base_model_mg) 
confint(base_model_mg) 
```

...

#### Residual checking of simple ls model

But model assumptions and violations:

```{r}
par(mfrow= c(2,2)) # divide the page into 2 x 2 frame
# par(mar=c(1,1,1,1))
plot( base_model_mg, which= c(1:4))
```

...

### Transformations

To improve model fit, we can transform the raw data to better conform to the regression assumptions, thereby amplifying predictive power and increasing the overall quality of the model.

We also check how the distribution of the data changes after each transformation to ensure that everything is working as it should.

```{r}

# Transforming variables
mg_model_data$log_biomass <- log10(mg_model_data$mean_biomass)
mg_model_data$log_market_grav <- log10(mg_model_data$market_gravity)

# Check transformations 

( log_market_grav_hist <- ggplot(mg_model_data, aes(x = log_market_grav)) +
  geom_histogram(bins = 20) +
  labs(
    x = expression("Log-Transformed Market Gravity"),
    y = "Frequency"
  ) +
  theme_bw() )
```

The log-transformed distribution is still somewhat skewed, but it is dramatically compressed relative to the raw scale. The extreme values are no longer dominating the range. The distribution also looks mildly bimodal or clustered rather than smoothly normal. That is not necessarily a transformation problem. It may reflect real structural differences (e.g., remote vs highly connected sites, or country-level clustering). Transformations address skewness, not underlying multimodality caused by real grouping.

Transformed models: Note that we will be centering the gravity data. In most ecological regression contexts, especially with interaction terms, centering is preferable because it avoids interpreting coefficients at unrealistic covariate values, It reduces collinearity with interaction terms, it improves numerical stability.

```{r}

# Centre logged market gravity
mg_model_data$log_mg_centred <- as.numeric(
  scale(mg_model_data$log_market_grav, center = TRUE, scale = FALSE)
)

# Transformed models 
log_biomass_mg <- lm(log_biomass~mg_centred,data = mg_model_data) 
biomass_log_mg<- lm(mean_biomass~log_mg_centred,data = mg_model_data) 
log_biomass_log_mg <- lm(log_biomass~log_mg_centred,data = mg_model_data) 

# Summaries of transformed models
summary(log_biomass_mg) 
summary(biomass_log_mg) 
summary(log_biomass_log_mg) 

# or arm::display(lm_biomass_chla)
#
```

...

### Residual checking of transformations

```{r}

# log_biomass_chla
par(mfrow= c(2,2)) # divide the page into 2 x 2 frame
plot(biomass_log_mg, which= c(1:4))

# biomass_log_chla
par(mfrow= c(2,2)) # divide the page into 2 x 2 frame
plot(log_biomass_mg, which= c(1:4))

# log_biomass_log_chla
par(mfrow= c(2,2)) # divide the page into 2 x 2 frame
plot(log_biomass_log_mg, which= c(1:4))
```

the log–log form better satisfies linear model assumptions ...

### Check AICs

Note that you cannot compare AIC across different response scales, so, given that log-biomass better fits the residuals, we are interested to know which model - log_biomass_chla or log-log - provides the best fit

```{r}
grav_base_models <- list(log_biomass_mg, log_biomass_log_mg)
aictab(grav_base_models)
```

...

### Comparing gravities

```{r}
grav_metric_models <- list(log_biomass_log_og, log_biomass_log_mg)
aictab(grav_metric_models)

anova(log_biomass_log_og,log_biomass_log_mg)
```

Perhaps as would be expected they are very very similar in fit...

### Taking into account eco-region (country)

#### Fixed-effects model (separate)

We will first separate this into each different eco-region (country for now)

```{r}
# plot of biomass vs chlorophyll for a 'no pooling analysis'
( mg_separate_plot <- ggplot(mg_model_data, aes(x = log_mg_centred,
           y = log_biomass)) + geom_point() + geom_smooth(method = "lm", formula = y ~ x) +
  facet_grid(. ~ country) + theme_bw() + labs( x = " Logged Old Market Gravity",
    y = "Log-Biomass", title = "No pooling analysis") )

```

Great differences across countries!!

This analysis treats the relationship at each site as unique and independent of the others.

#### Combined fixed-effects model

At the other extreme, we could simply pool all the data together in a single simple regression analysis: (same as above).

```{r}
# Graph the data - X-Y scatter plot
mg_combined_plot <- mg_model_data %>% 
  ggplot(aes(x=log_mg_centred, y=log_biomass)) + 
  geom_point() +
  geom_smooth(method=lm)

mg_combined_plot
```

This model helps us to understand the analysis we want to build from these initial simple extremes.

#### Country as a fixed effect

Note: - from above, we will carry on with the log_log model - we centre chla. In models including interactions, centering also makes main effects interpretable as effects evaluated at the mean of the interacting variable and can improve numerical stability in mixed-effects models. - We used sum-to-zero contrasts (effect coding) for country so that parameter estimates represent deviations from the overall mean rather than differences relative to an arbitrarily chosen reference country. Under effect coding, the intercept corresponds to the grand mean of the response (averaged across all observations), and each country coefficient represents that country’s deviation from this grand mean. In other words, coefficients quantify how much higher or lower a given country’s mean is compared to the overall mean.

##### Additive model

First we add it in an additive simple least squares regression. This additive model assumes that the biomass–chlorophyll slope is the same across all countries, but allows countries to differ in their mean biomass levels:

```{r}
# Set sum-to-zero contrasts for country (the grand mean is the baseline - not an arbitrary country level)
contrasts(mg_model_data$country) <- contr.sum(length(levels(mg_model_data$country)))

# Addtitive model
mg_country_additive <- lm(log_biomass ~ log_mg_centred + country, data = mg_model_data)
summary(mg_country_additive) 
```

Output interpretation: ...

##### Interaction model

This interaction model allows the biomass–chlorophyll slope to differ among countries rather than forcing a common slope.

```{r}

# Log-log interaction
mg_country_interaction <- lm(log_biomass ~ log_mg_centred * country, data = mg_model_data)
summary(mg_country_interaction) 

```

...

##### Test global, additive and interaction models

```{r}
#AIC
interaction_test_models_mg <- list(log_biomass_log_mg, mg_country_additive, mg_country_interaction)
aictab(interaction_test_models_mg) 

# Classical F-test
anova(log_biomass_log_mg, mg_country_additive, mg_country_interaction)
```

Note that again there isn't a huge different between global and additive model, but given that we want to keep the ecoregion effect, we will keep it for now :)

## Settlement gravity (settlement top pop)

```{r}

# Histogram market gravity 
( st_hist <- ggplot(gravity_metrics_2009, aes(x = settlement_tot_pop)) +
  geom_histogram(bins = 10) +
  labs(
    x = expression("Settlement tot pop"),
    y = "Number of sites"
  ) +
  theme_bw() )

# Boxcox 
boxcox(lm(settlement_tot_pop ~ 1, data = gravity_metrics_2009),
       lambda = seq(-2, 2, 0.1))

```

Outputs: Settlement gravity is extremely right skewed. The Box–Cox analysis confirms that settlement gravity should be log-transformed before modelling. Note that there is no need to use the exact maximum-likelihood λ; log is statistically justified and substantively coherent for a gravity-type variable spanning orders of magnitude.

### Exploratory Plot

```{r}
#| label: biomass-vs-st-plot

# Graph the data - X-Y scatter plot
biomass_st_fig1 <- merge(x = biomass_per_site, y = gravity_metrics_2009, by = "site", all = TRUE) %>%
  ggplot(aes(x=settlement_tot_pop, y=mean_biomass)) + 
  geom_point() +
  geom_smooth(method=lm)

biomass_st_fig1

# Correlation
with(merge(x = biomass_per_site, y = gravity_metrics_2009, by = "site", all = TRUE),
    cor(settlement_tot_pop, mean_biomass))

```

Human gravity seems to have a negative effect on biomass, but the are large confidence intervals, and a lot of uncertainty in the data.

### Untransformed ls model

model 1 - the simplest untransformed regression

```{r}
#| label: biomass-vs-chla-lm

# Merge needed data
st_model_data <- merge(biomass_per_site, gravity_metrics_2009[, c("site", "settlement_tot_pop")], by = "site", all = TRUE)

# Centre market gravity
st_model_data$st_centred <- as.numeric(scale(
  st_model_data$settlement_tot_pop,
  center = TRUE,
  scale = FALSE ))

# Build lm 
base_model_st <- lm(mean_biomass~st_centred,data = st_model_data) 
```

Analyse model:

```{r}
# Analyse the model 
coef(base_model_st)
summary(base_model_st) 
confint(base_model_st) 
```

...

#### Residual checking of simple ls model

But model assumptions and violations:

```{r}
par(mfrow= c(2,2)) # divide the page into 2 x 2 frame
# par(mar=c(1,1,1,1))
plot( base_model_st, which= c(1:4))
```

...

### Transformations

To improve model fit, we can transform the raw data to better conform to the regression assumptions, thereby amplifying predictive power and increasing the overall quality of the model.

We also check how the distribution of the data changes after each transformation to ensure that everything is working as it should.

```{r}

# Transforming variables
st_model_data$log_biomass <- log10(st_model_data$mean_biomass)
st_model_data$log_settlement_tot <- log10(st_model_data$settlement_tot_pop)

# Check transformations 

( log_st_hist <- ggplot(st_model_data, aes(x = log_settlement_tot)) +
  geom_histogram(bins = 20) +
  labs(
    x = expression("Log-Transformed Settlement Tot Pop"),
    y = "Frequency"
  ) +
  theme_bw() )
```

The distribution also looks mildly bimodal or clustered rather than smoothly normal. That is not necessarily a transformation problem. It may reflect real structural differences (e.g., remote vs highly connected sites, or country-level clustering). Transformations address skewness, not underlying multimodality caused by real grouping.

Transformed models: Note that we will be centering the gravity data. In most ecological regression contexts, especially with interaction terms, centering is preferable because it avoids interpreting coefficients at unrealistic covariate values, It reduces collinearity with interaction terms, it improves numerical stability.

```{r}

# Centre logged market gravity
st_model_data$log_st_centred <- as.numeric(
  scale(st_model_data$log_settlement_tot, center = TRUE, scale = FALSE)
)

# Transformed models 
log_biomass_st <- lm(log_biomass~st_centred,data = st_model_data) 
biomass_log_st<- lm(mean_biomass~log_st_centred,data = st_model_data) 
log_biomass_log_st <- lm(log_biomass~log_st_centred,data = st_model_data) 

# Summaries of transformed models
summary(log_biomass_st) 
summary(biomass_log_st) 
summary(log_biomass_log_st) 

```

...

### Residual checking of transformations

```{r}

# log_biomass_chla
par(mfrow= c(2,2)) # divide the page into 2 x 2 frame
plot(biomass_log_st, which= c(1:4))

# biomass_log_chla
par(mfrow= c(2,2)) # divide the page into 2 x 2 frame
plot(log_biomass_st, which= c(1:4))

# log_biomass_log_chla
par(mfrow= c(2,2)) # divide the page into 2 x 2 frame
plot(log_biomass_log_st, which= c(1:4))
```

the log–log form better satisfies linear model assumptions ...

### Check AICs

Note that you cannot compare AIC across different response scales, so, given that log-biomass better fits the residuals, we are interested to know which model - log_biomass_chla or log-log - provides the best fit

```{r}
st_base_models <- list(log_biomass_st, log_biomass_log_st)
aictab(st_base_models)
```

...

### Comparing gravities

```{r}
grav_metric_models2 <- list(log_biomass_log_mg, log_biomass_log_st)
aictab(grav_metric_models2)

anova(log_biomass_log_mg,log_biomass_log_st)
```

...

## Settlement gravity (nearest pop grav)

```{r}

#Change format of settlement pop gravity
gravity_metrics_2009$nearest_pop_grav <-  format(gravity_metrics_2009$nearest_pop_grav, scientific = FALSE)
gravity_metrics_2009$nearest_pop_grav <-  as.numeric(gravity_metrics_2009$nearest_pop_grav)

# Histogram market gravity 
( nearest_pop_hist <- ggplot(gravity_metrics_2009, aes(x = nearest_pop_grav)) +
  geom_histogram(bins = 20) +
  labs(
    x = expression("Nearest Pop Grav"),
    y = "Number of sites"
  ) +
  theme_bw() )

# Boxcox 
boxcox(lm(nearest_pop_grav ~ 1, data = gravity_metrics_2009),
       lambda = seq(-2, 2, 0.1))

```

Outputs: Settlement gravity is extremely right skewed. The Box–Cox analysis confirms that settlement gravity should be log-transformed before modelling. Note that there is no need to use the exact maximum-likelihood λ; log is statistically justified and substantively coherent for a gravity-type variable spanning orders of magnitude.

### Exploratory Plot

```{r}
#| label: biomass-vs-gravity-plot

# Graph the data - X-Y scatter plot
biomass_sg_fig1 <- merge(x = biomass_per_site, y = gravity_metrics_2009, by = "site", all = TRUE) %>%
  ggplot(aes(x=nearest_pop_grav, y=mean_biomass)) + 
  geom_point() +
  geom_smooth(method=lm)

biomass_sg_fig1

# Correlation
with(merge(x = biomass_per_site, y = gravity_metrics_2009, by = "site", all = TRUE),
    cor(nearest_pop_grav, mean_biomass))

```

Human gravity seems to have a negative effect on biomass, but the are large confidence intervals, and a lot of uncertainty in the data.

### Untransformed ls model

model 1 - the simplest untransforemd regression

```{r}
#| label: biomass-vs-chla-lm

# Merge needed data
sg_model_data <- merge(biomass_per_site, gravity_metrics_2009[, c("site", "nearest_pop_grav")], by = "site", all = TRUE)

# Centre market gravity
sg_model_data$sg_centred <- as.numeric(scale(
  sg_model_data$nearest_pop_grav,
  center = TRUE,
  scale = FALSE ))

# Build lm 
base_model_sg <- lm(mean_biomass~sg_centred,data = sg_model_data) 
```

Analyse model:

```{r}
# Analyse the model 
coef(base_model_sg)
summary(base_model_sg) 
confint(base_model_sg) 
```

...

#### Residual checking of simple ls model

But model assumptions and violations:

```{r}
par(mfrow= c(2,2)) # divide the page into 2 x 2 frame
# par(mar=c(1,1,1,1))
plot( base_model_sg, which= c(1:4))
```

...

### Transformations

To improve model fit, we can transform the raw data to better conform to the regression assumptions, thereby amplifying predictive power and increasing the overall quality of the model.

We also check how the distribution of the data changes after each transformation to ensure that everything is working as it should.

```{r}

# Transforming variables
sg_model_data$log_biomass <- log10(sg_model_data$mean_biomass)
sg_model_data$log_sg <- log10(sg_model_data$nearest_pop_grav)

# Check transformations 

( log_sg_hist <- ggplot(sg_model_data, aes(x = log_sg)) +
  geom_histogram(bins = 20) +
  labs(
    x = expression("Log-Transformed Settlement Tot Pop"),
    y = "Frequency"
  ) +
  theme_bw() )
```

The distribution also looks mildly bimodal or clustered rather than smoothly normal. That is not necessarily a transformation problem. It may reflect real structural differences (e.g., remote vs highly connected sites, or country-level clustering). Transformations address skewness, not underlying multimodality caused by real grouping.

Transformed models: Note that we will be centering the gravity data. In most ecological regression contexts, especially with interaction terms, centering is preferable because it avoids interpreting coefficients at unrealistic covariate values, It reduces collinearity with interaction terms, it improves numerical stability.

```{r}

# Centre logged market gravity
sg_model_data$log_sg_centred <- as.numeric(
  scale(sg_model_data$log_sg, center = TRUE, scale = FALSE)
)

# Transformed models 
log_biomass_sg <- lm(log_biomass~sg_centred,data = sg_model_data) 
biomass_log_sg <- lm(mean_biomass~log_sg_centred,data = sg_model_data) 
log_biomass_log_sg <- lm(log_biomass~log_sg_centred,data = sg_model_data) 

# Summaries of transformed models
summary(log_biomass_sg) 
summary(biomass_log_sg) 
summary(log_biomass_log_sg) 

```

...

### Residual checking of transformations

```{r}

# log_biomass_chla
par(mfrow= c(2,2)) # divide the page into 2 x 2 frame
plot(biomass_log_sg, which= c(1:4))

# biomass_log_chla
par(mfrow= c(2,2)) # divide the page into 2 x 2 frame
plot(log_biomass_sg, which= c(1:4))

# log_biomass_log_chla
par(mfrow= c(2,2)) # divide the page into 2 x 2 frame
plot(log_biomass_log_sg, which= c(1:4))
```

the log–log form better satisfies linear model assumptions ...

### Check AICs

Note that you cannot compare AIC across different response scales, so, given that log-biomass better fits the residuals, we are interested to know which model - log_biomass_chla or log-log - provides the best fit

```{r}
sg_base_models <- list(log_biomass_sg, log_biomass_log_sg)
aictab(sg_base_models)
```

Similar but...

### Comparing gravities

```{r}
grav_metric_models_all <- list(log_biomass_log_og, log_biomass_log_mg, log_biomass_log_st, log_biomass_log_sg)
aictab(grav_metric_models_all)

anova(log_biomass_log_og, log_biomass_log_mg, log_biomass_log_st, log_biomass_log_sg)
```

Settlement gravity seems to fit better!! So we keep this going forward

### Taking into account eco-region (country)

#### Fixed-effects model (separate)

We will first separate this into each different eco-region (country for now)

```{r}
# plot of biomass vs chlorophyll for a 'no pooling analysis'
( sg_separate_plot <- ggplot(sg_model_data, aes(x = log_sg_centred,
           y = log_biomass)) + geom_point() + geom_smooth(method = "lm", formula = y ~ x) +
  facet_grid(. ~ country) + theme_bw() + labs( x = " Logged Settlement Gravity",
    y = "Log-Biomass", title = "No pooling analysis") )

```

Great differences across countries!!

This analysis treats the relationship at each site as unique and independent of the others.

#### Combined fixed-effects model

At the other extreme, we could simply pool all the data together in a single simple regression analysis: (same as above).

```{r}
# Graph the data - X-Y scatter plot
sg_combined_plot <- sg_model_data %>% 
  ggplot(aes(x=log_sg_centred, y=log_biomass)) + 
  geom_point() +
  geom_smooth(method=lm)

sg_combined_plot
```

This model helps us to understand the analysis we want to build from these initial simple extremes.

#### Country as a fixed effect

Note: - from above, we will carry on with the log_log model - we centre chla. In models including interactions, centering also makes main effects interpretable as effects evaluated at the mean of the interacting variable and can improve numerical stability in mixed-effects models. - We used sum-to-zero contrasts (effect coding) for country so that parameter estimates represent deviations from the overall mean rather than differences relative to an arbitrarily chosen reference country. Under effect coding, the intercept corresponds to the grand mean of the response (averaged across all observations), and each country coefficient represents that country’s deviation from this grand mean. In other words, coefficients quantify how much higher or lower a given country’s mean is compared to the overall mean.

##### Additive model

First we add it in an additive simple least squares regression. This additive model assumes that the biomass–chlorophyll slope is the same across all countries, but allows countries to differ in their mean biomass levels:

```{r}
# Set sum-to-zero contrasts for country (the grand mean is the baseline - not an arbitrary country level)
contrasts(sg_model_data$country) <- contr.sum(length(levels(sg_model_data$country)))

# Addtitive model
sg_country_additive <- lm(log_biomass ~ log_sg_centred + country, data = sg_model_data)
summary(sg_country_additive) 
```

Output interpretation: ...

##### Interaction model

This interaction model allows the biomass–chlorophyll slope to differ among countries rather than forcing a common slope.

```{r}

# Log-log interaction
sg_country_interaction <- lm(log_biomass ~ log_sg_centred * country, data = sg_model_data)
summary(sg_country_interaction) 

```

...

##### Test global, additive and interaction models

```{r}
#AIC
interaction_test_models_sg <- list(log_biomass_log_sg, sg_country_additive, sg_country_interaction)
aictab(interaction_test_models_sg) 

# Classical F-test
anova(log_biomass_log_sg, sg_country_additive, sg_country_interaction)
```

Note that again the no-country model works a lot better!! Non-additive looks a lot better!

May be better to work with connectivity?

# GLM models

We fitted a generalized linear model (GLM) rather than applying a transformation to the response variable because GLMs allow the mean–variance relationship of the data to be modelled explicitly through an appropriate error distribution and link function. In contrast, transforming the response (e.g., log-transforming biomass) changes the scale of the outcome and can complicate interpretation, while still relying on Gaussian error assumptions.

First of all we will produce a comparative GLM to ensure that we can test the transformation again.

This is a base model but we will keep the log transformed data as mentioned before because it fits the assumptions better

```{r}
#Fitting a GLM
glm_gaussian_sg <- glm(
  mean_biomass ~ log_sg_centred,
  family = gaussian(link = "identity"),
  data = sg_model_data)

summary(glm_gaussian_sg)
confint(glm_gaussian_sg)
```

Next, a Gamma error distribution with a log link was used because site-level biomass is strictly positive and strongly right-skewed, with variance increasing as the mean increases. The Gamma distribution is well suited to modelling continuous, positive data with this mean–variance structure, and the log link allows effects of settlement gravity to be interpreted multiplicatively (i.e., proportional changes in biomass). Note that we are still using the centred version of settlement gravity. Centering does not change inference about the effect of gravity. It only improves interpretability of the intercept and helps if you later add interactions (e.g., country × gravity). We will aslo explicitly test the tranform x with the non-transformed again.

```{r}

#Fitting a GLM
glm_gamma_sg <- glm(
  mean_biomass ~ log_sg_centred,
  family = Gamma(link = "log"),
  data = sg_model_data)

summary(glm_gamma_sg)
confint(glm_gamma_sg)

#Fitting a GLM
glm_gamma_sg2 <- glm(
  mean_biomass ~ sg_centred,
  family = Gamma(link = "log"),
  data = sg_model_data)

summary(glm_gamma_sg2)
confint(glm_gamma_sg2)
```

Interpretation: Settlement gravity was significantly negatively associated with site-level fish biomass under a generalized linear model with a Gamma error distribution and log link. The coefficient for log-transformed and mean-centred settlement gravity was −0.174 (SE = 0.058, t = −3.01, p = 0.0038), indicating a statistically significant decline in biomass with increasing human gravity. The 95% confidence interval for the slope (−0.285 to −0.049) did not overlap zero, confirming the robustness of this negative relationship.

Because a log link was used, coefficients are interpreted multiplicatively on the original biomass scale. Exponentiating the slope indicates that a one-unit increase in log settlement gravity is associated with an approximately 16% reduction in expected biomass (95% CI: 5–25%). Thus, increases in human population gravity are associated with proportionally lower reef fish biomass across sites.

Model fit statistics indicated moderate explanatory power. The model reduced deviance from 34.90 (null model) to 30.88, corresponding to approximately 12% of deviance explained by settlement gravity alone. The estimated dispersion parameter (0.56) and simulation-based residual diagnostics indicated no evidence of overdispersion or distributional misspecification, supporting the adequacy of the Gamma GLM framework.

Note that both the x transformed and the x raw are significant. Here we are jsut trying to tune up the model.

Now we check the model fit more closely:

```{r}
library(DHARMa)

gamma_sg_res <- simulateResiduals(glm_gamma_sg)
plot(gamma_sg_res)
testDispersion(gamma_sg_res)
```

Diagnostic checks using DHARMa indicated that the Gamma generalized linear model with a log link provided an adequate fit to the biomass data. The simulated residuals did not deviate significantly from the expected uniform distribution (KS test: p = 0.496), suggesting that the assumed error distribution is appropriate. The dispersion test was non-significant (dispersion = 1.32, p = 0.272), indicating no evidence of overdispersion or underdispersion. Residuals plotted against fitted values showed no systematic patterns, and no significant outliers were detected (p = 0.391).

Overall, there was no evidence of model misspecification, supporting the use of the Gamma GLM for subsequent inference.

Taken together,these results provide statistically robust evidence that higher settlement gravity is associated with reduced reef fish biomass at the site level.

## Check Gaussian vs Gamma -log model

```{r}
glm_test_models <- list(glm_gaussian_sg, glm_gamma_sg, glm_gamma_sg2)
aictab(glm_test_models)
```

Model comparison using AICc strongly supported the Gamma GLM over the Gaussian alternative. The Gamma model had the lowest AICc (1345.76), whereas the Gaussian model had a substantially higher AICc (1388.41), resulting in a ΔAICc of 42.65. Differences in AICc greater than 10 are generally considered decisive evidence against the higher-AIC model; therefore, the Gaussian model receives effectively no support relative to the Gamma specification. The AICc weight for the Gamma model was 1.00, indicating overwhelming support among the candidate models.

These results confirm that modelling biomass with a Gamma error distribution and log link provides a substantially better fit to the data than assuming Gaussian errors. This further justifies the use of the Gamma GLM for inference.

And also using log-transformed x didnt seem to make a huge difference so maybe best to keep untransfromed?

## Check gravity metrics again

```{r}
# Old gravity glm
glm_gamma_og <- glm(
  mean_biomass ~ log_og_centred,
  family = Gamma(link = "log"),
  data = og_model_data)

summary(glm_gamma_og)
confint(glm_gamma_og)

# Market gravity glm
glm_gamma_mg <- glm(
  mean_biomass ~ log_mg_centred,
  family = Gamma(link = "log"),
  data = mg_model_data)

summary(glm_gamma_mg)
confint(glm_gamma_mg)

# Settlement tot pop glm
glm_gamma_st <- glm(
  mean_biomass ~ log_st_centred,
  family = Gamma(link = "log"),
  data = st_model_data)

summary(glm_gamma_st)
confint(glm_gamma_st)
```

```{r}
glm_gravity_models <- list(glm_gamma_sg, glm_gamma_og, glm_gamma_mg, glm_gamma_st)
aictab(glm_gravity_models)
```

Model comparison using AICc indicated that settlement gravity (Mod1) provided the best fit among the candidate gravity metrics. The settlement gravity model had the lowest AICc (1345.76) and the highest model weight (AICcWt = 0.86), indicating strong support relative to the alternatives.

Substantively, this means:

Settlement gravity is not just statistically significant.

It also provides the strongest explanatory power among alternative human-gravity measures.

The evidence for preferring settlement gravity over the others is strong but not absolutely overwhelming (since ΔAICc ≈ 4 for the second-best model).

## Check country as an ecoregion

```{r}

contrasts(sg_model_data$country) <- contr.sum(length(levels(sg_model_data$country)))

glm_country_additive <- glm(
  mean_biomass ~ log_sg_centred + country,
  data = sg_model_data,
  family = Gamma(link = "log"))

summary(glm_country_additive)

glm_country_interaction <- glm(
  mean_biomass ~ log_sg_centred*country,
  data = sg_model_data,
  family = Gamma(link = "log"))

summary(glm_country_interaction)

```

```{r}
glm_country_models <- list(glm_gamma_sg, glm_country_additive, glm_country_interaction)
aictab(glm_country_models)
```

Model comparison indicated that the gravity-only model provided the strongest support (AICc = 1345.76, AICcWt = 0.81). Adding country as an additive effect resulted in a modest increase in AICc (ΔAICc = 2.89), indicating limited additional explanatory value. Allowing the effect of gravity to vary among countries (interaction model) was strongly unsupported (ΔAICc = 12.57). These results suggest that the negative relationship between settlement gravity and biomass is broadly consistent across countries and does not vary substantially among eco-regions.

# Total Biomass (Laura) vs Human gravity

## Old market gravity

Firstly we explore the relationship between biomass and the market gravity extracted from \[site\] While this was calculated for \[year\], it has been used in other papers \[such as samoilys, 2025\] and therefore provides a good reference for comparison.

### Explore the structure of the data

We explore the distribution of the gravity variable to assess its suitability for parametric statistical modelling. Visualising the histogram allows us to evaluate distributional properties such as skewness, dispersion, and the presence of outliers, which inform assumptions of normality and homoscedasticity. The Box–Cox analysis formally evaluates whether a power transformation would improve normality and stabilize variance, thereby guiding appropriate transformation of the predictor prior to inclusion in subsequent models.

```{r}

# Histogram market gravity 
( gravity_hist_laura <- ggplot(gravity_metrics_2009_laura, aes(x = old_gravity)) +
  geom_histogram(bins = 10) +
  labs(
    x = expression(" Old market gravity"),
    y = "Number of sites"
  ) +
  theme_bw() )

# Boxcox 
boxcox(lm(old_gravity ~ 1, data = gravity_metrics_2009_laura),
       lambda = seq(-2, 2, 0.1))

```

The histogram shows that the market gravity metric is highly right-skewed, with most sites clustered at low values and a small number of extreme high values (strong positive skew and potential outliers). This indicates a clear deviation from normality and suggests that the raw variable is unlikely to meet linear modelling assumptions.

The Box–Cox analysis indicates an optimal negative λ (≈ −0.5), confirming that a decreasing power transformation is required and that the untransformed variable is inappropriate for parametric modelling. Because λ values near the optimum yield similar likelihoods and log transformations are more interpretable, it is reasonable to compare a log transformation with a reciprocal-type transformation and select the most appropriate form based on subsequent model diagnostics.

### Exploratory Plot

Although prior diagnostics indicated that old_gravity is strongly right-skewed and likely requires transformation, we first fit a simple linear regression using the untransformed predictor as a baseline model. This provides a reference against which transformed models can be formally compared, allowing us to evaluate whether transformation improves model fit, residual behaviour, and inferential robustness.

```{r}
#| label: biomass-vs-gravity-plot

# Graph the data - X-Y scatter plot
biomass_gravity_fig2 <- merge(x = biomass_per_site_laura, y = gravity_metrics_2009_laura, by = "site", all = TRUE) %>%
  ggplot(aes(x=old_gravity, y=mean_biomass)) + 
  geom_point() +
  geom_smooth(method=lm)

biomass_gravity_fig2

# Correlation
with(merge(x = biomass_per_site_laura, y = gravity_metrics_2009_laura, by = "site", all = TRUE),
    cor(old_gravity, mean_biomass))

```

The scatterplot and regression line suggest a weak negative relationship between human gravity and mean biomass, with biomass tending to decline slightly as gravity increases. However, variability is high, confidence intervals are wide—particularly at higher gravity values—and the correlation is small (r = −0.136), indicating that gravity explains little of the observed variation in biomass in the untransformed linear model.

### Untransformed ls model

We fitted a simple linear regression model with mean biomass as the response variable and centred old market gravity as the sole predictor. This represents the most parsimonious, untransformed baseline model against which subsequent transformed models can be compared.

Note that the gravity predictor was mean-centred by subtracting its sample mean from each observation, so that a value of zero corresponds to the average gravity across sites. This reparameterisation does not change the estimated slope or its statistical significance, but it shifts the intercept to represent the expected biomass at average gravity rather than at gravity = 0, thereby improving the interpretability of the results.

```{r}
#| label: biomass-vs-chla-lm

# Merge needed data
og_model_data2 <- merge(biomass_per_site_laura, gravity_metrics_2009_laura[, c("site", "old_gravity")], by = "site", all = TRUE)

# Centre old market gravity
og_model_data2$og_centred <- as.numeric(scale(
  og_model_data2$old_gravity,
  center = TRUE,
  scale = FALSE ))

# Build lm 
base_model_og2 <- lm(mean_biomass~og_centred,data = og_model_data2) 
```

Analyse model:

```{r}
# Analyse the model 
coef(base_model_og2)
summary(base_model_og2) 
confint(base_model_og2) 
```

The model indicates a negative but non-significant association between gravity and biomass (β = −3.81 ± 3.59 SE, p = 0.292), with the 95% confidence interval spanning zero. The model explains very little variation in biomass (R² = 0.018), and the overall regression is not statistically significant (F₁,₆₀ = 1.13, p = 0.292). Thus, in its untransformed form, human gravity does not provide meaningful explanatory power for variation in biomass.

#### Checking assumptions

```{r}

# Extracting fitted values and residuals 
og_model_data$predicted <- fitted.values(base_model_og)
og_model_data$residuals <- residuals(base_model_og) 

# Histogram of residuals 
( og_residual_hist <- ggplot(og_model_data, aes(x = residuals)) +
  # Change y to density so the curve matches the scale
  geom_histogram(aes(y = after_stat(density)), bins = 10, fill = "lightgray", color = "white") +
  geom_density(color = "black", size = 1) +
  labs(
    x = expression("Residuals"),
    y = "Number of sites"
  ) +
  theme_bw() )
```

Residual checking of simple ls model To evaluate whether the assumptions of the linear model are satisfied, we examined standard diagnostic plots: residuals versus fitted values, a normal Q–Q plot, a scale–location plot, and Cook’s distance. These diagnostics assess linearity, homoscedasticity, normality of residuals, and the presence of influential observations.

But model assumptions and violations:

```{r}
par(mfrow= c(2,2)) # divide the page into 2 x 2 frame
# par(mar=c(1,1,1,1))
plot( base_model_og, which= c(1:4))
```

The residuals versus fitted plot indicates increasing dispersion at higher fitted values, suggesting heteroscedasticity rather than constant error variance. This pattern is reinforced by the scale–location plot, where residual spread clearly increases with fitted biomass. The Q–Q plot further reveals departures from normality, particularly in the upper tail, consistent with right-skewed residuals and the presence of extreme values. Cook’s distance highlights several moderately influential observations, likely associated with sites exhibiting high gravity values and leverage.

Taken together, these patterns indicate that key assumptions of normality and homoscedasticity are not satisfied under the untransformed specification, supporting the need for transformation before proceeding with formal inference.

### Transformations

Transformations were applied to improve adherence to linear regression assumptions, particularly normality and homoscedasticity, and to reduce the influence of extreme values. After each transformation, the distribution of the variable was re-examined to verify that skewness was reduced and that the transformation behaved as expected. These transformations amplifying predictive power and increasing the overall quality of the model

```{r}

# Transforming variables
og_model_data2$log_biomass <- log10(og_model_data2$mean_biomass)
og_model_data2$log_og <- log10(og_model_data2$old_gravity)

# Check transformations 

( log_market_grav_hist2 <- ggplot(og_model_data2, aes(x = log_og)) +
  geom_histogram(bins = 20) +
  labs(
    x = expression("Log-Transformed Old Gravity"),
    y = "Frequency"
  ) +
  theme_bw() )

# Reciprocal square-root transformation (λ ≈ -0.5)
og_model_data2$og_recipr_sqrt <- 1 / sqrt(og_model_data2$old_gravity)

# Check distribution

( recip_sqrt_market_grav_hist2 <- ggplot(og_model_data2, 
                                        aes(x = og_recipr_sqrt)) +
  geom_histogram(bins = 20) +
  labs(
    x = expression("Reciprocal Square-Root Old Gravity (1 / sqrt(Old Gravity))"),
    y = "Frequency"
  ) +
  theme_bw() )
```

The log-transformed gravity variable exhibits a substantially improved distribution relative to the raw scale. The strong right skew observed in the untransformed data is markedly reduced, and the values are more evenly distributed across the range. Although some mild skewness persists, the transformation compresses extreme high values and reduces leverage from previously influential observations. This suggests that a log transformation is effective in stabilising variance and improving conformity with linear modelling assumptions.

The reciprocal square-root transformation also substantially alters the distribution relative to the raw scale. By strongly compressing large gravity values and stretching differences among smaller values, it reduces the influence of extreme high-gravity observations. The resulting distribution appears more symmetric than the original variable, although some skewness and clustering remain. Compared to the log transformation, the reciprocal square-root places greater weight on variation at the lower end of the gravity gradient, which may reduce leverage from high values but can also concentrate variability among low-gravity sites. Overall, this transformation is effective in mitigating extreme right-tail influence, but it should be noted that its interpretability is less straightforward than the log scale.

#### Transformed models: log

Model specification To evaluate whether transformation improves model performance, we fitted three alternative linear models: (i) log-transformed biomass regressed on untransformed gravity, (ii) untransformed biomass regressed on log-transformed gravity, and (iii) a fully log–log model with both variables log-transformed. The gravity predictor was mean-centred prior to modelling to ensure that intercepts represent expected biomass at average gravity and to maintain numerical stability:

```{r}

# Centre logged market gravity
og_model_data2$log_og_centred <- as.numeric(
  scale(og_model_data2$log_og, center = TRUE, scale = FALSE)
)

# Transformed models 
log_biomass_og2 <- lm(log_biomass~og_centred,data = og_model_data2) 
biomass_log_og2 <- lm(mean_biomass~log_og_centred,data = og_model_data2) 
log_biomass_log_og2 <- lm(log_biomass~log_og_centred,data = og_model_data2) 

# Summaries of transformed models
summary(log_biomass_og2) 
summary(biomass_log_og2) 
summary(log_biomass_log_og2) 

# or arm::display(lm_biomass_chla)
```

Across all transformed specifications, there remains no statistically supported relationship between human gravity and biomass. In the model with log-transformed biomass and untransformed gravity, the slope is negative but non-significant (β = −7.69 × 10⁻⁵, p = 0.268), explaining approximately 2% of the variance (R² = 0.020). When biomass is modelled against log-transformed gravity, explanatory power further declines (R² = 0.003), and the slope remains non-significant (p = 0.687). The fully log–log model similarly shows no evidence of association (β = −0.0196, p = 0.812), with negligible explanatory power (R² \< 0.001).

Thus, although transformations improve distributional properties, they do not materially strengthen the relationship between gravity and biomass. The consistently small effect sizes and very low R² values indicate that human gravity, as measured here, provides minimal explanatory value for variation in site-level biomass.

#### Transformed models: reciprocal sqrt

Model specification To assess whether the Box–Cox–suggested reciprocal square-root transformation (λ ≈ −0.5) improves model performance, we fitted two linear models using the transformed gravity predictor: (i) untransformed biomass regressed on reciprocal square-root gravity, and (ii) log-transformed biomass regressed on reciprocal square-root gravity. The transformed gravity variable was mean-centred prior to analysis so that intercepts represent expected biomass at average transformed gravity and to maintain numerical stability:

```{r}

# Centre reciprocal sqrt market gravity
og_model_data$og_recipr_sqrt_centred <- as.numeric(
  scale(og_model_data$og_recipr_sqrt, center = TRUE, scale = FALSE)
)

# Transformed models 
biomass_rsqrt_og <- lm(mean_biomass~og_recipr_sqrt_centred,data = og_model_data)
log_biomass_rsqrt_og <- lm(log_biomass~og_recipr_sqrt_centred,data = og_model_data) 

# Summaries of transformed models
summary(biomass_rsqrt_og) 
summary(log_biomass_rsqrt_og) 

```

The model with untransformed biomass shows no evidence of association between biomass and reciprocal square-root gravity (β = −2011, SE = 36874, p = 0.957). The explanatory power is effectively zero (R² ≈ 0.00005), indicating that the transformed predictor accounts for none of the variation in biomass.

Similarly, when biomass is log-transformed, the relationship remains non-significant (β = −0.164, SE = 0.707, p = 0.818), with negligible explanatory power (R² ≈ 0.0009).

Thus, although the reciprocal square-root transformation improves the distributional properties of the gravity variable, it does not strengthen the statistical relationship between gravity and biomass. Across specifications, effect sizes remain small, highly uncertain, and biologically negligible.

#### Residual checking of transformations

To determine whether transformation improved model adequacy, we examined the same set of diagnostic plots (residuals vs fitted, Q–Q plots, scale–location plots, and Cook’s distance) for each transformed specification:

```{r}

# log_biomass_old_market_grav
par(mfrow= c(2,2)) # divide the page into 2 x 2 frame
plot(biomass_log_og, which= c(1:4))

# biomass_log_old_market_grav
par(mfrow= c(2,2)) # divide the page into 2 x 2 frame
plot(log_biomass_og, which= c(1:4))

# log_biomass_log_old_market_grav
par(mfrow= c(2,2)) # divide the page into 2 x 2 frame
plot(log_biomass_log_og, which= c(1:4))

# biomass_rsqrt_old_market_grav
par(mfrow= c(2,2)) # divide the page into 2 x 2 frame
plot(biomass_rsqrt_og, which= c(1:4))

# log_biomass_rsqrt_old_market_grav
par(mfrow= c(2,2)) # divide the page into 2 x 2 frame
plot(log_biomass_rsqrt_og, which= c(1:4))
```

Results: In the model with log-transformed gravity only (mean_biomass \~ log_gravity), the residual structure remains broadly similar to the untransformed model. Heteroscedasticity persists, with increasing spread at higher fitted values, and the Q–Q plot still shows deviation in the upper tail. Influential observations remain present.

When biomass alone is log-transformed (log_biomass \~ gravity), residual variance is noticeably more stable across fitted values, and the Q–Q plot shows improved alignment with the normal reference line, although mild tail deviations remain. This indicates partial improvement in normality and homoscedasticity.

The fully log–log model (log_biomass \~ log_gravity) shows the most consistent variance structure, with residuals more evenly distributed and closer to normality, though slight upper-tail deviations and a few influential observations persist.

Models incorporating the reciprocal square-root transformation of gravity show a similar pattern: transforming gravity alone does not resolve heteroscedasticity or tail deviations, whereas models with log-transformed biomass exhibit improved variance stability and closer adherence to normality. However, the reciprocal square-root transformation provides no clear diagnostic advantage over the log transformation of gravity.

Overall, transforming the response variable (log biomass) substantially improves adherence to model assumptions, whereas transforming gravity alone—whether via log or reciprocal square-root—offers limited benefit.

### Check AICs

Because AIC values are only comparable among models fitted to the same response variable on the same scale, model selection was restricted to specifications with log-transformed biomass as the response. We therefore compared the models including untransformed gravity, log-transformed gravity, and reciprocal square-root gravity as predictors:

```{r}
og_base_models <- list(log_biomass_og, log_biomass_log_og, log_biomass_rsqrt_og)
aictab(og_base_models)
```

Model selection based on AICc indicates that the model with untransformed gravity predicting log biomass (Mod1) provides the best relative fit (AICc = 44.51). However, the log–log and reciprocal square-root models (Mod2 and Mod3) are within ΔAICc = 1.22 of the top model. Because models within ΔAICc \< 2 are generally considered to have substantial support, all three models are statistically competitive.

So, to conclude: When comparing functional forms of the gravity predictor (untransformed, log-transformed, and reciprocal square-root), there is no meaningful difference in inference. All three specifications yield very small, non-significant slope estimates, extremely low R² values, and nearly identical AICc scores. The ΔAICc values (\< 2) indicate that the models are statistically indistinguishable in terms of fit.

Residual diagnostics show that improvements in model assumptions are primarily driven by transforming the response variable (log biomass), not by transforming gravity. Changing the functional form of gravity—whether log or reciprocal square-root—does not materially improve variance structure, normality, or explanatory power relative to using gravity on its original scale.

Overall, there is no statistical or ecological justification for preferring a transformed gravity predictor over the untransformed version. The absence of a detectable effect of gravity on biomass is robust to the choice of transformation.

### Taking into account eco-region (country)

#### Fixed-effects model (separate)

We will first separate this into each different eco-region (country for now) We will use log for ease of interpretation

```{r}
# plot of biomass vs chlorophyll for a 'no pooling analysis'
( separate_lm_plot <- ggplot(og_model_data, aes(x = log_og_centred,
           y = log_biomass)) + geom_point() + geom_smooth(method = "lm", formula = y ~ x) +
  facet_grid(. ~ country) + theme_bw() + labs( x = " Logged Market Gravity",
    y = "Log-Biomass", title = "No pooling analysis") )

```

Differences across countries!

This analysis treats the relationship at each site as unique and independent of the others.

#### Combined fixed-effects model

At the other extreme, we could simply pool all the data together in a single simple regression analysis: (same as above).

```{r}
# Graph the data - X-Y scatter plot
og_combined_lm_plot <- og_model_data %>% 
  ggplot(aes(x=log_og_centred, y=log_biomass)) + 
  geom_point() +
  geom_smooth(method=lm)

og_combined_lm_plot
```

This model helps us to understand the analysis we want to build from these initial simple extremes.

#### Country as a fixed effect

Note: - from above, we will carry on with the log_log model - we centre chla. In models including interactions, centering also makes main effects interpretable as effects evaluated at the mean of the interacting variable and can improve numerical stability in mixed-effects models. - We used sum-to-zero contrasts (effect coding) for country so that parameter estimates represent deviations from the overall mean rather than differences relative to an arbitrarily chosen reference country. Under effect coding, the intercept corresponds to the grand mean of the response (averaged across all observations), and each country coefficient represents that country’s deviation from this grand mean. In other words, coefficients quantify how much higher or lower a given country’s mean is compared to the overall mean.

##### Additive model

First we add it in an additive simple least squares regression. This additive model assumes that the biomass–chlorophyll slope is the same across all countries, but allows countries to differ in their mean biomass levels:

```{r}
# Set sum-to-zero contrasts for country (the grand mean is the baseline - not an arbitrary country level)
contrasts(og_model_data$country) <- contr.sum(length(levels(og_model_data$country)))

# Addtitive model
og_country_additive <- lm(log_biomass ~ log_og_centred + country, data = og_model_data)
summary(og_country_additive) 
```

Output interpretation: ...

##### Interaction model

This interaction model allows the biomass–chlorophyll slope to differ among countries rather than forcing a common slope.

```{r}

# Log-log interaction
og_country_interaction <- lm(log_biomass ~ log_og_centred * country, data = og_model_data)
summary(og_country_interaction) 

```

...

#### Test global, additive and interaction models

```{r}
#AIC
interaction_test_models_og <- list(log_biomass_log_og, og_country_additive, og_country_interaction)
aictab(interaction_test_models_og) 

# Classical F-test
anova(log_biomass_log_og, og_country_additive, og_country_interaction)
```

It seems that not interaction, but additive mdoel does not add loads...

## Market gravity

```{r}

# Histogram market gravity 
( gravity_hist <- ggplot(gravity_metrics_2009_laura, aes(x = market_gravity)) +
  geom_histogram(bins = 10) +
  labs(
    x = expression("Market  gravity"),
    y = "Number of sites"
  ) +
  theme_bw() )

# Boxcox 
boxcox(lm(market_gravity ~ 1, data = gravity_metrics_2009_laura),
       lambda = seq(-2, 2, 0.1))

```

Outputs: Market gravity is extremely right skewed. The vast majority of sites have very low (or zero) values of market gravity, with only a small number of sites exhibiting very large values (in the millions). This indicates a heavy-tailed distribution with strong imbalance. Since it is not normally distributed, it will violate Gaussian assumptions if used on the raw scale. Any linear model on the untransformed scale would be dominated by the few extreme values.

Additionally, since the dynamic range spans several orders of magnitude. That strongly suggests that effects, if present, are likely multiplicative rather than additive.

The Box–Cox analysis confirms that market gravity should be log-transformed before modelling. Note that there is no need to use the exact maximum-likelihood λ; log is statistically justified and substantively coherent for a gravity-type variable spanning orders of magnitude.

### Exploratory Plot

```{r}
#| label: biomass-vs-gravity-plot

# Graph the data - X-Y scatter plot
biomass_mg_fig2 <- merge(x = biomass_per_site_laura, y = gravity_metrics_2009_laura, by = "site", all = TRUE) %>%
  ggplot(aes(x=market_gravity, y=mean_biomass)) + 
  geom_point() +
  geom_smooth(method=lm)
biomass_mg_fig2

# Correlation
with(merge(x = biomass_per_site_laura, y = gravity_metrics_2009_laura, by = "site", all = TRUE),
    cor(market_gravity, mean_biomass))

```

Human gravity seems to have a negative effect on biomass, but the are large confidence intervals, and a lot of uncertainty in the data.

### Untransformed ls model

model 1 - the simplest untransforemd regression

```{r}
#| label: biomass-vs-chla-lm

# Merge needed data
mg_model_data2 <- merge(biomass_per_site_laura, gravity_metrics_2009_laura[, c("site", "market_gravity")], by = "site", all = TRUE)


# Centre market gravity
mg_model_data2$mg_centred <- as.numeric(scale(
  mg_model_data2$market_gravity,
  center = TRUE,
  scale = FALSE ))

# Build lm 
base_model_mg2 <- lm(mean_biomass~mg_centred,data = mg_model_data2) 
```

Analyse model:

```{r}
# Analyse the model 
coef(base_model_mg)
summary(base_model_mg) 
confint(base_model_mg) 
```

...

#### Residual checking of simple ls model

But model assumptions and violations:

```{r}
par(mfrow= c(2,2)) # divide the page into 2 x 2 frame
# par(mar=c(1,1,1,1))
plot( base_model_mg, which= c(1:4))
```

...

### Transformations

To improve model fit, we can transform the raw data to better conform to the regression assumptions, thereby amplifying predictive power and increasing the overall quality of the model.

We also check how the distribution of the data changes after each transformation to ensure that everything is working as it should.

```{r}

# Transforming variables
mg_model_data2$log_biomass <- log10(mg_model_data2$mean_biomass)
mg_model_data2$log_market_grav <- log10(mg_model_data2$market_gravity)

# Check transformations 

( log_market_grav_hist <- ggplot(mg_model_data, aes(x = log_market_grav)) +
  geom_histogram(bins = 20) +
  labs(
    x = expression("Log-Transformed Market Gravity"),
    y = "Frequency"
  ) +
  theme_bw() )
```

The log-transformed distribution is still somewhat skewed, but it is dramatically compressed relative to the raw scale. The extreme values are no longer dominating the range. The distribution also looks mildly bimodal or clustered rather than smoothly normal. That is not necessarily a transformation problem. It may reflect real structural differences (e.g., remote vs highly connected sites, or country-level clustering). Transformations address skewness, not underlying multimodality caused by real grouping.

Transformed models: Note that we will be centering the gravity data. In most ecological regression contexts, especially with interaction terms, centering is preferable because it avoids interpreting coefficients at unrealistic covariate values, It reduces collinearity with interaction terms, it improves numerical stability.

```{r}

# Centre logged market gravity
mg_model_data2$log_mg_centred <- as.numeric(
  scale(mg_model_data2$log_market_grav, center = TRUE, scale = FALSE)
)

# Transformed models 
log_biomass_mg <- lm(log_biomass~mg_centred,data = mg_model_data) 
biomass_log_mg<- lm(mean_biomass~log_mg_centred,data = mg_model_data) 
log_biomass_log_mg <- lm(log_biomass~log_mg_centred,data = mg_model_data) 

# Summaries of transformed models
summary(log_biomass_mg) 
summary(biomass_log_mg) 
summary(log_biomass_log_mg) 

# or arm::display(lm_biomass_chla)
#
```

...

### Residual checking of transformations

```{r}

# log_biomass_chla
par(mfrow= c(2,2)) # divide the page into 2 x 2 frame
plot(biomass_log_mg, which= c(1:4))

# biomass_log_chla
par(mfrow= c(2,2)) # divide the page into 2 x 2 frame
plot(log_biomass_mg, which= c(1:4))

# log_biomass_log_chla
par(mfrow= c(2,2)) # divide the page into 2 x 2 frame
plot(log_biomass_log_mg, which= c(1:4))
```

the log–log form better satisfies linear model assumptions ...

### Check AICs

Note that you cannot compare AIC across different response scales, so, given that log-biomass better fits the residuals, we are interested to know which model - log_biomass_chla or log-log - provides the best fit

```{r}
grav_base_models <- list(log_biomass_mg, log_biomass_log_mg)
aictab(grav_base_models)
```

...

### Comparing gravities

```{r}
grav_metric_models <- list(log_biomass_log_og, log_biomass_log_mg)
aictab(grav_metric_models)

anova(log_biomass_log_og,log_biomass_log_mg)
```

Perhaps as would be expected they are very very similar in fit...

### Taking into account eco-region (country)

#### Fixed-effects model (separate)

We will first separate this into each different eco-region (country for now)

```{r}
# plot of biomass vs chlorophyll for a 'no pooling analysis'
( mg_separate_plot <- ggplot(mg_model_data, aes(x = log_mg_centred,
           y = log_biomass)) + geom_point() + geom_smooth(method = "lm", formula = y ~ x) +
  facet_grid(. ~ country) + theme_bw() + labs( x = " Logged Old Market Gravity",
    y = "Log-Biomass", title = "No pooling analysis") )

```

Great differences across countries!!

This analysis treats the relationship at each site as unique and independent of the others.

#### Combined fixed-effects model

At the other extreme, we could simply pool all the data together in a single simple regression analysis: (same as above).

```{r}
# Graph the data - X-Y scatter plot
mg_combined_plot <- mg_model_data %>% 
  ggplot(aes(x=log_mg_centred, y=log_biomass)) + 
  geom_point() +
  geom_smooth(method=lm)

mg_combined_plot
```

This model helps us to understand the analysis we want to build from these initial simple extremes.

#### Country as a fixed effect

Note: - from above, we will carry on with the log_log model - we centre chla. In models including interactions, centering also makes main effects interpretable as effects evaluated at the mean of the interacting variable and can improve numerical stability in mixed-effects models. - We used sum-to-zero contrasts (effect coding) for country so that parameter estimates represent deviations from the overall mean rather than differences relative to an arbitrarily chosen reference country. Under effect coding, the intercept corresponds to the grand mean of the response (averaged across all observations), and each country coefficient represents that country’s deviation from this grand mean. In other words, coefficients quantify how much higher or lower a given country’s mean is compared to the overall mean.

##### Additive model

First we add it in an additive simple least squares regression. This additive model assumes that the biomass–chlorophyll slope is the same across all countries, but allows countries to differ in their mean biomass levels:

```{r}
# Set sum-to-zero contrasts for country (the grand mean is the baseline - not an arbitrary country level)
contrasts(mg_model_data$country) <- contr.sum(length(levels(mg_model_data$country)))

# Addtitive model
mg_country_additive <- lm(log_biomass ~ log_mg_centred + country, data = mg_model_data)
summary(mg_country_additive) 
```

Output interpretation: ...

##### Interaction model

This interaction model allows the biomass–chlorophyll slope to differ among countries rather than forcing a common slope.

```{r}

# Log-log interaction
mg_country_interaction <- lm(log_biomass ~ log_mg_centred * country, data = mg_model_data)
summary(mg_country_interaction) 

```

...

##### Test global, additive and interaction models

```{r}
#AIC
interaction_test_models_mg <- list(log_biomass_log_mg, mg_country_additive, mg_country_interaction)
aictab(interaction_test_models_mg) 

# Classical F-test
anova(log_biomass_log_mg, mg_country_additive, mg_country_interaction)
```

Note that again there isn't a huge different between global and additive model, but given that we want to keep the ecoregion effect, we will keep it for now :)

## Settlement gravity (settlement top pop)

```{r}

# Histogram market gravity 
( st_hist <- ggplot(gravity_metrics_2009_laura, aes(x = settlement_tot_pop)) +
  geom_histogram(bins = 10) +
  labs(
    x = expression("Settlement tot pop"),
    y = "Number of sites"
  ) +
  theme_bw() )

# Boxcox 
boxcox(lm(settlement_tot_pop ~ 1, data = gravity_metrics_2009_laura),
       lambda = seq(-2, 2, 0.1))

```

Outputs: Settlement gravity is extremely right skewed. The Box–Cox analysis confirms that settlement gravity should be log-transformed before modelling. Note that there is no need to use the exact maximum-likelihood λ; log is statistically justified and substantively coherent for a gravity-type variable spanning orders of magnitude.

### Exploratory Plot

```{r}
#| label: biomass-vs-st-plot

# Graph the data - X-Y scatter plot
biomass_st_fig2 <- merge(x = biomass_per_site_laura, y = gravity_metrics_2009_laura, by = "site", all = TRUE) %>%
  ggplot(aes(x=settlement_tot_pop, y=mean_biomass)) + 
  geom_point() +
  geom_smooth(method=lm)

biomass_st_fig2

# Correlation
with(merge(x = biomass_per_site_laura, y = gravity_metrics_2009_laura, by = "site", all = TRUE),
    cor(settlement_tot_pop, mean_biomass))

```

Human gravity seems to have a negative effect on biomass, but the are large confidence intervals, and a lot of uncertainty in the data.

### Untransformed ls model

model 1 - the simplest untransformed regression

```{r}
#| label: biomass-vs-chla-lm

# Merge needed data
st_model_data2 <- merge(biomass_per_site_laura, gravity_metrics_2009_laura[, c("site", "settlement_tot_pop")], by = "site", all = TRUE)

# Centre market gravity
st_model_data2$st_centred <- as.numeric(scale(
  st_model_data2$settlement_tot_pop,
  center = TRUE,
  scale = FALSE ))

# Build lm 
base_model_st <- lm(mean_biomass~st_centred,data = st_model_data2) 
```

Analyse model:

```{r}
# Analyse the model 
coef(base_model_st)
summary(base_model_st) 
confint(base_model_st) 
```

...

#### Residual checking of simple ls model

But model assumptions and violations:

```{r}
par(mfrow= c(2,2)) # divide the page into 2 x 2 frame
# par(mar=c(1,1,1,1))
plot( base_model_st, which= c(1:4))
```

...

### Transformations

To improve model fit, we can transform the raw data to better conform to the regression assumptions, thereby amplifying predictive power and increasing the overall quality of the model.

We also check how the distribution of the data changes after each transformation to ensure that everything is working as it should.

```{r}

# Transforming variables
st_model_data2$log_biomass <- log10(st_model_data2$mean_biomass)
st_model_data2$log_settlement_tot <- log10(st_model_data2$settlement_tot_pop)

# Check transformations 

( log_st_hist <- ggplot(st_model_data, aes(x = log_settlement_tot)) +
  geom_histogram(bins = 20) +
  labs(
    x = expression("Log-Transformed Settlement Tot Pop"),
    y = "Frequency"
  ) +
  theme_bw() )
```

The distribution also looks mildly bimodal or clustered rather than smoothly normal. That is not necessarily a transformation problem. It may reflect real structural differences (e.g., remote vs highly connected sites, or country-level clustering). Transformations address skewness, not underlying multimodality caused by real grouping.

Transformed models: Note that we will be centering the gravity data. In most ecological regression contexts, especially with interaction terms, centering is preferable because it avoids interpreting coefficients at unrealistic covariate values, It reduces collinearity with interaction terms, it improves numerical stability.

```{r}

# Centre logged market gravity
st_model_data2$log_st_centred <- as.numeric(
  scale(st_model_data2$log_settlement_tot, center = TRUE, scale = FALSE)
)

# Transformed models 
log_biomass_st <- lm(log_biomass~st_centred,data = st_model_data) 
biomass_log_st<- lm(mean_biomass~log_st_centred,data = st_model_data) 
log_biomass_log_st <- lm(log_biomass~log_st_centred,data = st_model_data) 

# Summaries of transformed models
summary(log_biomass_st) 
summary(biomass_log_st) 
summary(log_biomass_log_st) 

```

...

### Residual checking of transformations

```{r}

# log_biomass_chla
par(mfrow= c(2,2)) # divide the page into 2 x 2 frame
plot(biomass_log_st, which= c(1:4))

# biomass_log_chla
par(mfrow= c(2,2)) # divide the page into 2 x 2 frame
plot(log_biomass_st, which= c(1:4))

# log_biomass_log_chla
par(mfrow= c(2,2)) # divide the page into 2 x 2 frame
plot(log_biomass_log_st, which= c(1:4))
```

the log–log form better satisfies linear model assumptions ...

### Check AICs

Note that you cannot compare AIC across different response scales, so, given that log-biomass better fits the residuals, we are interested to know which model - log_biomass_chla or log-log - provides the best fit

```{r}
st_base_models_2 <- list(log_biomass_st, log_biomass_log_st)
aictab(st_base_models)
```

...

### Comparing gravities

```{r}
grav_metric_models2 <- list(log_biomass_log_mg, log_biomass_log_st)
aictab(grav_metric_models2)

anova(log_biomass_log_mg,log_biomass_log_st)
```

...

## Settlement gravity (nearest pop grav)

```{r}

# Histogram market gravity 
( nearest_pop_hist <- ggplot(gravity_metrics_2009_laura, aes(x = nearest_pop_grav)) +
  geom_histogram(bins = 20) +
  labs(
    x = expression("Nearest Pop Grav"),
    y = "Number of sites"
  ) +
  theme_bw() )

# Boxcox 
boxcox(lm(nearest_pop_grav ~ 1, data = gravity_metrics_2009_laura),
       lambda = seq(-2, 2, 0.1))

```

Outputs: Settlement gravity is extremely right skewed. The Box–Cox analysis confirms that settlement gravity should be log-transformed before modelling. Note that there is no need to use the exact maximum-likelihood λ; log is statistically justified and substantively coherent for a gravity-type variable spanning orders of magnitude.

### Exploratory Plot

```{r}
#| label: biomass-vs-gravity-plot

# Graph the data - X-Y scatter plot
biomass_sg_fig2 <- merge(x = biomass_per_site_laura, y = gravity_metrics_2009_laura, by = "site", all = TRUE) %>%
  ggplot(aes(x=nearest_pop_grav, y=mean_biomass)) + 
  geom_point() +
  geom_smooth(method=lm)

biomass_sg_fig2

# Correlation
with(merge(x = biomass_per_site_laura, y = gravity_metrics_2009_laura, by = "site", all = TRUE),
    cor(nearest_pop_grav, mean_biomass))

```

Human gravity seems to have a negative effect on biomass, but the are large confidence intervals, and a lot of uncertainty in the data.

### Untransformed ls model

model 1 - the simplest untransforemd regression

```{r}
#| label: biomass-vs-chla-lm

# Merge needed data
sg_model_data2 <- merge(biomass_per_site_laura, gravity_metrics_2009_laura[, c("site", "nearest_pop_grav")], by = "site", all = TRUE)

# Centre market gravity
sg_model_data2$sg_centred <- as.numeric(scale(
  sg_model_data2$nearest_pop_grav,
  center = TRUE,
  scale = FALSE ))

# Build lm 
base_model_sg2 <- lm(mean_biomass~sg_centred,data = sg_model_data2) 
```

Analyse model:

```{r}
# Analyse the model 
coef(base_model_sg2)
summary(base_model_sg2) 
confint(base_model_sg2) 
```

...

#### Residual checking of simple ls model

But model assumptions and violations:

```{r}
par(mfrow= c(2,2)) # divide the page into 2 x 2 frame
# par(mar=c(1,1,1,1))
plot( base_model_sg2, which= c(1:4))
```

...

### Transformations

To improve model fit, we can transform the raw data to better conform to the regression assumptions, thereby amplifying predictive power and increasing the overall quality of the model.

We also check how the distribution of the data changes after each transformation to ensure that everything is working as it should.

```{r}

# Transforming variables
sg_model_data2$log_biomass <- log10(sg_model_data2$mean_biomass)
sg_model_data2$log_sg <- log10(sg_model_data2$nearest_pop_grav)

# Check transformations 

( log_sg_hist <- ggplot(sg_model_data, aes(x = log_sg)) +
  geom_histogram(bins = 20) +
  labs(
    x = expression("Log-Transformed Settlement Tot Pop"),
    y = "Frequency"
  ) +
  theme_bw() )
```

The distribution also looks mildly bimodal or clustered rather than smoothly normal. That is not necessarily a transformation problem. It may reflect real structural differences (e.g., remote vs highly connected sites, or country-level clustering). Transformations address skewness, not underlying multimodality caused by real grouping.

Transformed models: Note that we will be centering the gravity data. In most ecological regression contexts, especially with interaction terms, centering is preferable because it avoids interpreting coefficients at unrealistic covariate values, It reduces collinearity with interaction terms, it improves numerical stability.

```{r}

# Centre logged market gravity
sg_model_data2$log_sg_centred <- as.numeric(
  scale(sg_model_data2$log_sg, center = TRUE, scale = FALSE)
)

# Transformed models 
log_biomass_sg2 <- lm(log_biomass~sg_centred,data = sg_model_data2) 
biomass_log_sg2 <- lm(mean_biomass~log_sg_centred,data = sg_model_data2) 
log_biomass_log_sg2 <- lm(log_biomass~log_sg_centred,data = sg_model_data2) 

# Summaries of transformed models
summary(log_biomass_sg2) 
summary(biomass_log_sg2) 
summary(log_biomass_log_sg2) 

```

...

### Residual checking of transformations

```{r}

# log_biomass_chla
par(mfrow= c(2,2)) # divide the page into 2 x 2 frame
plot(biomass_log_sg, which= c(1:4))

# biomass_log_chla
par(mfrow= c(2,2)) # divide the page into 2 x 2 frame
plot(log_biomass_sg, which= c(1:4))

# log_biomass_log_chla
par(mfrow= c(2,2)) # divide the page into 2 x 2 frame
plot(log_biomass_log_sg, which= c(1:4))
```

the log–log form better satisfies linear model assumptions ...

### Check AICs

Note that you cannot compare AIC across different response scales, so, given that log-biomass better fits the residuals, we are interested to know which model - log_biomass_chla or log-log - provides the best fit

```{r}
sg_base_models <- list(log_biomass_sg, log_biomass_log_sg)
aictab(sg_base_models)
```

Similar but...

### Comparing gravities

```{r}
grav_metric_models_all <- list(log_biomass_log_og, log_biomass_log_mg, log_biomass_log_st, log_biomass_log_sg)
aictab(grav_metric_models_all)

anova(log_biomass_log_og, log_biomass_log_mg, log_biomass_log_st, log_biomass_log_sg)
```

Settlement gravity seems to fit better!! So we keep this going forward

### Taking into account eco-region (country)

#### Fixed-effects model (separate)

We will first separate this into each different eco-region (country for now)

```{r}
# plot of biomass vs chlorophyll for a 'no pooling analysis'
( sg_separate_plot <- ggplot(sg_model_data, aes(x = log_sg_centred,
           y = log_biomass)) + geom_point() + geom_smooth(method = "lm", formula = y ~ x) +
  facet_grid(. ~ country) + theme_bw() + labs( x = " Logged Settlement Gravity",
    y = "Log-Biomass", title = "No pooling analysis") )

```

Great differences across countries!!

This analysis treats the relationship at each site as unique and independent of the others.

#### Combined fixed-effects model

At the other extreme, we could simply pool all the data together in a single simple regression analysis: (same as above).

```{r}
# Graph the data - X-Y scatter plot
sg_combined_plot <- sg_model_data %>% 
  ggplot(aes(x=log_sg_centred, y=log_biomass)) + 
  geom_point() +
  geom_smooth(method=lm)

sg_combined_plot
```

This model helps us to understand the analysis we want to build from these initial simple extremes.

#### Country as a fixed effect

Note: - from above, we will carry on with the log_log model - we centre chla. In models including interactions, centering also makes main effects interpretable as effects evaluated at the mean of the interacting variable and can improve numerical stability in mixed-effects models. - We used sum-to-zero contrasts (effect coding) for country so that parameter estimates represent deviations from the overall mean rather than differences relative to an arbitrarily chosen reference country. Under effect coding, the intercept corresponds to the grand mean of the response (averaged across all observations), and each country coefficient represents that country’s deviation from this grand mean. In other words, coefficients quantify how much higher or lower a given country’s mean is compared to the overall mean.

##### Additive model

First we add it in an additive simple least squares regression. This additive model assumes that the biomass–chlorophyll slope is the same across all countries, but allows countries to differ in their mean biomass levels:

```{r}
# Set sum-to-zero contrasts for country (the grand mean is the baseline - not an arbitrary country level)
contrasts(sg_model_data$country) <- contr.sum(length(levels(sg_model_data$country)))

# Addtitive model
sg_country_additive <- lm(log_biomass ~ log_sg_centred + country, data = sg_model_data)
summary(sg_country_additive) 
```

Output interpretation: ...

##### Interaction model

This interaction model allows the biomass–chlorophyll slope to differ among countries rather than forcing a common slope.

```{r}

# Log-log interaction
sg_country_interaction <- lm(log_biomass ~ log_sg_centred * country, data = sg_model_data)
summary(sg_country_interaction) 

```

...

##### Test global, additive and interaction models

```{r}
#AIC
interaction_test_models_sg <- list(log_biomass_log_sg, sg_country_additive, sg_country_interaction)
aictab(interaction_test_models_sg) 

# Classical F-test
anova(log_biomass_log_sg, sg_country_additive, sg_country_interaction)
```

Note that again the no-country model works a lot better!! Non-additive looks a lot better!

May be better to work with connectivity?

# GLM models

We fitted a generalized linear model (GLM) rather than applying a transformation to the response variable because GLMs allow the mean–variance relationship of the data to be modelled explicitly through an appropriate error distribution and link function. In contrast, transforming the response (e.g., log-transforming biomass) changes the scale of the outcome and can complicate interpretation, while still relying on Gaussian error assumptions.

First of all we will produce a comparative GLM to ensure that we can test the transformation again.

This is a base model but we will keep the log transformed data as mentioned before because it fits the assumptions better

```{r}
#Fitting a GLM
glm_gaussian_sg2 <- glm(
  mean_biomass ~ log_sg_centred,
  family = gaussian(link = "identity"),
  data = sg_model_data2)

summary(glm_gaussian_sg2)
confint(glm_gaussian_sg2)
```

Next, a Gamma error distribution with a log link was used because site-level biomass is strictly positive and strongly right-skewed, with variance increasing as the mean increases. The Gamma distribution is well suited to modelling continuous, positive data with this mean–variance structure, and the log link allows effects of settlement gravity to be interpreted multiplicatively (i.e., proportional changes in biomass). Note that we are still using the centred version of settlement gravity. Centering does not change inference about the effect of gravity. It only improves interpretability of the intercept and helps if you later add interactions (e.g., country × gravity). We will aslo explicitly test the tranform x with the non-transformed again.

```{r}

#Fitting a GLM
glm_gamma_sg2 <- glm(
  mean_biomass ~ sg_centred,
  family = Gamma(link = "log"),
  data = sg_model_data2)

summary(glm_gamma_sg2)
confint(glm_gamma_sg2)

glm_gamma_sg3 <- glm(
  mean_biomass ~ log_sg_centred,
  family = Gamma(link = "log"),
  data = sg_model_data2)

summary(glm_gamma_sg3)
confint(glm_gamma_sg3)
```

Interpretation: Settlement gravity was significantly negatively associated with site-level fish biomass under a generalized linear model with a Gamma error distribution and log link. The coefficient for log-transformed and mean-centred settlement gravity was −0.174 (SE = 0.058, t = −3.01, p = 0.0038), indicating a statistically significant decline in biomass with increasing human gravity. The 95% confidence interval for the slope (−0.285 to −0.049) did not overlap zero, confirming the robustness of this negative relationship.

Because a log link was used, coefficients are interpreted multiplicatively on the original biomass scale. Exponentiating the slope indicates that a one-unit increase in log settlement gravity is associated with an approximately 16% reduction in expected biomass (95% CI: 5–25%). Thus, increases in human population gravity are associated with proportionally lower reef fish biomass across sites.

Model fit statistics indicated moderate explanatory power. The model reduced deviance from 34.90 (null model) to 30.88, corresponding to approximately 12% of deviance explained by settlement gravity alone. The estimated dispersion parameter (0.56) and simulation-based residual diagnostics indicated no evidence of overdispersion or distributional misspecification, supporting the adequacy of the Gamma GLM framework.

Note that both the x transformed and the x raw are significant. Here we are jsut trying to tune up the model.

Now we check the model fit more closely:

```{r}
library(DHARMa)

gamma_sg_res3 <- simulateResiduals(glm_gamma_sg3)
plot(gamma_sg_res3)
testDispersion(gamma_sg_res3)
```

Diagnostic checks using DHARMa indicated that the Gamma generalized linear model with a log link provided an adequate fit to the biomass data. The simulated residuals did not deviate significantly from the expected uniform distribution (KS test: p = 0.496), suggesting that the assumed error distribution is appropriate. The dispersion test was non-significant (dispersion = 1.32, p = 0.272), indicating no evidence of overdispersion or underdispersion. Residuals plotted against fitted values showed no systematic patterns, and no significant outliers were detected (p = 0.391).

Overall, there was no evidence of model misspecification, supporting the use of the Gamma GLM for subsequent inference.

Taken together,these results provide statistically robust evidence that higher settlement gravity is associated with reduced reef fish biomass at the site level.

## Check Gaussian vs Gamma -log model

```{r}
glm_test_models <- list(glm_gaussian_sg2, glm_gamma_sg2, glm_gamma_sg3)
aictab(glm_test_models)
```

Model comparison using AICc strongly supported the Gamma GLM over the Gaussian alternative. The Gamma model had the lowest AICc (1345.76), whereas the Gaussian model had a substantially higher AICc (1388.41), resulting in a ΔAICc of 42.65. Differences in AICc greater than 10 are generally considered decisive evidence against the higher-AIC model; therefore, the Gaussian model receives effectively no support relative to the Gamma specification. The AICc weight for the Gamma model was 1.00, indicating overwhelming support among the candidate models.

These results confirm that modelling biomass with a Gamma error distribution and log link provides a substantially better fit to the data than assuming Gaussian errors. This further justifies the use of the Gamma GLM for inference.

And also using log-transformed x didnt seem to make a huge difference but the residuals are better

## Check gravity metrics again

```{r}
# Old gravity glm
glm_gamma_og2 <- glm(
  mean_biomass ~ log_og_centred,
  family = Gamma(link = "log"),
  data = og_model_data2)

summary(glm_gamma_og2)
confint(glm_gamma_og2)

# Market gravity glm
glm_gamma_mg2 <- glm(
  mean_biomass ~ log_mg_centred,
  family = Gamma(link = "log"),
  data = mg_model_data2)

summary(glm_gamma_mg2)
confint(glm_gamma_mg2)

# Settlement tot pop glm
glm_gamma_st2 <- glm(
  mean_biomass ~ log_st_centred,
  family = Gamma(link = "log"),
  data = st_model_data2)

summary(glm_gamma_st2)
confint(glm_gamma_st2)
```

```{r}
glm_gravity_models <- list(glm_gamma_sg3, glm_gamma_og2, glm_gamma_mg2, glm_gamma_st2)
aictab(glm_gravity_models)
```

Model comparison using AICc indicated that settlement gravity (Mod1) provided the best fit among the candidate gravity metrics. The settlement gravity model had the lowest AICc (1345.76) and the highest model weight (AICcWt = 0.86), indicating strong support relative to the alternatives.

Substantively, this means:

Settlement gravity is not just statistically significant.

It also provides the strongest explanatory power among alternative human-gravity measures.

The evidence for preferring settlement gravity over the others is strong but not absolutely overwhelming (since ΔAICc ≈ 4 for the second-best model).

## Check country as an ecoregion

```{r}

contrasts(sg_model_data$country) <- contr.sum(length(levels(sg_model_data$country)))

glm_country_additive2 <- glm(
  mean_biomass ~ log_sg_centred + country,
  data = sg_model_data2,
  family = Gamma(link = "log"))

summary(glm_country_additive2)

glm_country_interaction2 <- glm(
  mean_biomass ~ log_sg_centred*country,
  data = sg_model_data2,
  family = Gamma(link = "log"))

summary(glm_country_interaction2)

```

```{r}
glm_country_models <- list(glm_gamma_sg3, glm_country_additive2, glm_country_interaction2)
aictab(glm_country_models)
```

Model comparison indicated that the gravity-only model provided the strongest support (AICc = 1345.76, AICcWt = 0.81). Adding country as an additive effect resulted in a modest increase in AICc (ΔAICc = 2.89), indicating limited additional explanatory value. Allowing the effect of gravity to vary among countries (interaction model) was strongly unsupported (ΔAICc = 12.57). These results suggest that the negative relationship between settlement gravity and biomass is broadly consistent across countries and does not vary substantially among eco-regions.

# Total Biomass vs Chlorophyll a

```{r}
#| label: calculate-env-var-means

# Calculate mean dhw per site
chla_2009_sites <- chla_2009 %>% 
  group_by(site) %>% 
  summarise(
    mean_annual_chla = mean(chla_annual_mean),
    .groups = "drop"
  ) %>%
  semi_join(biomass_per_site_laura, by = "site")

```

## Chlor-a
```{r}

# Overview of the data
head(chla_2009_sites,3) # see first three rows
str(chla_2009_sites) #structure of data
summary(chla_2009_sites) #summary of data

# Histogram 
( chla_hist <- ggplot(chla_2009_sites, aes(x = mean_annual_chla)) +
  geom_histogram(bins = 20) +
  labs(
    x = expression("Mean annual chlorophyll-a (mg m"^{-3}*")"),
    y = "Frequency"
  ) +
  theme_bw() )

# Boxcox 
boxcox(lm(mean_annual_chla ~ 1, data = chla_2009_sites),
       lambda = seq(-2, 2, 0.1))

```

Outputs: - 62 sites - Right-skewed distribution: mean \> median and upper tail extends to 3.24 (which is then \~4x median). - The boxcox suggests log-transformation

## Exploratory Plot

```{r}
#| label: biomass-vs-chla-plot

# Graph the data - X-Y scatter plot
biomass_chla_fig1 <- merge(x = biomass_per_site_laura, y = chla_2009_sites, by = "site", all = TRUE) %>% 
  ggplot(aes(x=mean_annual_chla, y=mean_biomass)) + 
  geom_point() +
  geom_smooth(method=lm)

biomass_chla_fig1

# Correlation
with(merge(x = biomass_per_site_laura, y = chla_2009_sites, by = "site", all = TRUE),
    cor(mean_annual_chla, mean_biomass))

```

Chlorophyll a and biomass seem to be weakly positively related, with a correlation of 0.16, but the are large confidence intervals, and a lot of uncertainty in the data.

## Simple ls model

Note that chlorophyll-a, as a continuous predictors, was mean-centred prior to analysis to improve interpretability of model parameters. Centering ensures that the intercept represents the expected response at the average value of the predictor, rather than at zero, which would be biologically unrealistic.

```{r}
#| label: biomass-vs-chla-lm

# Merge needed data
chla_model_data <- merge(x = biomass_per_site_laura, y = chla_2009_sites, by = "site", all = TRUE)

# Centre chla
chla_model_data$mean_chla_centred <- as.numeric(scale(
  chla_model_data$mean_annual_chla,
  center = TRUE,
  scale = FALSE
))

# Build lm 
lm_biomass_chla <- lm(mean_biomass~mean_chla_centred,data = chla_model_data) 
```

Analyse model:

```{r}
# Analyse the model 
coef(lm_biomass_chla)
#residuals(lm_biomass_chla)
summary(lm_biomass_chla) 
# or arm::display(lm_biomass_chla)

confint(lm_biomass_chla) 

```

Outputs:

1.  Intercept: Because chlorophyll was centred, the intercept (21,397) represents the expected biomass at average chlorophyll concentration. The 95% CI (16,982 to 25,811) gives the plausible range for that mean biomass level.

2.  Slopes The slope estimate (3,934) represents the expected increase in biomass for each one-unit increase in chlorophyll above the mean. However, this effect is not statistically significant (p = 0.206), and the confidence interval includes zero (−2,218 to 10,085). This means the data do not provide evidence of a linear relationship between chlorophyll and biomass on the raw scale.

3.  Correlation: 0.162924 (weakly positive)

4.  Effect size The model explains very little variation (R² ≈ 0.027), indicating that chlorophyll alone accounts for only about 3% of the variability in biomass.

Overall, there is no strong support for a simple linear biomass–chlorophyll relationship in this specification.

### Residual checking of simple ls model

Checking model assumptions and violations:

```{r}
par(mfrow= c(2,2)) # divide the page into 2 x 2 frame
plot( lm_biomass_chla, which= c(1:4))

```

Diagnostic plots: - Slight downward trend in residual smooth line - Increasing spread at higher fitted values - Extreme positive outliers (obs 16, 21, 26)

This suggests possible non-linearity (the curve is not flat) and heteroscedasticity (variance increases as fitted values increase) - i.e. residuals are not normally distributed

## Tranformations

To improve model fit, we can transform the raw data to better conform to the regression assumptions, thereby amplifying predictive power and increasing the overall quality of the model.

We also check how the distribution of the data changes after each transformation to ensure that everything is working as it should.

```{r}

# Transforming variables
chla_model_data$log_biomass <- log10(chla_model_data$mean_biomass)
chla_model_data$log_chla <- log10(chla_model_data$mean_annual_chla)

# Check transformations 
( log_biomass_hist <- ggplot(chla_model_data, aes(x = log_biomass)) +
  geom_histogram(bins = 20) +
  labs(
    x = expression("Log-Transformed Biomass"),
    y = "Frequency"
  ) +
  theme_bw() )

( log_chla_hist <- ggplot(chla_model_data, aes(x = log_chla)) +
  geom_histogram(bins = 20) +
  labs(
    x = expression("Log-Transformed Chlorophyll-a"),
    y = "Frequency"
  ) +
  theme_bw() )
```

The data is now looking much more normal!! Great :)

We can now fit models with the transformed data. Ensure you are also centering the log_chla!

```{r}

# Centre logged chla
chla_model_data$log_chla_centred <- as.numeric(
  scale(chla_model_data$log_chla, center = TRUE, scale = FALSE)
)


# Transformed models 
log_biomass_chla <- lm(log_biomass~mean_chla_centred,data = chla_model_data) 
biomass_log_chla <- lm(mean_biomass~log_chla_centred,data = chla_model_data) 
log_biomass_log_chla <- lm(log_biomass~log_chla_centred,data = chla_model_data) 

# Summaries of transformed models
summary(log_biomass_chla) 
summary(biomass_log_chla) 
summary(log_biomass_log_chla) 

# or arm::display(lm_biomass_chla)
#
```

The pattern is very consistent: none of the simple bivariate models show evidence of a relationship between chlorophyll and biomass.

First model: log_biomass \~ mean_annual_chla

The slope is −0.016 (p = 0.79), with R² ≈ 0.001. This indicates essentially no relationship between raw chlorophyll and log-transformed biomass. Chlorophyll explains virtually none of the variation in log biomass.

Second model: mean_biomass \~ log_chla_centred

The slope is 6,048 (p = 0.34), R² ≈ 0.015. This suggests that a one-unit increase in log10 chlorophyll (i.e., a tenfold increase in chlorophyll) would be associated with an increase of about 6,000 biomass units, but this effect is not statistically supported (the confidence interval spans zero).

Because log_chla is centred, the intercept (21,397) represents mean biomass at average log chlorophyll. That makes the intercept interpretable, but it does not change inference.

Third model: log_biomass \~ log_chla_centred (log–log model)

The slope is −0.007 (p = 0.95), R² ≈ 0.00006. This is effectively zero. In scaling terms, the estimated exponent is essentially zero, meaning biomass does not scale with chlorophyll in a simple global power-law relationship.

### Residual checking of transformations

```{r}

# biomass_log_chla
par(mfrow= c(2,2)) # divide the page into 2 x 2 frame
plot(biomass_log_chla, which= c(1:4))

# log_biomass_chla
par(mfrow= c(2,2)) # divide the page into 2 x 2 frame
plot(log_biomass_chla, which= c(1:4))

# log_biomass_log_chla
par(mfrow= c(2,2)) # divide the page into 2 x 2 frame
plot(log_biomass_log_chla, which= c(1:4))
```

Even though neither model shows a significant slope, the log–log form better satisfies linear model assumptions slightly, as it reduces The residuals look the best for the log_log model, with the assumptions now satisfied -\> The log-transformed predictor slightly improves linearity.

### Check AICs

Note that you cannot compare AIC across different response scales, so, given that log-biomass better fits the residuals, we are interested to know which model - log_biomass_chla or log-log - provides the best fit

```{r}
models <- list(log_biomass_chla, log_biomass_log_chla)
aictab(models)
```

When we compare these models we see that they are essentially identical:

ΔAICc = 0.00/0.07 → essentially identical. AIC weights ≈ 0.5 / 0.5 → no meaningful support difference. Log-likelihoods almost identical.

Statistically: The data cannot distinguish between these two models. R²: 0.0012 vs 0.00006s Both slopes non-significant (≈ 0)

So, model comparison based on AICc indicated no meaningful difference between the log–linear and log–log formulations (ΔAICc = 0.07), and both models explained negligible variation in biomass (R² ≈ 0).

However, diagnostic plots suggested slightly improved homoscedasticity and linearity under the log–log specification. Given the equivalent statistical support and the expectation of power-law scaling in ecological productivity relationships, the log–log model was retained for subsequent analyses.

\< Competing models were indistinguishable based on AICc (ΔAICc \< 0.1); however, the log–log specification exhibited improved residual structure and provides a biologically interpretable scaling relationship. We therefore retained the log-transformed predictor. \>

## Taking into account eco-region (country)

### Fixed-effects model (separate)

We will first separate this into each different eco-region (country for now)

```{r}
# plot of biomass vs chlorophyll for a 'no pooling analysis'
( separate_lm_plot <- ggplot(chla_model_data, aes(x = log_chla_centred,
           y = log_biomass)) + geom_point() + geom_smooth(method = "lm", formula = y ~ x) +
  facet_grid(. ~ country) + theme_bw() + labs( x = "Chlorophyll a",
    y = "Log-Biomass", title = "No pooling analysis") )

```

Great differences across countries!!

This analysis treats the relationship at each site as unique and independent of the others.

### Combined fixed-effects model

At the other extreme, we could simply pool all the data together in a single simple regression analysis: (same as above).

```{r}
# Graph the data - X-Y scatter plot
combined_lm_plot <- chla_model_data %>% 
  ggplot(aes(x=log_chla_centred, y=log_biomass)) + 
  geom_point() +
  geom_smooth(method=lm)

combined_lm_plot
```

This model helps us to understand the analysis we want to build from these initial simple extremes.

### Country as a fixed effect

Note: - from above, we will carry on with the log_log model - we centre chla. In models including interactions, centering also makes main effects interpretable as effects evaluated at the mean of the interacting variable and can improve numerical stability in mixed-effects models. - We used sum-to-zero contrasts (effect coding) for country so that parameter estimates represent deviations from the overall mean rather than differences relative to an arbitrarily chosen reference country. Under effect coding, the intercept corresponds to the grand mean of the response (averaged across all observations), and each country coefficient represents that country’s deviation from this grand mean. In other words, coefficients quantify how much higher or lower a given country’s mean is compared to the overall mean.

#### Additive model

First we add it in an additive simple least squares regression. This additive model assumes that the biomass–chlorophyll slope is the same across all countries, but allows countries to differ in their mean biomass levels:

```{r}
# Set sum-to-zero contrasts for country (the grand mean is the baseline - not an arbitrary country level)
contrasts(chla_model_data$country) <- contr.sum(length(levels(chla_model_data$country)))


chla_country_additive <- lm(log_biomass ~ log_chla_centred + country, data = chla_model_data)
summary(chla_country_additive) 
```

Output interpretation:

1.  Intercept Because log chlorophyll is centred, the intercept (4.208) represents the grand mean log biomass at average chlorophyll concentration. Under effect coding, each country coefficient represents that country’s deviation from this overall mean (evaluated at mean chlorophyll), not from a specific reference country.

2.  Slope The slope for log chlorophyll is estiamted to be −0.364 (p = 0.019). This indicates that, when a common slope is imposed across countries, biomass declines with increasing chlorophyll on the log–log scale. In scaling terms, this corresponds to a negative exponent: at higher chlorophyll, biomass is lower, on average. However, this interpretation must be treated cautiously because earlier analyses showed evidence of slope heterogeneity across countries. If slopes differ, this “average” negative slope may reflect opposing patterns cancelling out. So beware with this interpretation!

Country1 has significantly lower biomass than the grand mean (−0.418), while country2 has significantly higher biomass (+0.323), both evaluated at mean chlorophyll. Countries 3 and 4 do not differ significantly from the overall mean.

The model explains about 19% of the variance (R² = 0.19), and the overall model is statistically significant (p = 0.033), indicating that adding country improves fit relative to a simple log–log model without country.

In summary, this additive model suggests that biomass differs among countries and shows a negative average biomass–chlorophyll relationship.

However, given your earlier interaction results, the assumption of a common slope across countries is likely too restrictive, and the interaction model is probably the more appropriate representation of the data.

#### Interaction model

This interaction model allows the biomass–chlorophyll slope to differ among countries rather than forcing a common slope.

```{r}

# Log-log interaction
chla_country_interaction <- lm(log_biomass ~ log_chla_centred * country,
  data = chla_model_data)
summary(chla_country_interaction)

anova(chla_country_interaction)

```

Interpration of outputs:

1.  Intercept Because log chlorophyll is centred and effect coding is used, the intercept (4.276) represents the grand mean log biomass at average chlorophyll across all countries.

2.  Slopes The main effect of log chlorophyll (0.058, p = 0.81) represents the average slope across countries, but in the presence of an interaction this average is not biologically meaningful on its own. The country main effects now represent deviations from the grand mean biomass at mean chlorophyll, conditional on the interaction being included.

3.  Interaction coefficients These are the key terms of the model. These quantify how each country’s slope differs from the overall (grand) slope. Most interaction terms are not statistically different from zero, but the interaction for country4 is strongly negative (−0.899, p = 0.004), indicating that the biomass–chlorophyll scaling relationship in that country differs substantially from the overall trend. Country3 also shows a positive deviation from the mean slope (0.757), with marginal evidence (p = 0.078).

Importantly, the overall model fit improves considerably relative to the additive model. The R² increases from 0.19 to 0.36, and the overall F-test is highly significant (p = 0.003). This indicates that allowing slopes to vary by country explains substantially more variation in biomass.

In substantive terms, there is no single universal biomass–chlorophyll scaling relationship across the region. Instead, the strength and direction of the relationship depend on country, with at least one country exhibiting a markedly different (negative) scaling exponent. This supports the conclusion that the biomass–productivity relationship is spatially heterogeneous rather than globally consistent.

ANOVa - ancova...

#### Test global, additive and interaction models

```{r}
#AIC
interaction_test_models <- list(log_biomass_log_chla, chla_country_additive, chla_country_interaction)
aictab(interaction_test_models) 

# Classical F-test
anova(log_biomass_log_chla, chla_country_additive, chla_country_interaction)
```

The three models represent increasing structural complexity: a global log–log relationship only, an additive model including country-level differences in mean biomass, and a full interaction model allowing country-specific slopes.

The AICc comparison clearly favours the interaction model. It has the lowest AICc (38.91) and carries 83% of the model weight. The additive model is moderately supported (ΔAICc = 3.45, weight = 0.15), while the simple global log–log model has very little support (ΔAICc = 6.88, weight = 0.03). A ΔAICc greater than 6 indicates substantially less support, so the global model is effectively ruled out.

The nested F-tests lead to the same conclusion. Adding country to the global log–log model significantly improves fit (p = 0.008), indicating that mean biomass differs among countries. Adding the interaction further improves fit (p = 0.014), demonstrating that slopes differ among countries. Thus, both country-level differences in mean biomass and heterogeneity in scaling exponents are statistically supported.

Taken together, both information-theoretic (AICc) and classical hypothesis testing approaches converge on the same conclusion: the interaction model provides the best representation of the data. There is no single biomass–chlorophyll relationship across the region; instead, both intercepts and slopes vary among countries.

### Try a GLM

Given that biomass is positive and right-skewed, we additionally fitted generalized linear models using a Gamma error distribution with a log link, which models multiplicative effects directly on the original biomass scale and accommodates mean–variance relationships typical of ecological biomass data.

First we begin with a global and an additive model:

```{r}

# Global 
glm_gamma_global <- glm(
  mean_biomass ~ log_chla_centred,
  family = Gamma(link = "log"),
  data = chla_model_data
)

summary(glm_gamma_global)

# Additive
glm_gamma_additive <- glm(
  mean_biomass ~ log_chla_centred + country,
  family = Gamma(link = "log"),
  data = chla_model_data
)

summary(glm_gamma_additive)
```

Note that the Gamma GLM results closely mirror your Gaussian log–log results in terms of structure.

In this additive Gamma model, the chlorophyll effect is negative but not statistically significant (p = 0.14), and country1 and country2 differ significantly from the regional mean. That is broadly consistent with your additive log–log model.

Now let's check residuals:

```{r}
plot(glm_gamma_additive)
```

...

Then we try the interaction, multiplicative model:

```{r}

chla_model_data$country <- droplevels(chla_model_data$country)
glm_gamma_interaction <- glm(
  mean_biomass ~ log_chla_centred * country,
  family = Gamma(link = "log"),
  data = chla_model_data
)

summary(glm_gamma_interaction)
```

In this interaction Gamma model, as with the lm model, the interaction term for country4 is strongly negative and significant (p = 0.004), and country3 shows marginal evidence of slope deviation. This is almost identical to what you saw in the Gaussian log–log interaction model. So the mean-structure conclusions are robust to distributional assumptions.

```{r}
plot(glm_gamma_interaction)
```

```{r}
glm_test_models <- list(glm_gamma_global, glm_gamma_additive, glm_gamma_interaction)
aictab(glm_test_models)
```
The interaction Gamma model ... hwih emans ....

# Total Biomass vs DHW

```{r}

# Calculate mean dhw per site
dhw_2009_sites_categories <- dhw_2009 %>% 
  group_by(site) %>% 
  summarise(
    mean_max_dhw = mean(max_dhw),
    .groups = "drop"
  ) %>%
  semi_join(biomass_per_site_laura, by = "site") 

# Calculate mean dhw per site
dhw_2009_sites_categories <- dhw_2009 %>% 
  group_by(site) %>% 
  summarise(
    mean_max_dhw = mean(max_dhw),
    .groups = "drop"
  ) %>%
  semi_join(biomass_per_site_laura, by = "site") %>% 
  mutate(
    dhw_category = case_when(
      mean_max_dhw == 0 ~ "No Thermal Stress",
      mean_max_dhw > 0 & mean_max_dhw < 4 ~ "Low Thermal Stress",
      mean_max_dhw >= 4 & mean_max_dhw < 8 ~ "High Thermal Stress",
      mean_max_dhw >= 8 ~ "Severe Thermal Stress"
    ),
    
    # Set the plotting order
    dhw_category = factor(
      dhw_category,
      levels = c(
        "No Thermal Stress",
        "Low Thermal Stress",
        "High Thermal Stress",
        "Severe Thermal Stress"
      )
    )
  )

```

## DHW
```{r}

# Overview of the data
head(dhw_2009_sites_categories,3) # see first three rows
str(dhw_2009_sites_categories) #structure of data
summary(dhw_2009_sites_categories) #summary of data

```

Outputs: - 62 sites - Right-skewed distribution: mean \> median and upper tail extends to 3.24 (which is then \~4x median). - The boxcox suggests log-transformation

## Exploratory Plot

```{r}
#| label: biomass-vs-dhw-plot

# Histogram 
( dhw_boxplot <- merge(
  x = biomass_per_site_laura,
  y = dhw_2009_sites_categories,
  by = "site",
  all = TRUE
) %>% 
  ggplot(aes(x = dhw_category, y = mean_biomass)) +
  geom_boxplot() +
  geom_jitter(width = 0.1, alpha = 0.5) +
  labs(
    x = "Thermal Stress Category",
    y = "Mean Fish Biomass"
  ) +
  theme_bw() )


# Correlation
with(merge(x = biomass_per_site_laura, y = dhw_2009_sites_categories, by = "site", all = TRUE),
    cor(mean_max_dhw, mean_biomass))

```

Chlorophyll a and biomass seem to be weakly positively related, with a correlation of 0.16, but the are large confidence intervals, and a lot of uncertainty in the data.

## Simple ls model

Note that chlorophyll-a, as a continuous predictors, was mean-centred prior to analysis to improve interpretability of model parameters. Centering ensures that the intercept represents the expected response at the average value of the predictor, rather than at zero, which would be biologically unrealistic.

```{r}
#| label: biomass-vs-dhw-lm

# Merge needed data
dhw_model_data <- merge(x = biomass_per_site_laura, y = dhw_2009_sites_categories, by = "site", all = TRUE)

# Centre dhw
dhw_model_data$mean_dhw_centred <- as.numeric(scale(
  dhw_model_data$mean_max_dhw,
  center = TRUE,
  scale = FALSE
))

# Build lm 
lm_biomass_dhw <- lm(mean_biomass~mean_dhw_centred,data = dhw_model_data) 
```

Analyse model:

```{r}
# Analyse the model 
coef(lm_biomass_dhw)
#residuals(lm_biomass_dhw)
summary(lm_biomass_dhw) 
# or arm::display(lm_biomass_dhw)

confint(lm_biomass_dhw) 

```

Outputs:

1.  Intercept: Because chlorophyll was centred, the intercept (21,397) represents the expected biomass at average chlorophyll concentration. The 95% CI (16,982 to 25,811) gives the plausible range for that mean biomass level.

2.  Slopes The slope estimate (3,934) represents the expected increase in biomass for each one-unit increase in chlorophyll above the mean. However, this effect is not statistically significant (p = 0.206), and the confidence interval includes zero (−2,218 to 10,085). This means the data do not provide evidence of a linear relationship between chlorophyll and biomass on the raw scale.

3.  Correlation: 0.162924 (weakly positive)

4.  Effect size The model explains very little variation (R² ≈ 0.027), indicating that chlorophyll alone accounts for only about 3% of the variability in biomass.

Overall, there is no strong support for a simple linear biomass–chlorophyll relationship in this specification.

### Residual checking of simple ls model

Checking model assumptions and violations:

```{r}
par(mfrow= c(2,2)) # divide the page into 2 x 2 frame
plot( lm_biomass_dhw, which= c(1:4))

```

Diagnostic plots: - Slight downward trend in residual smooth line - Increasing spread at higher fitted values - Extreme positive outliers (obs 16, 21, 26)

This suggests possible non-linearity (the curve is not flat) and heteroscedasticity (variance increases as fitted values increase) - i.e. residuals are not normally distributed

## Tranformations

To improve model fit, we can transform the raw data to better conform to the regression assumptions, thereby amplifying predictive power and increasing the overall quality of the model.

We also check how the distribution of the data changes after each transformation to ensure that everything is working as it should.

```{r}

# Transforming variables
dhw_model_data$log_biomass <- log10(dhw_model_data$mean_biomass)
dhw_model_data$log_dhw <- log10(dhw_model_data$mean_annual_chla)

# Check transformations 
( log_biomass_hist <- ggplot(chla_model_data, aes(x = log_biomass)) +
  geom_histogram(bins = 20) +
  labs(
    x = expression("Log-Transformed Biomass"),
    y = "Frequency"
  ) +
  theme_bw() )

( log_chla_hist <- ggplot(chla_model_data, aes(x = log_chla)) +
  geom_histogram(bins = 20) +
  labs(
    x = expression("Log-Transformed Chlorophyll-a"),
    y = "Frequency"
  ) +
  theme_bw() )
```

The data is now looking much more normal!! Great :)

We can now fit models with the transformed data. Ensure you are also centering the log_chla!

```{r}

# Centre logged chla
chla_model_data$log_chla_centred <- as.numeric(
  scale(chla_model_data$log_chla, center = TRUE, scale = FALSE)
)


# Transformed models 
log_biomass_chla <- lm(log_biomass~mean_chla_centred,data = chla_model_data) 
biomass_log_chla <- lm(mean_biomass~log_chla_centred,data = chla_model_data) 
log_biomass_log_chla <- lm(log_biomass~log_chla_centred,data = chla_model_data) 

# Summaries of transformed models
summary(log_biomass_chla) 
summary(biomass_log_chla) 
summary(log_biomass_log_chla) 

# or arm::display(lm_biomass_chla)
#
```

The pattern is very consistent: none of the simple bivariate models show evidence of a relationship between chlorophyll and biomass.

First model: log_biomass \~ mean_annual_chla

The slope is −0.016 (p = 0.79), with R² ≈ 0.001. This indicates essentially no relationship between raw chlorophyll and log-transformed biomass. Chlorophyll explains virtually none of the variation in log biomass.

Second model: mean_biomass \~ log_chla_centred

The slope is 6,048 (p = 0.34), R² ≈ 0.015. This suggests that a one-unit increase in log10 chlorophyll (i.e., a tenfold increase in chlorophyll) would be associated with an increase of about 6,000 biomass units, but this effect is not statistically supported (the confidence interval spans zero).

Because log_chla is centred, the intercept (21,397) represents mean biomass at average log chlorophyll. That makes the intercept interpretable, but it does not change inference.

Third model: log_biomass \~ log_chla_centred (log–log model)

The slope is −0.007 (p = 0.95), R² ≈ 0.00006. This is effectively zero. In scaling terms, the estimated exponent is essentially zero, meaning biomass does not scale with chlorophyll in a simple global power-law relationship.

### Residual checking of transformations

```{r}

# biomass_log_chla
par(mfrow= c(2,2)) # divide the page into 2 x 2 frame
plot(biomass_log_chla, which= c(1:4))

# log_biomass_chla
par(mfrow= c(2,2)) # divide the page into 2 x 2 frame
plot(log_biomass_chla, which= c(1:4))

# log_biomass_log_chla
par(mfrow= c(2,2)) # divide the page into 2 x 2 frame
plot(log_biomass_log_chla, which= c(1:4))
```

Even though neither model shows a significant slope, the log–log form better satisfies linear model assumptions slightly, as it reduces The residuals look the best for the log_log model, with the assumptions now satisfied -\> The log-transformed predictor slightly improves linearity.

### Check AICs

Note that you cannot compare AIC across different response scales, so, given that log-biomass better fits the residuals, we are interested to know which model - log_biomass_chla or log-log - provides the best fit

```{r}
models <- list(log_biomass_chla, log_biomass_log_chla)
aictab(models)
```

When we compare these models we see that they are essentially identical:

ΔAICc = 0.00/0.07 → essentially identical. AIC weights ≈ 0.5 / 0.5 → no meaningful support difference. Log-likelihoods almost identical.

Statistically: The data cannot distinguish between these two models. R²: 0.0012 vs 0.00006s Both slopes non-significant (≈ 0)

So, model comparison based on AICc indicated no meaningful difference between the log–linear and log–log formulations (ΔAICc = 0.07), and both models explained negligible variation in biomass (R² ≈ 0).

However, diagnostic plots suggested slightly improved homoscedasticity and linearity under the log–log specification. Given the equivalent statistical support and the expectation of power-law scaling in ecological productivity relationships, the log–log model was retained for subsequent analyses.

\< Competing models were indistinguishable based on AICc (ΔAICc \< 0.1); however, the log–log specification exhibited improved residual structure and provides a biologically interpretable scaling relationship. We therefore retained the log-transformed predictor. \>

## Taking into account eco-region (country)

### Fixed-effects model (separate)

We will first separate this into each different eco-region (country for now)

```{r}
# plot of biomass vs chlorophyll for a 'no pooling analysis'
( separate_lm_plot <- ggplot(chla_model_data, aes(x = log_chla_centred,
           y = log_biomass)) + geom_point() + geom_smooth(method = "lm", formula = y ~ x) +
  facet_grid(. ~ country) + theme_bw() + labs( x = "Chlorophyll a",
    y = "Log-Biomass", title = "No pooling analysis") )

```

Great differences across countries!!

This analysis treats the relationship at each site as unique and independent of the others.

### Combined fixed-effects model

At the other extreme, we could simply pool all the data together in a single simple regression analysis: (same as above).

```{r}
# Graph the data - X-Y scatter plot
combined_lm_plot <- chla_model_data %>% 
  ggplot(aes(x=log_chla_centred, y=log_biomass)) + 
  geom_point() +
  geom_smooth(method=lm)

combined_lm_plot
```

This model helps us to understand the analysis we want to build from these initial simple extremes.

### Country as a fixed effect

Note: - from above, we will carry on with the log_log model - we centre chla. In models including interactions, centering also makes main effects interpretable as effects evaluated at the mean of the interacting variable and can improve numerical stability in mixed-effects models. - We used sum-to-zero contrasts (effect coding) for country so that parameter estimates represent deviations from the overall mean rather than differences relative to an arbitrarily chosen reference country. Under effect coding, the intercept corresponds to the grand mean of the response (averaged across all observations), and each country coefficient represents that country’s deviation from this grand mean. In other words, coefficients quantify how much higher or lower a given country’s mean is compared to the overall mean.

#### Additive model

First we add it in an additive simple least squares regression. This additive model assumes that the biomass–chlorophyll slope is the same across all countries, but allows countries to differ in their mean biomass levels:

```{r}
# Set sum-to-zero contrasts for country (the grand mean is the baseline - not an arbitrary country level)
contrasts(chla_model_data$country) <- contr.sum(length(levels(chla_model_data$country)))


chla_country_additive <- lm(log_biomass ~ log_chla_centred + country, data = chla_model_data)
summary(chla_country_additive) 
```

Output interpretation:

1.  Intercept Because log chlorophyll is centred, the intercept (4.208) represents the grand mean log biomass at average chlorophyll concentration. Under effect coding, each country coefficient represents that country’s deviation from this overall mean (evaluated at mean chlorophyll), not from a specific reference country.

2.  Slope The slope for log chlorophyll is estiamted to be −0.364 (p = 0.019). This indicates that, when a common slope is imposed across countries, biomass declines with increasing chlorophyll on the log–log scale. In scaling terms, this corresponds to a negative exponent: at higher chlorophyll, biomass is lower, on average. However, this interpretation must be treated cautiously because earlier analyses showed evidence of slope heterogeneity across countries. If slopes differ, this “average” negative slope may reflect opposing patterns cancelling out. So beware with this interpretation!

Country1 has significantly lower biomass than the grand mean (−0.418), while country2 has significantly higher biomass (+0.323), both evaluated at mean chlorophyll. Countries 3 and 4 do not differ significantly from the overall mean.

The model explains about 19% of the variance (R² = 0.19), and the overall model is statistically significant (p = 0.033), indicating that adding country improves fit relative to a simple log–log model without country.

In summary, this additive model suggests that biomass differs among countries and shows a negative average biomass–chlorophyll relationship.

However, given your earlier interaction results, the assumption of a common slope across countries is likely too restrictive, and the interaction model is probably the more appropriate representation of the data.

#### Interaction model

This interaction model allows the biomass–chlorophyll slope to differ among countries rather than forcing a common slope.

```{r}

# Log-log interaction
chla_country_interaction <- lm(log_biomass ~ log_chla_centred * country,
  data = chla_model_data)
summary(chla_country_interaction)

anova(chla_country_interaction)

```

Interpration of outputs:

1.  Intercept Because log chlorophyll is centred and effect coding is used, the intercept (4.276) represents the grand mean log biomass at average chlorophyll across all countries.

2.  Slopes The main effect of log chlorophyll (0.058, p = 0.81) represents the average slope across countries, but in the presence of an interaction this average is not biologically meaningful on its own. The country main effects now represent deviations from the grand mean biomass at mean chlorophyll, conditional on the interaction being included.

3.  Interaction coefficients These are the key terms of the model. These quantify how each country’s slope differs from the overall (grand) slope. Most interaction terms are not statistically different from zero, but the interaction for country4 is strongly negative (−0.899, p = 0.004), indicating that the biomass–chlorophyll scaling relationship in that country differs substantially from the overall trend. Country3 also shows a positive deviation from the mean slope (0.757), with marginal evidence (p = 0.078).

Importantly, the overall model fit improves considerably relative to the additive model. The R² increases from 0.19 to 0.36, and the overall F-test is highly significant (p = 0.003). This indicates that allowing slopes to vary by country explains substantially more variation in biomass.

In substantive terms, there is no single universal biomass–chlorophyll scaling relationship across the region. Instead, the strength and direction of the relationship depend on country, with at least one country exhibiting a markedly different (negative) scaling exponent. This supports the conclusion that the biomass–productivity relationship is spatially heterogeneous rather than globally consistent.

ANOVa - ancova...

#### Test global, additive and interaction models

```{r}
#AIC
interaction_test_models <- list(log_biomass_log_chla, chla_country_additive, chla_country_interaction)
aictab(interaction_test_models) 

# Classical F-test
anova(log_biomass_log_chla, chla_country_additive, chla_country_interaction)
```

The three models represent increasing structural complexity: a global log–log relationship only, an additive model including country-level differences in mean biomass, and a full interaction model allowing country-specific slopes.

The AICc comparison clearly favours the interaction model. It has the lowest AICc (38.91) and carries 83% of the model weight. The additive model is moderately supported (ΔAICc = 3.45, weight = 0.15), while the simple global log–log model has very little support (ΔAICc = 6.88, weight = 0.03). A ΔAICc greater than 6 indicates substantially less support, so the global model is effectively ruled out.

The nested F-tests lead to the same conclusion. Adding country to the global log–log model significantly improves fit (p = 0.008), indicating that mean biomass differs among countries. Adding the interaction further improves fit (p = 0.014), demonstrating that slopes differ among countries. Thus, both country-level differences in mean biomass and heterogeneity in scaling exponents are statistically supported.

Taken together, both information-theoretic (AICc) and classical hypothesis testing approaches converge on the same conclusion: the interaction model provides the best representation of the data. There is no single biomass–chlorophyll relationship across the region; instead, both intercepts and slopes vary among countries.

### Try a GLM

Given that biomass is positive and right-skewed, we additionally fitted generalized linear models using a Gamma error distribution with a log link, which models multiplicative effects directly on the original biomass scale and accommodates mean–variance relationships typical of ecological biomass data.

First we begin with a global and an additive model:

```{r}

# Global 
glm_gamma_global <- glm(
  mean_biomass ~ log_chla_centred,
  family = Gamma(link = "log"),
  data = chla_model_data
)

summary(glm_gamma_global)

# Additive
glm_gamma_additive <- glm(
  mean_biomass ~ log_chla_centred + country,
  family = Gamma(link = "log"),
  data = chla_model_data
)

summary(glm_gamma_additive)
```

Note that the Gamma GLM results closely mirror your Gaussian log–log results in terms of structure.

In this additive Gamma model, the chlorophyll effect is negative but not statistically significant (p = 0.14), and country1 and country2 differ significantly from the regional mean. That is broadly consistent with your additive log–log model.

Now let's check residuals:

```{r}
plot(glm_gamma_additive)
```

...

Then we try the interaction, multiplicative model:

```{r}

chla_model_data$country <- droplevels(chla_model_data$country)
glm_gamma_interaction <- glm(
  mean_biomass ~ log_chla_centred * country,
  family = Gamma(link = "log"),
  data = chla_model_data
)

summary(glm_gamma_interaction)
```

In this interaction Gamma model, as with the lm model, the interaction term for country4 is strongly negative and significant (p = 0.004), and country3 shows marginal evidence of slope deviation. This is almost identical to what you saw in the Gaussian log–log interaction model. So the mean-structure conclusions are robust to distributional assumptions.

```{r}
plot(glm_gamma_interaction)
```

```{r}
glm_test_models <- list(glm_gamma_global, glm_gamma_additive, glm_gamma_interaction)
aictab(glm_test_models)
```
The interaction Gamma model ... hwih emans ....


