---
title: "cleaning_data"
author: "Candela Ferrer Diez"
date: 2025-11-06
format: html
editor: visual
---

## Packages

Load the necessary packages (these can be found in ...)

Use the renv::restore() to load up the necessary package versions along with the necessary dependencies.

```{r}
#| echo: false
#| include: false
renv::init() #renv file for reproducibility 

# Install Packages -> BAD CODE (delete once renv snapshot)
install.packages("lubridate")
install.packages("dplyr") 
install.packages("here") 
install.packages("tidyverse")
install.packages("janitor")
install.packages("ggplot2")

#Load packages -> BAD CODE (delete once renv snapshot)
library(tidyverse)  # Data manipulation (dplyr, ggplot2), pipes (%>%), string functions (str_trim, str_to_title)
library(here)       # Reproducible file paths that work across different computers
library(janitor)    # Clean column names to lowercase snake_case
library(lubridate)  # Parse and manipulate dates easily
library(dplyr)      # ...

#renv::snapshot() - run when all the packages are loaded 


```

*Important note: Comprehensive instructions, including all prerequisites and detailed guidelines for running this analysis and relevant code, are available in the README file of my [GitHub repository](...) for this project. Please refer to the README ... to ensure proper setup and usage.*

```{r}
#| label: renv-restore
renv::restore()
```

## Load data 

### 2009-2016 data

This is ...

```{r}
#| label: load 2009-2016 data
#| echo: false

# Load 2009-2016 data 
fish_abund_2009_raw <- read.csv(here("raw_data","cordio_2009_2016_fish_abund.csv"))
master_location_2009_raw <- read.csv(here("raw_data","cordio_2009_2016_master_location.csv"))
dive_details_2009_raw <- read.csv(here("raw_data","cordio_2009_2016_dive_details.csv"))

```

### 2018-2023 data
This is ...

```{r}
#| label: load 2018-2023 data
#| echo: false

# Load 2018-2023 data 
fish_abund_2018_raw <- read.csv(here("raw_data","cordio_2018_2023_fish_abund.csv"))
master_location_2018_raw <- read.csv(here("raw_data","cordio_2018_2023_master_location.csv"))
dive_details_2018_raw <- read.csv(here("raw_data","cordio_2018_2023_dive_details.csv"))

```

## Explore data structure

...

```{r}
#| label: exploring data structure
#| warning: false
#| message: false

# Source structure exploration function
source(here::here("functions", "cleaning_&_processing.R"))

# Explore 2009-2016 datasets
explore_data_structure(fish_abund_2009_raw, "FISH ABUNDANCE 2009-2016")
explore_data_structure(master_location_2009_raw, "LOCATION 2009-2016")
explore_data_structure(dive_details_2009_raw, "DIVE DETAILS 2009-2016")

# Explore 2018-2023 datasets
explore_data_structure(fish_abund_2018_raw, "FISH ABUNDANCE 2018-2023")
explore_data_structure(master_location_2018_raw, "LOCATION 2018-2023")
explore_data_structure(dive_details_2018_raw, "DIVE DETAILS 2018-2023")

```

## Cleaning data

### Clean fish abundance data
...

```{r}
#| label: clean location data
#| warning: false
#| message: false

# Source cleaning functions
source(here::here("functions", "cleaning_&_processing.R"))

# 2009-2016
clean_fish_2009 <- fish_abund_2009_raw %>%
  clean_column_names() %>%
  drop_columns(drop_cols = c("species_old", "notes", "note_on_analysis", "x")) %>%
  # standardise the textual columns
  standardize_text(c("site", "station", "country", "location", "family", "species_new", "taxon_group", "functional_group")) %>% 
  convert_column_types(
    date_cols = intersect(c("date"), names(.)),
    #columns that should be integers
    integer_cols = intersect(
      c("ts_no", "number"),
      names(.)
    ),
    # columns that should be numeric
    numeric_cols = intersect(
      c("size_class", "l", "a", "b", "weight_g", "tot_wt_g"),
      names(.)
    ),
    # columns that should be factors (categorical)
    factor_cols = intersect(c("country", "location", "site", "station", "family", "taxon_group", "functional_group"), names(.))
  ) 

# 2018-2023
clean_fish_2018 <- fish_abund_2018_raw %>%
  clean_column_names() %>%
  drop_columns(drop_cols = c("x", "x_1", "x_2","x_3")) %>%
  # standardise only the textual ID-like columns
  standardize_text(c("site", "station", "country", "family", "species", "trophic_groups")) %>%
  convert_column_types(
    date_cols = intersect(c("date"), names(.)),
    
    # columns that should be integers
    integer_cols = intersect(
      c("transect_no", "number"), 
      names(.)
    ),
    
   # columns that should be numeric
    numeric_cols = intersect(
      c("size_class", "l", "a", "b", "weight_g", "total_wt_g"), 
      names(.)
    ),

   # columns that should be factors (categorical)
    factor_cols = intersect(
      c("station", "site", "country", "family", "species", "trophic_groups"), 
      names(.)
    )
  )

```

### Clean location data
...

```{r}
#| label: clean location data
#| warning: false
#| message: false

# Source cleaning functions
source(here::here("functions", "cleaning_&_processing.R"))

# 2009-2016
clean_location_2009 <- master_location_2009_raw %>%
  clean_column_names() %>%
  drop_columns(drop_cols = c("reeftype_old",        
                "extra_notes_for_missing_data"  )) %>%
                 
  # standardise textual columns
  standardize_text(c("station", "site", "location",
      "country", "orientation", "geology",
      "reef_geomorphology", "reef_type")) %>% 
  convert_column_types(
    date_cols = intersect(c("date"), names(.)),
    
    #columns that should be integers
    integer_cols = intersect(
      c("exposure"), names(.)),
    
    # columns that should be numeric
    numeric_cols = intersect(
      c("latitude", "longitude"), names(.)),
    
    # columns that should be factors (categorical)
    factor_cols = intersect(c("station", "site", 
      "location", "country", "orientation", "geology",
      "reef_geomorphology", "reef_type"), names(.))) %>%
    
  # validate coordinates
  { if(all(c("latitude","longitude") %in% names(.)))
    fix_coordinates(., lat_col = "latitude", lon_col =
                      "longitude") else . }

# 2018-2023
clean_location_2018 <- master_location_2018_raw %>%
  clean_column_names() %>%
  drop_columns(drop_cols = 
                 c("notes")) %>%
                 
  # standardise only the textual ID-like columns
  standardize_text(c("station", "site", "location",
      "country", "orientation", "geology",
      "reef_geomorphology", "reef_type")) %>% 
  convert_column_types(
    date_cols = intersect(c("date"), names(.)),
    
    #columns that should be integers
    integer_cols = intersect(
      c("exposure", "slope"), names(.)),
    
    # columns that should be numeric
    numeric_cols = intersect(
      c("latitude", "longitude"), names(.)),
    
    # columns that should be factors (categorical)
    factor_cols = intersect(c("station", "site", 
      "location", "country", "orientation", "geology",
      "reef_geomorphology", "reef_type"), names(.))) %>%
    
  # validate coordinates
  { if(all(c("latitude","longitude") %in% names(.)))
    fix_coordinates(., lat_col = "latitude", lon_col =
                      "longitude") else . }
```

### Clean dive data
...

```{r}
#| label: clean dive data
#| warning: false
#| message: false

# Source cleaning function
source(here::here("functions", "cleaning_&_processing.R"))

# 2009-2016
clean_dive_details_2009 <- dive_details_2009_raw %>%
  clean_column_names() %>%
  drop_columns(drop_cols = c("dive_time", "other_notes",
                "other_notes_1")) %>%
  # standardise only the textual ID-like columns
  standardize_text(c("station", "site", 
        "depth_categories", "assoc_habitats")) %>% 
  convert_column_types(
    date_cols = intersect(c("date"), names(.)),
    
    #columns that should be integers
    integer_cols = NULL,
    
    # columns that should be numeric
    numeric_cols = intersect(
      c("max_depth_m", "min_depth_m", "rugosity"),
      names(.)),
    
    # columns that should be factors (categorical)
    factor_cols = intersect(c("station", "site",
      "depth_categories", "assoc_habitats"), names(.))) 

# 2018-2023
clean_dive_details_2018 <- dive_details_2018_raw %>%
  clean_column_names() %>%
  drop_columns(drop_cols = c("total_dive_time_minutes")) %>%
  # standardise only the textual ID-like columns
  standardize_text(c("station", "site", "location", "country")) %>% 
  convert_column_types(
    date_cols = intersect(c("date"), names(.)),
    
    #columns that should be integers
    integer_cols = NULL,
    
    # columns that should be numeric
    numeric_cols = intersect(
      c("latitude", "longitude", "rugosity", "viz", 
        "temp", "max_depth_m", "min_depth_m",
        "div_maxd"),names(.)),
    
    # columns that should be factors (categorical)
    factor_cols = intersect(c("station", "site", 
        "location", "country"), names(.))) %>% 
  { if(all(c("latitude","longitude") %in% names(.)))
       fix_coordinates(., lat_col = "latitude", lon_col =
                       "longitude") else . }

```

## Verify cleaning
...
```{r}
#| label: verify cleaning process
#| warning: false
#| message: false

# Source cleaning function
source(here::here("functions", "cleaning_&_processing.R"))

# Explore cleaned 2009-2016 data
explore_data_structure(clean_fish_2009, "CLEANED FISH ABUNDANCE 2009-2016")
explore_data_structure(clean_location_2009, "CLEANED LOCATION 2009-2016")
explore_data_structure(clean_dive_details_2009, "CLEANED DIVE DETAILS 2009-2016")

# Explore cleaned 2018-2023 datasets data
explore_data_structure(clean_fish_2018, "CLEANED FISH ABUNDANCE 2018-2023")
explore_data_structure(clean_location_2018, "CLEANED LOCATION 2018-2023")
explore_data_structure(clean_dive_details_2018, "CLEANED DIVE DETAILS 2018-2023")

```

## Save clean data 
...

```{r}
#| label: save cleaned data
#| warning: false
#| message: false

# 2009-2016
write_csv(clean_fish_2009, here("processed_data", "clean_fish_2009.csv"))
write_csv(clean_location_2009, here("processed_data", "clean_location_2009.csv"))
write_csv(clean_dive_details_2009, here("processed_data", "clean_dive_details_2009.csv"))

# 2018-2023
write_csv(clean_fish_2018, here("processed_data", "clean_fish_2018.csv"))
write_csv(clean_location_2018, here("processed_data", "clean_location_2018.csv"))
write_csv(clean_dive_details_2018, here("processed_data", "clean_dive_details_2018.csv"))

```




