---
title: "exploring_script"
author: "Candela Ferrer Diez"
date: 2025-11-06
format: 
  html:
    toc: true
    toc-depth: 2
    code-fold: false
    theme: cosmo
    embed-resources: true
editor: visual
execute:
  warning: false
  message: false
---

## Overview

This document explores ecological fish survey data from Kenya across two time periods: 2009-2016 and 2018-2023.

## R Set-up

*Important note: Comprehensive instructions, including all prerequisites and detailed guidelines for running this analysis and relevant code, are available in the README file of my [GitHub repository](...) for this project. Please refer to the README ... to ensure proper setup and usage.*

Load the necessary packages (these can be found in ...)

*Important note: Comprehensive instructions, including all prerequisites and detailed guidelines for running this analysis and relevant code, are available in the README file of my [GitHub repository](...) for this project. Please refer to the README ... to ensure proper setup and usage.*

Use the renv::restore() to load up the necessary package versions along with the necessary dependencies.

```{r}
#| label: r set-up
#| echo: false
#| include: false

#Load packages -> BAD CODE (delete once renv snapshot)
library(tidyverse)
library(sf)
library(rnaturalearth)
library(rnaturalearthdata)

```

```{r}
#| label: renv-restore
renv::restore()
```

## 2009-2016 data

### Load data

```{r}
#| label: load cleaned data 2009
#| echo: false
#| warning: false
#| message: false

# Load 2009-2016 datasets
clean_fish_2009 <- read.csv(here("processed_data","clean_fish_2009.csv"))
clean_location_2009 <- read.csv(here("processed_data","clean_location_2009.csv"))
clean_dive_details_2009 <- read.csv(here("processed_data","clean_dive_details_2009.csv"))

```

### Site Observations Map
...

```{r}
#| label: map 2009 site observations
#| echo: false
#| warning: false
#| message: false

# Filter for Kenya only and count observations per station
site_counts_2009 <- clean_fish_2009 %>%
  filter(country == "kenya") %>%
  group_by(site) %>%
  summarise(n_observations = n()) %>%
  ungroup()

# Join with location data (Kenya only)
site_map_data_2009 <- clean_location_2009 %>%
  filter(country == "kenya") %>%
  left_join(site_counts_2009, by = "site") %>%
  filter(!is.na(latitude) & !is.na(longitude))

# Get map data for Kenya and surrounding region
world <- ne_countries(scale = "medium", returnclass = "sf")
kenya_region <- world %>%
  filter(admin %in% c("Kenya", "Tanzania", "Somalia", "Uganda"))

# Create the map with a tighter zoom
map_plot_2009 <- ggplot() +
  
  # Plot all countries in Africa as the base layer
  geom_sf(data = world[world$continent == "Africa",], 
          fill = "gray95", color = "gray50", size = 0.3) + 
  
  geom_sf(data = kenya_region, fill = "lightblue", color = "gray30", size = 0.3) +
  
  geom_point(data = site_map_data_2009, 
             aes(x = longitude, y = latitude, size = n_observations),
             alpha = 0.6, color = "#440154") +
  
  scale_size_continuous(name = "Number of\nObservations", 
                        range = c(0.5, 5),
                        breaks = c(10, 25, 50, 75, 100, 150, 200)) +
  
  # Specify region of focus - Kenya in frame
  coord_sf(xlim = c(38.9, 41.8), ylim = c(-5, -1.5)) + 
  
  labs(title = "Marine Survey Sites in Kenya",
       subtitle = "Point size represents number of fish observations per site",
       x = "Longitude", y = "Latitude") +
  
  theme_minimal() +
  theme(panel.grid.major = element_line(color = "gray90", size = 0.2),
        panel.background = element_rect(fill = "aliceblue"),
        plot.title = element_text(face = "bold", size = 14),
        legend.position = "right")

print(map_plot_2009)

```

### Species observations per site in 2009

```{r}
#| label: species observations per site 2009
#| echo: false
#| warning: false
#| message: false


# Calculate species richness and total observations per site (Kenya only)
species_per_site_2009 <- clean_fish_2009 %>%
  filter(country == "kenya") %>%
  group_by(site) %>%
  summarise(
    n_species = n_distinct(species_new),
    n_observations = n(),
    .groups = "drop"
  ) %>%
  arrange(desc(n_species))

# Create bar plot of species richness
species_plot_2009 <- ggplot(species_per_site_2009, 
                       aes(x = reorder(site, n_species), y = n_species)) +
  geom_col(fill = "#440154") +
  geom_text(aes(label = n_species), hjust = -0.2, size = 3) +
  coord_flip() +
  labs(title = "Species Richness by Survey Site (Kenya)",
       subtitle = "Number of unique fish species observed at each site",
       x = "Site", y = "Number of Species") +
  theme_minimal() +
  theme(plot.title = element_text(face = "bold", size = 14),
        axis.text.y = element_text(size = 9),
        legend.position = "bottom")

print(species_plot_2009)

```

### Species observations vs sampling effort 2009

```{r}
#| label: species observations vs sampling effort 2009
#| echo: false
#| warning: false
#| message: false

# Create plot 
sampling_effort_plot_2009 <- ggplot(species_per_site_2009, 
                               aes(x = n_observations, y = n_species, size = n_observations)) +
  geom_point(alpha = 0.7, color = "#440154") +
  geom_text(aes(label = site), hjust = -0.1, vjust = 0.5, 
            size = 2.5, check_overlap = TRUE) +
  scale_size_continuous(guide = "none") +
  labs(title = "Species Observations vs Sampling Effort (Kenya) 2009",
       subtitle = "Relationship between number of observations and species observations",
       x = "Number of Observations", y = "Number of Species") +
  theme_minimal() +
  theme(plot.title = element_text(face = "bold", size = 14),
        legend.position = "bottom")

print(sampling_effort_plot_2009)

```

### Summary statistics 2009

```{r}
#| label: summary statistics 2009
#| echo: false
#| warning: false
#| message: false

cat("\n=== SUMMARY STATISTICS (KENYA ONLY) ===\n")
cat("\nTotal sites:", n_distinct(species_per_site_2009$site))
cat("\nTotal species:", n_distinct(clean_fish_2009 %>% filter(country == "kenya") %>% pull(species_new)))
cat("\nTotal observations:", nrow(clean_fish_2009 %>% filter(country == "kenya")))
cat("\n\nTop 5 sites by species richness:\n")
print(head(species_per_site_2009))

```


## 2018-2023 data

### Load data

```{r}
#| label: load cleaned data 2018
#| echo: false
#| warning: false
#| message: false

##Load 2018-2023 datasets
clean_fish_2018 <- read.csv(here("processed_data","clean_fish_2018.csv"))
clean_location_2018 <- read.csv(here("processed_data","clean_location_2018.csv"))
clean_dive_details_2018 <- read.csv(here("processed_data","clean_dive_details_2018.csv"))

```


### Site Observations Map
...

```{r}
#| label: map 2018 site observations
#| echo: false
#| warning: false
#| message: false

# Filter for Kenya only and count observations per station
site_counts_2018 <- clean_fish_2018 %>%
  filter(country == "kenya") %>%
  group_by(site) %>%
  summarise(n_observations = n()) %>%
  ungroup()

# Join with location data (Kenya only)
site_map_data_2018 <- clean_location_2018 %>%
  filter(country == "kenya") %>%
  left_join(site_counts_2018, by = "site") %>%
  filter(!is.na(latitude) & !is.na(longitude))

# Get map data for Kenya and surrounding region
world <- ne_countries(scale = "medium", returnclass = "sf")
kenya_region <- world %>%
  filter(admin %in% c("Kenya", "Tanzania", "Somalia", "Uganda"))

# Create the map with a tighter zoom
map_plot_2018 <- ggplot() +
  
  # Plot all countries in Africa as the base layer
  geom_sf(data = world[world$continent == "Africa",], 
          fill = "gray95", color = "gray50", size = 0.3) + 
  
  geom_sf(data = kenya_region, fill = "lightblue", color = "gray30", size = 0.3) +
  
  geom_point(data = site_map_data_2018, 
             aes(x = longitude, y = latitude, size = n_observations),
             alpha = 0.6, color = "#440154") +
  
  scale_size_continuous(name = "Number of\nObservations", 
                        range = c(0.5, 5),
                        breaks = c(10, 25, 50, 75, 100, 150, 200)) +
  
  # Specify region of focus - Kenya in frame
  coord_sf(xlim = c(38.9, 41.8), ylim = c(-5, -1.5)) + 
  
  labs(title = "Marine Survey Sites in Kenya",
       subtitle = "Point size represents number of fish observations per site",
       x = "Longitude", y = "Latitude") +
  
  theme_minimal() +
  theme(panel.grid.major = element_line(color = "gray90", size = 0.2),
        panel.background = element_rect(fill = "aliceblue"),
        plot.title = element_text(face = "bold", size = 14),
        legend.position = "right")

print(map_plot_2018)

```

### Species observations per site in 2018

```{r}
#| label: species observations per site 2018
#| echo: false
#| warning: false
#| message: false


# Calculate species richness and total observations per site (Kenya only)
species_per_site_2018 <- clean_fish_2018 %>%
  filter(country == "kenya") %>%
  group_by(site) %>%
  summarise(
    n_species = n_distinct(species),
    n_observations = n(),
    .groups = "drop"
  ) %>%
  arrange(desc(n_species))

# Create bar plot of species richness
species_plot_2018 <- ggplot(species_per_site_2018, 
                       aes(x = reorder(site, n_species), y = n_species)) +
  geom_col(fill = "#440154") +
  geom_text(aes(label = n_species), hjust = -0.2, size = 3) +
  coord_flip() +
  labs(title = "Species Richness by Survey Site (Kenya)",
       subtitle = "Number of unique fish species observed at each site",
       x = "Site", y = "Number of Species") +
  theme_minimal() +
  theme(plot.title = element_text(face = "bold", size = 14),
        axis.text.y = element_text(size = 9),
        legend.position = "bottom")

print(species_plot_2018)

```

### Species observations vs sampling effort 2018

```{r}
#| label: species observations vs sampling effort 2018
#| echo: false
#| warning: false
#| message: false

# # Create plot 
sampling_effort_plot_2018 <- ggplot(species_per_site_2018, 
                               aes(x = n_observations, y = n_species, size = n_observations)) +
  geom_point(alpha = 0.7, color = "#440154") +
  geom_text(aes(label = site), hjust = -0.1, vjust = 0.5, 
            size = 2.5, check_overlap = TRUE) +
  scale_size_continuous(guide = "none") +
  labs(title = "Species Observations vs Sampling Effort (Kenya) 2018",
       subtitle = "Relationship between number of observations and species observations",
       x = "Number of Observations", y = "Number of Species") +
  theme_minimal() +
  theme(plot.title = element_text(face = "bold", size = 14),
        legend.position = "bottom")

print(sampling_effort_plot_2018)

```

### Summary statistics 2018

```{r}
#| label: summary statistics 2018
#| echo: false
#| warning: false
#| message: false

cat("\n=== SUMMARY STATISTICS (KENYA ONLY) ===\n")
cat("\nTotal sites:", n_distinct(species_per_site_2018$site))
cat("\nTotal species:", n_distinct(clean_fish_2018 %>% filter(country == "kenya") %>% pull(species)))
cat("\nTotal observations:", nrow(clean_fish_2018 %>% filter(country == "kenya")))
cat("\n\nTop 5 sites by species richness:\n")
print(head(species_per_site_2018))

```


