---
title: "exploring_script"
author: "Candela Ferrer Diez"
date: 2025-11-06
format: 
  html:
    toc: true
    toc-depth: 2
    code-fold: false
    theme: cosmo
    embed-resources: true
editor: visual
execute:
  warning: false
  message: false
---

## Overview

This document explores ecological fish survey data from Kenya across two time periods: 2009-2016 and 2018-2023.

## R Set-up

*Important note: Comprehensive instructions, including all prerequisites and detailed guidelines for running this analysis and relevant code, are available in the README file of my [GitHub repository](...) for this project. Please refer to the README ... to ensure proper setup and usage.*

Load the necessary packages (these can be found in ...)

*Important note: Comprehensive instructions, including all prerequisites and detailed guidelines for running this analysis and relevant code, are available in the README file of my [GitHub repository](...) for this project. Please refer to the README ... to ensure proper setup and usage.*

Use the renv::restore() to load up the necessary package versions along with the necessary dependencies.

```{r}
#| label: r set-up
#| echo: false
#| include: false

#Load packages -> BAD CODE (delete once renv snapshot)
library(tidyverse)
library(here)
library(maps)
library(knitr)


```

```{r}
#| label: renv-restore
renv::restore()
```

## Load data 

```{r}
#| label: load cleaned data
#| echo: false
#| warning: false
#| message: false

# Load 2009-2016 datasets
fish_2009 <- read_csv(here("processed_data","clean_fish_2009.csv"))
location_2009 <- read_csv(here("processed_data","clean_location_2009.csv"))
dive_2009 <- read_csv(here("processed_data","clean_dive_details_2009.csv"))

# Load 2018-2023 datasets
fish_2018 <- read_csv(here("processed_data","clean_fish_2018.csv"))
location_2018 <- read_csv(here("processed_data","clean_location_2018.csv"))
dive_2018 <- read_csv(here("processed_data","clean_dive_details_2018.csv"))
```

## Process 2009-2016 Kenya Data

```{r process-2009}
# Filter Kenya locations and join with dive details, then fish data
kenya_data_2009 <- location_2009 %>%
  filter(country == "kenya") %>%
  select(date, station, site, location, country, latitude, longitude) %>%
  inner_join(
    dive_2009 %>% select(date, station, max_depth_m, min_depth_m, rugosity),
    by = c("station", "date")
  ) %>%
  inner_join(
    fish_2009 %>% select(date, station, species_new, family, number),
    by = c("station", "date")
  )

# Count unique sites for 2009-2016
sites_2009 <- kenya_data_2009 %>%
  select(date, station, site, location, latitude, longitude) %>%
  distinct() %>%
  mutate(period = "2009-2016")

# Count unique species for 2009-2016
species_2009 <- kenya_data_2009 %>%
  select(species_new) %>%
  distinct() %>%
  mutate(period = "2009-2016")
```

## Process 2018-2023 Kenya Data

```{r process-2018}
# Filter Kenya locations and join with dive details, then fish data
kenya_data_2018 <- location_2018 %>%
  filter(country == "kenya") %>%
  select(date, station, site, location, country, latitude, longitude) %>%
  inner_join(
    dive_2018 %>% select(date, station, site, max_depth_m, min_depth_m, rugosity),
    by = c("station", "date", "site")
  ) %>%
  inner_join(
    fish_2018 %>% select(date, station, site, species, family, number),
    by = c("station", "date", "site")
  )

# Count unique sites for 2018-2023
sites_2018 <- kenya_data_2018 %>%
  select(date, station, site, location, latitude, longitude) %>%
  distinct() %>%
  mutate(period = "2018-2023")

# Count unique species for 2018-2023
species_2018 <- kenya_data_2018 %>%
  select(species) %>%
  distinct() %>%
  mutate(period = "2018-2023")
```

## Summary Statistics

### Site Counts

```{r site-counts}
site_summary <- tibble(
  Period = c("2009-2016", "2018-2023"),
  `Number of Sites` = c(nrow(sites_2009), nrow(sites_2018))
)

knitr::kable(site_summary, caption = "Kenya Survey Sites by Period")
```

**2009-2016 period:** `r nrow(sites_2009)` unique sites  
**2018-2023 period:** `r nrow(sites_2018)` unique sites

### Species Counts

```{r species-counts}
species_summary <- tibble(
  Period = c("2009-2016", "2018-2023"),
  `Number of Species` = c(nrow(species_2009), nrow(species_2018))
)

knitr::kable(species_summary, caption = "Kenya Species Richness by Period")
```

**2009-2016 period:** `r nrow(species_2009)` unique species  
**2018-2023 period:** `r nrow(species_2018)` unique species

### Sites Monitored in Both Periods

```{r common-sites}
common_sites <- intersect(sites_2009$site, sites_2018$site)
```

**Common sites:** `r length(common_sites)` sites monitored in both periods

```{r common-sites-list}
if(length(common_sites) > 0) {
  common_sites_df <- tibble(
    `Site Name` = sort(common_sites)
  )
  knitr::kable(common_sites_df, caption = "Sites Monitored in Both Time Periods")
}
```

## Geographic Coverage

```{r coverage-table}
# Combine sites for analysis
all_kenya_sites <- bind_rows(sites_2009, sites_2018)

# Geographic coverage summary
coverage_summary <- all_kenya_sites %>%
  group_by(period) %>%
  summarise(
    `Total Sites` = n(),
    `Locations` = n_distinct(location),
    `Min Latitude` = round(min(latitude, na.rm = TRUE), 3),
    `Max Latitude` = round(max(latitude, na.rm = TRUE), 3),
    `Min Longitude` = round(min(longitude, na.rm = TRUE), 3),
    `Max Longitude` = round(max(longitude, na.rm = TRUE), 3)
  )

knitr::kable(coverage_summary, caption = "Geographic Coverage by Period")
```

### Sites by Location

```{r location-breakdown}
location_breakdown <- all_kenya_sites %>%
  group_by(period, location) %>%
  summarise(n_sites = n(), .groups = "drop") %>%
  pivot_wider(names_from = period, values_from = n_sites, values_fill = 0)

knitr::kable(location_breakdown, 
      col.names = c("Location", "2009-2016", "2018-2023"),
      caption = "Number of Sites per Location")
```

## Map Visualization

### Kenya Marine Survey Sites

```{r kenya-map, fig.width=10, fig.height=8}
# Get world map data
world <- map_data("world")

# Create map
kenya_map <- ggplot() +
  geom_polygon(data = world, 
               aes(x = long, y = lat, group = group),
               fill = "lightgray", color = "white", linewidth = 0.2) +
  geom_point(data = all_kenya_sites,
             aes(x = longitude, y = latitude, color = period, shape = period),
             size = 3.5, alpha = 0.8) +
  coord_fixed(1.3, xlim = c(38.5, 40.5), ylim = c(-5, -3.5)) +
  scale_color_manual(values = c("2009-2016" = "#E69F00", 
                                 "2018-2023" = "#56B4E9")) +
  scale_shape_manual(values = c("2009-2016" = 16, "2018-2023" = 17)) +
  labs(title = "Kenya Marine Survey Sites",
       subtitle = sprintf("2009-2016: %d sites | 2018-2023: %d sites", 
                         nrow(sites_2009), nrow(sites_2018)),
       x = "Longitude",
       y = "Latitude",
       color = "Period",
       shape = "Period") +
  theme_minimal() +
  theme(plot.title = element_text(face = "bold", size = 16),
        plot.subtitle = element_text(size = 12, color = "gray40"),
        legend.position = "bottom",
        panel.grid.minor = element_blank())

print(kenya_map)
```

## Key Findings

- **Total sites surveyed:** `r nrow(sites_2009)` (2009-2016) and `r nrow(sites_2018)` (2018-2023)
- **Total species observed:** `r nrow(species_2009)` (2009-2016) and `r nrow(species_2018)` (2018-2023)
- **Sites monitored in both periods:** `r length(common_sites)`
- **Geographic range:** Sites span approximately `r round(max(all_kenya_sites$latitude, na.rm=TRUE) - min(all_kenya_sites$latitude, na.rm=TRUE), 2)` degrees latitude along Kenya's coast

---

*Generated: `r Sys.time()`*

```{r}

# Load necessary libraries
library(tidyverse)
library(ggplot2)
library(rnaturalearth) # For fetching map boundaries
library(sf)            # For handling spatial data

# --- 1. Data Aggregation (Using your existing cleaned data frames) ---

# Count the total number of fish observations (rows) per station
station_counts_2018 <- clean_fish_2018 %>%
  group_by(station) %>%
  summarise(data_count = n(), .groups = 'drop')

# Get the unique coordinates for each station
station_locations_2018 <- clean_location_2018 %>%
  select(station, latitude, longitude) %>%
  distinct()

# Merge the data counts with the unique geographic coordinates
df_plot_data <- inner_join(
  station_locations_2018,
  station_counts_2018,
  by = "station"
)

# --- 2. Fetch Map Data and Filter to Kenya Region ---

# Get medium-scale country boundaries data from the rnaturalearth package
world_map <- ne_countries(scale = "medium", returnclass = "sf")

# Define the bounding box for East Africa/Kenya
# Kenya: Lat approx -5 to 5, Lon approx 34 to 42
bbox <- c(
  xmin = 32, # West (covers Uganda, Tanzania)
  ymin = -7, # South (covers Tanzania)
  xmax = 44, # East (covers Somalia)
  ymax = 6   # North (covers Ethiopia)
)


# --- 3. Create the Bubble Map with Geographic Background ---

data_volume_map_kenya <- ggplot() +
  
  # A. Add the country boundaries layer
  geom_sf(
    data = world_map,
    fill = "antiquewhite",
    color = "gray80",
    size = 0.3
  ) +
  
  # B. Add the station points layer (Bubble Map)
  geom_point(
    data = df_plot_data,
    aes(x = longitude, y = latitude, size = data_count, color = data_count),
    alpha = 0.7,
    shape = 21, # Change shape to 21 for easier visual layering
    fill = "steelblue"
  ) +
  
  # C. Set the color and size scales
  scale_color_viridis_c(
    option = "magma",
    name = "Observation Count"
  ) +
  scale_size_continuous(
    range = c(3, 15), # Min and max dot size
    name = "Data Volume"
  ) +
  
  # D. Crop the map to the defined Kenya region and set projection
  coord_sf(
    xlim = c(bbox["xmin"], bbox["xmax"]),
    ylim = c(bbox["ymin"], bbox["ymax"]),
    expand = FALSE # Prevents adding extra space around the bounding box
  ) +
  
  # E. Add labels and theme
  labs(
    title = "Kenya Station Data Volume (2018-2023)",
    subtitle = "Dot size and color are proportional to the number of fish observations",
    x = "Longitude",
    y = "Latitude"
  ) +
  theme_bw() +
  theme(
    panel.grid.major = element_line(color = "gray80", linetype = "dotted"),
    panel.background = element_rect(fill = "aliceblue") # Light blue ocean background
  )

# Display the map
print(data_volume_map_kenya)

```

