---
title: "merging_additional_data"
author: "Candela Ferrer Diez"
date: 2025-11-06
format: 
  html:
    toc: true
    toc-depth: 2
    code-fold: false
    theme: cosmo
    embed-resources: true
editor: visual
execute:
  warning: false
  message: false
---

## R Set-up

Load the necessary packages (these can be found in ...)

*Important note: Comprehensive instructions, including all prerequisites and detailed guidelines for running this analysis and relevant code, are available in the README file of my [GitHub repository](...) for this project. Please refer to the README ... to ensure proper setup and usage.*

Use the renv::restore() to load up the necessary package versions along with the necessary dependencies.


```{r}
#| label: r set-up
#| echo: false
#| include: false

#Load packages -> BAD CODE (delete once renv snapshot)
library(sf)           # For spatial data handling
library(dplyr)        # For data manipulation
library(terra)        # Alternative for raster operations (if needed)
library(here)
library(dplyr)
library(ggplot2)
```

```{r}
#| label: renv-restore
renv::restore()
```

# Adding Human Gravity

## Location data
We will be adding the merged data to the master lcoation - as this contians..

### Load cleaned data
...

```{r}
#| label: load location data
#| echo: false
#| warning: false

##load location data
clean_location_2018 <- read.csv(here::here("processed_data","clean_location_2018.csv"))

```

### Convert to Spatial Format

Transform the location data into a spatial object (simple features) to enable spatial operations and joins with the gravity data.
```{r}
#| label: convert-to-spatial
#| echo: false
#| warning: false


# Convert to sf object with WGS84 coordinate system
sites_sf <- st_as_sf(clean_location_2018, 
                     coords = c("longitude", "latitude"),
                     crs = 4326,  # WGS84
                     remove = FALSE)  # Keep original lon/lat columns

```

## Gravity Data

### Load Gravity Data

Load the human gravity shapefile which contains modeled human pressure on coral reefs based on population density and accessibility within a 500 km radius.
... (more about this)

```{r}
#| label: load-gravity-data
#| echo: false
#| warning: false

# Load the shapefile
gravity_data <- st_read(here::here("raw_data", "Total Gravity of Coral Reefs 1.0.shp"))

# Check the structure
glimpse(gravity_data)

```

### Clean Gravity Data
Standardise column names to match the celaned data
```{r}
#| label: clean-gravity-data
#| echo: false
#| warning: false

# Clean gravity data
clean_gravity_data <- clean_column_names(gravity_data)

```

### Extract Gravity Values

Match each survey site to the nearest gravity grid cell and extract the human pressure values.
I will be doing this through a spatial join by finding the nearest gravity grid cell for each cell - For each site in sites_sf, find the nearest polygon in gravity_data and copy all its attributes (reef_id, grav_tot) to that site's row

```{r}
#| label: extract-gravity
#| echo: false
#| warning: false

# Switch from spherical (S2) to planar geometry calculations
sf_use_s2(FALSE) 

# Spatial join - attach gravity data to each survey site
sites_with_gravity <- st_join(sites_sf, 
              clean_gravity_data,
              join = st_nearest_feature) # Match to nearest grid cell

# Calculate how far each site is from its matched grid cell
# This helps assess match quality (closer = better)
nearest_indices <- st_nearest_feature(sites_sf, clean_gravity_data)

# Calculate the distance between each site and its matched grid cell
distances <- st_distance(sites_sf,
                        clean_gravity_data[nearest_indices, ],  
                        by_element = TRUE)

# Add distance as a new column (to be able to check matches)
sites_with_gravity$distance_to_grid_km <- as.numeric(distances) / 1000

# Convert to dataframe and organize columns
results <- sites_with_gravity %>%
  mutate(distance_to_grid_km = as.numeric(distances) / 1000) %>%
  st_drop_geometry() %>%
  dplyr::select(
    date, station, site, location, country, 
    latitude, longitude,
    contains("grav"),  # All gravity-related columns
    distance_to_grid_km,
    everything()
  )

```

### Checking results

Convert the spatial object back to a regular dataframe and organize columns for easier interpretation.
```{r}
#| label: check-results

# Check for any sites without gravity values
missing_gravity <- results %>%
  filter(is.na(grav_tot))  # Use lowercase after clean_column_names

if(nrow(missing_gravity) > 0) {
  print(paste("⚠ Warning:", nrow(missing_gravity), "sites missing gravity values"))
  print(missing_gravity %>% select(station, site, location))
} else {
  print("All sites successfully matched with gravity values")
}
```

# Adding Management Data







# Save data  
... 

```{r}

# ============================================================
# STEP 7: EXPORT RESULTS (to be mofidied)
# ============================================================

# Export full results
write.csv(results, 
          "clean_location_2018_with_gravity.csv", 
          row.names = FALSE)

# Export summaries
write.csv(country_summary, 
          "gravity_summary_by_country.csv", 
          row.names = FALSE)

write.csv(location_summary, 
          "gravity_summary_by_location.csv", 
          row.rows = FALSE)

# Save as spatial file if needed for GIS
st_write(sites_with_gravity, 
         "sites_with_gravity.shp", 
         delete_dsn = TRUE)  # Overwrite if exists

print("\n✓ All results exported successfully!")
print("Files created:")
print("  - clean_location_2018_with_gravity.csv")
print("  - gravity_summary_by_country.csv")
print("  - gravity_summary_by_location.csv")
print("  - sites_with_gravity.shp (and associated files)")


```
