---
title: "adding_chla"
author: "Candela Ferrer Diez"
date: 2025-12-01
format: html
editor: visual
---

# Overview

This document processes NASA Ocean Color Level-3 binned (L3b) chlorophyll-a data.. and sea-surface temperature data (SST) for the Western Indian Ocean (WIO) region. The script:

-   Reads HDF5/NetCDF files containing binned satellite ocean color data
-   Converts NASA bin numbers to geographic coordinates (lat/lon)
-   Filters data to the WIO Area of Interest (37°E - 47°E, 15°S - 1°N)
-   Exports combined data to CSV and Parquet format

# R Set-up

*Important note: Comprehensive instructions, including all prerequisites and detailed guidelines for running this analysis and relevant code, are available in the README file of my [GitHub repository](...) for this project. Please refer to the README to ensure proper setup and usage.*

## Load required libraries

```{r}
#| label: load-packages
#| echo: false
#| include: false

# Load required libraries
library(here)
library(ncdf4)
library(data.table)
library(arrow)
library(sf)
library(tidyverse) 
library(dplyr)      
library(readr)

```

renv::restore() loads the necessary package versions along with the necessary dependencies. Information about this can be found in the lockfile (\[where...\] in the Github repository).

```{r}
#| label: renv-restore
renv::restore()
```

## Load functions

Loads the necessary extraction functions to find, extract and process satellite-derived chlorophyll a and SST

```{r}
#| label: call-for-functions
#| echo: false
#| message: false

# Source structure exploration function
source(here::here("functions", "extracting.R"))
```

# Chlorophyll a data (chla)

Chlorophyll a is used as a proxy for ... It will be extracted from ... and processed like so... \[general descripiton\]

## Chlor a data for 2009

### Define path to raw chlor a files for 2009

```{r}
#| label: find-chla-files-2009

# Define path to all the raw chla files 
chla_filenames_2009 <- list.files(
  here::here("raw_data", "chla_raw_2009"),
  full.names = TRUE)
```

### Extract raw chlor a data for 2009 from NASA L3b files

This extracts the relevant data from each of the satellite extarction files.

*See extracting.r document in functions folder to find the data extraction functions*

```{r}
#| label: process-chla-files-2009

chla_data_list_2009 <- lapply(chla_filenames_2009, function(f) {
  extract_sat_data(filepath = f)
})

# Combine into a single data table
raw_chla_2009 <- data.table::rbindlist(chla_data_list_2009, use.names = TRUE, fill = TRUE)

raw_chla_2009 <- raw_chla_2009 %>% filter(!is.na(chlor_a))

message(sprintf("Processing complete. Total rows extracted: %d", nrow(raw_chla_2009)))
```

#### Save raw chlor-a data for 2009

Save the data table with the extracted chlor-a for ease of use

```{r}
#| label: save-chla-data-2009

# Save csv
fwrite(raw_chla_2009, here::here("raw_data", "raw_chla_2009.csv"))

# Load if needed
# raw_chla_2009 <- read.csv(
#   here::here("raw_data", "raw_chla_2009.csv"))

```

### Match raw chlor-a data to each station in 2009-2016 data set

#### Load 2009-2016 fish location data and empty columns to add chor-a data values

```{r}
#| label: load-2009-location-data

# Load location data and add columns to add relevant chlor-a data
locations_2009 <- read_rds(
  here::here("processed_data", "clean_location_2009.rds")) %>% 
  dplyr::select(date, station, site, country, latitude, longitude) %>% 
  dplyr::mutate(closest_cell_lat =0, closest_cell_lon=0, cell_distance_km = 0)
```

#### Find nearest satellite cells of chlor-a values to 2009 stations

\[explain what this chuck is doing exactly...\]

```{r}
#| label: find-nearest-satellite-cells-2009

# For each station, find and assign the nearest satellite cell
locations_with_chla_sat_2009 <- assign_nearest_cell(raw_chla_2009, locations_2009) %>% 
  dplyr::mutate(chla_annual_mean = 0)
```

#### Add annual chlorophyll a means to fish location data

```{r}
#| label: add-annual-mean-2009

# Adding annual chlorophyll a means
locations_with_chla_2009 <- compute_annual_mean(locations_with_chla_sat_2009, raw_chla_2009, 'chlor_a', 'chla_annual_mean')
```

### Save merged data

```{r}
#| label: save-merged-chla-data-2009

# Save csv
fwrite(locations_with_chla_2009 , here::here("processed_data", "locations_with_chla_2009.csv"))

# Load if needed
# locations_with_chla_2009 <- read.csv(
#   here::here("processed_data", "locations_with_chla_2009.csv"))
```

## Chlor a data for 2018

### Define path to raw chlor a files for 2018

```{r}
#| label: find-chla-files-2018

# Define path to all the raw chla files 
chla_filenames_2018 <- list.files(
  here::here("raw_data", "chla_raw_2018"),
  full.names = TRUE)
```

### Extract raw chlor a data for 2018 from NASA L3b files

This extracts the relevant data from each of the statellite extarction files.

*See extracting.r document in functions folder to find the data extraction functions*

```{r}
#| label: process-chla-files-2018

chla_data_list_2018 <- lapply(chla_filenames_2018, function(f) {
  extract_sat_data(filepath = f)
})

# Combine into a single data table
raw_chla_2018 <- data.table::rbindlist(chla_data_list_2018, use.names = TRUE, fill = TRUE)
raw_chla_2018 <- raw_chla_2018 %>% filter(!is.na(chlor_a))
message(sprintf("Processing complete. Total rows extracted: %d", nrow(raw_chla_2018)))
```

#### Save raw chlor-a data for 2018

Save the data table with the extracted chlor-a for ease of use

```{r}
#| label: save-chla-data-2018

# Save csv
fwrite(raw_chla_2018, here::here("raw_data", "raw_chla_2018.csv"))
```

### Match raw chlor-a data to each station in 2018-2016 data set

#### Load 2018-2016 fish location data and empty columns to add chor-a data values

```{r}
#| label: load-2018-location-data

# Load location data and add columns to add relevant chlor-a data
locations_2018 <- read_rds(
  here::here("processed_data", "clean_location_2018.rds")) %>% 
  dplyr::select(date, station, site, country, latitude, longitude) %>% 
  dplyr::mutate(closest_cell_lat =0, closest_cell_lon=0, cell_distance_km = 0)
```

#### Find nearest satellite cells of chlor-a values to 2018 stations

\[explain what this chuck is doing exactly...\]

```{r}
#| label: find-nearest-satellite-cells-2018

# For each station, find and assign the nearest satellite cell
locations_with_chla_sat_2018 <- assign_nearest_cell(raw_chla_2018, locations_2018)%>% 
  dplyr::mutate(chla_annual_mean = 0)
```

#### Add annual chlorophyll a means to fish location data

```{r}
#| label: add-annual-mean-2018

# Adding annual chlorophyll a means
locations_with_chla_2018 <- compute_annual_mean(locations_with_chla_sat_2018, raw_chla_2018, 'chlor_a', 'chla_annual_mean')
```

### Save merged data

```{r}
#| label: save-merged-chla-data-2018

# Save csv
fwrite(locations_with_chla_2018 , here::here("processed_data", "locations_with_chla_2018.csv"))
```

## Quality checks

#### Notes on Data Quality

**Important considerations:**

-   This is a **monthly composite** product, which reduces cloud gaps
-   The spatial resolution is **4km**, so nearby sites will have similar values

### Visualise satellite match on a map for 2009

```{r}
#| label: visualise-sat-match-2009

library(sf)
library(ggplot2)
library(dplyr)
library(rnaturalearth)
library(rnaturalearthdata)


world <- ne_countries(scale = "medium", returnclass = "sf")
focus_region <- world %>%
  filter(admin %in% c("Kenya", "Tanzania", "Comoros"))

# Create the map with a tighter zoom
map_plot_2009 <- ggplot() +
  
  # Plot all countries in Africa as the base layer
  geom_sf(data = world[world$continent == "Africa",], 
          fill = "gray95", color = "gray50", size = 0.3) + 
  
  geom_sf(data = focus_region, fill = "lightblue", color = "gray30", size = 0.3) +
  
  geom_point(data = locations_2009,
             aes(x = longitude, y = latitude, ),
             alpha = 0.6, color = "#440154") +
            # each country gets a different 

  #geom_point(data = raw_chla_2009, 
  #           aes(x = lon, y = lat),
  #           alpha = 0.6, color = "darkblue")+
  geom_point(data = raw_chla_2009, 
             aes(x = lon, y = lat),
             alpha = 0.01, color = "green")+
  
  
  scale_size_continuous(name = "Number of\nObservations", 
                        range = c(0.5, 5),
                        breaks = c(10, 25, 50, 75, 100, 150, 200)) +
  
  # Specify region of focus
  coord_sf(xlim = c(36, 55), ylim = c(1, -19)) + 
  
  labs(title = "Marine Survey Sites in Kenya",
       subtitle = "Point size represents number of fish observations per site",
       x = "Longitude", y = "Latitude") +
  
  theme_minimal() +
  theme(panel.grid.major = linewidth(color = "gray90", size = 0.2),
        panel.background = element_rect(fill = "aliceblue"),
        plot.title = element_text(face = "bold", size = 14),
        legend.position = "right")

print(map_plot_2009)
```

```{r}
#| label: visualise-sat-match-2018

library(sf)
library(ggplot2)
library(dplyr)
library(rnaturalearth)
library(rnaturalearthdata)


world <- ne_countries(scale = "medium", returnclass = "sf")
focus_region <- world %>%
  filter(admin %in% c("Kenya", "Tanzania", "Comoros"))

# Create the map with a tighter zoom
map_plot_2018 <- ggplot() +
  
  # Plot all countries in Africa as the base layer
  geom_sf(data = world[world$continent == "Africa",], 
          fill = "gray95", color = "gray50", size = 0.3) + 
  
  geom_sf(data = focus_region, fill = "lightblue", color = "gray30", size = 0.3) +
  
  geom_point(data = locations_2018,
             aes(x = longitude, y = latitude, ),
             alpha = 0.6, color = "#440154") +
            # each country gets a different 

  geom_point(data = raw_chla_2018, 
             aes(x = lon, y = lat),
             alpha = 0.01, color = "green")+
  
  
  scale_size_continuous(name = "Number of\nObservations", 
                        range = c(0.5, 5),
                        breaks = c(10, 25, 50, 75, 100, 150, 200)) +
  
  # Specify region of focus - Kenya in frame
  coord_sf(xlim = c(36, 49), ylim = c(1, -15)) + 
  
  labs(title = "Marine Survey Sites in Kenya",
       subtitle = "Point size represents number of fish observations per site",
       x = "Longitude", y = "Latitude") +
  
  theme_minimal() +
  theme(panel.grid.major = element_line(color = "gray90", size = 0.2),
        panel.background = element_rect(fill = "aliceblue"),
        plot.title = element_text(face = "bold", size = 14),
        legend.position = "right")

print(map_plot_2018)
```

### Match Quality Histogram 2009

```{r}
#| label: plot-match-quality
#| fig-width: 10
#| fig-height: 6

locations_with_chla_2009 %>%
  filter(!is.na(cell_distance_km)) %>%
  ggplot(aes(x = cell_distance_km)) +
  geom_histogram(bins = 30, fill = "coral", alpha = 0.7) +
  labs(
    title = "Distribution of Spatial Match Distances",
    subtitle = "Distance between survey location and nearest chlorophyll-a grid cell",
    x = "Distance (km)",
    y = "Count"
  ) +
  theme_minimal()
```

### Match Quality Histogram 2018

```{r}
#| label: plot-match-quality
#| fig-width: 10
#| fig-height: 6

locations_with_chla_2018 %>%
  filter(!is.na(cell_distance_km)) %>%
  ggplot(aes(x = cell_distance_km)) +
  geom_histogram(bins = 30, fill = "coral", alpha = 0.7) +
  labs(
    title = "Distribution of Spatial Match Distances",
    subtitle = "Distance between survey location and nearest chlorophyll-a grid cell",
    x = "Distance (km)",
    y = "Count"
  ) +
  theme_minimal()
```