---
title: "human_gravity_metric"
format: html
editor: visual
---

# Set up

```{r}
#|label: load-libraries
#|echo: false

#Load libraries 
library(terra)      # modern raster/vector handling
library(sf)         # vector data
library(gdistance)  # cost-distance modelling
library(units)
library(raster)

library(rnaturalearth)
library(rnaturalearthdata)

```

# Make friction map

## Get land cover raster

```{r}

# Load land cover legend
legend <- read.csv(here::here("land_cover_data", "ESACCI-LC-Legend.csv"), row.names = NULL,
 stringsAsFactors = FALSE,
check.names = FALSE)

# Load and rasterise land cover tif file
land_cover <- rast(here::here("land_cover_data", "ESACCI-LC-L4-LCCS-Map-300m-P5Y-2010-v1.6.1.tif"))
crs(land_cover) <- "EPSG:4326"


# Define area of interest (AOI)
#aoi <- ext(37, 47, 1, -15) # actual coordinates 
aoi <- ext(30, 55, -21,1) #slightly bigger area

# Crop the land cover raster to AOI
land_cover_crop <- crop(land_cover, aoi)

# Plot cropped land cover raster 
plot(land_cover_crop)

# Behrmann is a projection from the earth spherical surface to a flat surface 
# This line specifies how coords should be transformed (manually for consistency)
behrmann <- "+proj=cea +lat_ts=30 +lon_0=0 +datum=WGS84 +units=m +no_defs"

# Transforms land cover raster to behrmann form
land_cover_behr <- project(
  land_cover_crop,
  behrmann,
  method = "near",
  res = 1000 # resolution = 1km
)

```

## Get roads

```{r}
# Load world roads shapefile
roads <- st_read(here::here("road_data", "trans-road-l.shp"))

# Crop roads shapefile to region of interest
roads <- st_crop(st_transform(roads, 4326), aoi)

# transform to behrmann form
roads <- st_transform(roads, behrmann)

# removes all non-line features from roads (e.g. removes polygons)
roads <- roads[st_geometry_type(roads) %in% c("LINESTRING", "MULTILINESTRING"), ]

# Plot 
plot(st_geometry(roads), col = "black", lwd = 0.3,
     main = "VMAP0 Roads and Tracks")

# Convert roads to a spatial vector suitable for use with terra operations
roads_sp <- as(roads, "Spatial")
roads_v  <- terra::vect(roads_sp)

#Make roads 'thicker' in order to make it more continuous (set thickness to resolution which is 1km)
roads_buf <- terra::buffer(roads_v, width = res(land_cover_behr)[1] )

#build road raster
roads_raster <- terra::rasterize(
  roads_buf,
  land_cover_behr,
  field = 1,        # assign value 1 to road cells
  touches = TRUE    # include cells touched by lines
)
plot(roads_raster, col = "red")

```

## Get shoreline shapefile

```{r}

# Load shoreline shapefile
shoreline <- st_read(here::here("shoreline_data", "GSHHS_i_L1.shp"))

# Convert to behrmann form
shoreline <- st_transform(shoreline, behrmann)

# Rasterise shapefile
water_raster <- rasterize(vect(shoreline), land_cover_behr, field = 1)

# Plot 
plot(water_raster)
```

## Creating friction map

```{r}

# Create reference speed table
speed_table <- data.frame(
  land_cover_code = legend$nb_lab,  # example codes
  speed_kmh = legend$speed
)

# set speeds on land cover
speed <- classify(land_cover_behr, speed_table)

# set speeds for roads and water
speed[roads_raster == 1]  <- 60
speed[water_raster != 1] <- 20

# set speed for an NA cell
speed[is.na(speed)] <- 1.4

# create friction raster (ie time taken to travel a cell)
cell_km <- res(speed)[1] / 1000
friction <- (cell_km / speed) # Time in hours to cross a cell

# Plot 
plot(speed)
plot(friction)
```

```{r}
# rasterize friction surface (values = hours per km)
friction_raster <- raster(friction)

# make raster Behrmann-compatible (projected, metres)
crs(friction_raster) <- behrmann

# convert friction from hours per km -> hours per metre
friction_raster_hpm <- friction_raster / 1000

# create transition graph (conductance = metres per hour)
transition_graph <- transition(
  friction_raster_hpm,
  transitionFunction = function(x) 1 / mean(x),
  directions = 8
)

# apply geographic correction so diagonal distances are √2 × cellsize (pythagoras)
transition_graph <- geoCorrection(transition_graph, type = "c")

# save to file
saveRDS(transition_graph, file = here::here("land_cover_data", "transition_graph.rds"))

# load from file
transition_graph <- readRDS(here::here("land_cover_data", "transition_graph.rds"))
```

```{r}
# Function to calculate travel cost btwn two lat/lon pairs
cost_between_points <- function(lon1, lat1, lon2, lat2,
                                transition_graph, friction_raster) {

  p1 <- st_as_sf(data.frame(lon = lon1, lat = lat1),
                 coords = c("lon", "lat"), crs = 4326)
  p2 <- st_as_sf(data.frame(lon = lon2, lat = lat2),
                 coords = c("lon", "lat"), crs = 4326)

  p1p <- st_transform(p1, st_crs(crs(friction_raster)))
  p2p <- st_transform(p2, st_crs(crs(friction_raster)))

  from_xy <- st_coordinates(p1p)
  to_xy   <- st_coordinates(p2p)

  as.numeric(costDistance(transition_graph, from_xy, to_xy))
}
```

```{r}
## TEST THE COST FUNCTION ##
library(raster)
library(gdistance)
library(sf)

# choose a central 10x10 window
center_cell <- cellFromRowCol(
  friction_raster,
  row = round(nrow(friction_raster) / 2),
  col = round(ncol(friction_raster) / 2)
)

rc <- rowColFromCell(friction_raster, center_cell)

rows <- (rc[1] - 4):(rc[1] + 5)
cols <- (rc[2] - 4):(rc[2] + 5)

cells_10x10 <- cellFromRowColCombine(friction_raster, rows, cols)

friction_10x10 <- crop(
  friction_raster,
  extentFromCells(friction_raster, cells_10x10)
)

friction_10x10_hpm <- friction_10x10 / 1000

transition_10x10 <- transition(
  friction_10x10_hpm,
  transitionFunction = function(x) 1 / mean(x),  # metres per hour
  directions = 8
)

transition_10x10 <- geoCorrection(transition_10x10, type = "c")

cell_a <- 1
cell_b <- 13

xy_a <- xyFromCell(friction_10x10, cell_a)
xy_b <- xyFromCell(friction_10x10, cell_b)

p_a_ll <- st_transform(
  st_as_sf(data.frame(x = xy_a[1], y = xy_a[2]),
           coords = c("x", "y"),
           crs = crs(friction_10x10)),
  4326
)

p_b_ll <- st_transform(
  st_as_sf(data.frame(x = xy_b[1], y = xy_b[2]),
           coords = c("x", "y"),
           crs = crs(friction_10x10)),
  4326
)

lon1 <- st_coordinates(p_a_ll)[1]
lat1 <- st_coordinates(p_a_ll)[2]

lon2 <- st_coordinates(p_b_ll)[1]
lat2 <- st_coordinates(p_b_ll)[2]

time_hours <- cost_between_points(
  lon1, lat1,
  lon2, lat2,
  transition_10x10,
  friction_10x10_hpm
)

message(time_hours)

plot(friction_10x10)
points(xy_a[1], xy_a[2], col = "red", pch = 16)
points(xy_b[1], xy_b[2], col = "blue", pch = 16)

```

```{r}
#load world city data
world_cities <- read.csv(here::here("city_data", "world_cities.csv"))

#Remove cities that are too small
cities_to_remove <- c("Mkokotoni","Mtwara", "Koani", "Lindi", "Pemba", "Antsiranana", "Mocambique", "Mamoutzou", "Mahajanga")
world_cities <- world_cities[!world_cities$CITY_NAME %in% cities_to_remove, ]

# function to convert xy coords to lat/lon
xy_to_lonlat <- function(x, y, crs_in = 3857) {
  st_coordinates(
    st_transform(
      st_as_sf(data.frame(x, y), coords = c("x", "y"), crs = crs_in),
      4326
    )
  )
}

city_coords <- xy_to_lonlat(world_cities$X, world_cities$Y)

# get lat lon for cities (previously only had xy coords)
world_cities$lon <- city_coords[, 1]
world_cities$lat <- city_coords[, 2]


# convert cities to sf
cities_sf <- st_as_sf(
  world_cities,
  coords = c("lon", "lat"),
  crs = 4326,
  remove = FALSE
)


filter_cities_bbox <- function(df, lon_min, lon_max, lat_min, lat_max) {
  df[
    df$lon >= lon_min & df$lon <= lon_max &
    df$lat >= lat_min & df$lat <= lat_max,
  ]
}

#filter cities to our wanted region
cities_sf <- filter_cities_bbox(
  cities_sf,
  lon_min = 30, lon_max = 50,
  lat_min = -20, lat_max = 0
)


# get world map for plotting
world <- ne_countries(scale = "medium", returnclass = "sf")

# plot
plot(st_geometry(world), xlim = c(45, 47), ylim = c(-20, 0), col = "grey90", border = "grey60")
points(st_coordinates(cities_sf), pch = 20, cex = 0.6, col = "red")
```

```{r}
# load reef locations data
location_2009 <- readr::read_rds(here::here("processed_data", "clean_location_2009.rds"))


locations_sf <- st_as_sf(
  location_2009,
  coords = c("longitude", "latitude"),
  crs = 4326,
  remove = FALSE
)

# cities (from previous steps)
#cities_sf <- st_as_sf(
#  world_cities,
#  coords = c("lon", "lat"),
#  crs = 4326,
#  remove = FALSE
#)

## For each reef location, find the nearest city:

locations_p <- st_transform(locations_sf, 3857)
cities_p   <- st_transform(cities_sf, 3857)

#Find nearest city and update the columbs
nearest_city_idx <- st_nearest_feature(locations_p, cities_p)
locations_sf$nearest_city <- cities_sf$CITY_NAME[nearest_city_idx]
locations_sf$nearest_city_lat <- cities_sf$lat[nearest_city_idx]
locations_sf$nearest_city_lon <- cities_sf$lon[nearest_city_idx]
locations_sf$nearest_city_pop <- 0
locations_sf$nearest_city_cost_distance <- 0

# Use graph to calculate the travel cost to that nearest city
locations_sf$nearest_city_cost_distance <- mapply(
  cost_between_points,
  locations_sf$lon,
  locations_sf$lat,
  locations_sf$nearest_city_lon,
  locations_sf$nearest_city_lat, 
  MoreArgs = list(transition_graph = transition_graph, friction_raster = friction_raster)
)

unique(locations_sf$nearest_city)

readr::write_rds(locations_sf,here::here("city_data","locations_sf.rds"))

#write.csv(locations_sf, here::here("city_data","locations_sf.csv"))


```

defined as provincial capital cities, major population centres, landmark cities, national capitals, and ports - metropolitan area population

```{r}
# Load locations with closest cities
locations_with_grav <- readr::read_rds(here::here("city_data", "locations_sf.rds"))

# Manually set 2010 pops
cities = c("Zanzibar","Tanga","Dar es Salaam","Mombasa","Kismaayo", "Moroni", "Nampula", "Toamasina")
pops =  c(460000,    260000,  3870000,        940000,    296000,   54000    , 539000,   246000)

locations_with_grav$nearest_city_pop <- pops[match(locations_with_grav$nearest_city, cities)]

locations_with_grav$market_grav = locations_with_grav$nearest_city_pop / (locations_with_grav$nearest_city_cost_distance ** 2)

```

# Settlement Gravity

This is getting the sum of the population in a 50km radius

```{r}
# Load raster
landscan <- rast(here::here("city_data", "landscan-global-2010.tif"))

#Test checks:
#aoi2 <- ext(39.170, 39.358, -6.965, -6.718) #slightly bigger area
#landscan_crop2 <- crop(landscan, aoi2)
#message(global(landscan_crop2, fun = "sum", na.rm = TRUE)
#)


# Crop (optional)
landscan_crop <- crop(landscan, aoi)
plot(landscan_crop,)
res(landscan_crop)


# Convert points
points <- vect(
  location_2009,
  geom = c("longitude", "latitude"),
  crs = "EPSG:4326"
)
# Project BOTH raster and points to a metric CRS
landscan_m <- project(landscan_crop, "EPSG:54009")   # World Mollweide
points_m   <- project(points, crs(landscan_m))

# Create 50 km buffers (meters!)
buffers <- buffer(points_m, width = 50000)

# Extract population
pop_sum <- extract(
  landscan_m,
  buffers,
  fun = sum,
  na.rm = TRUE
)

# Assign correctly by name
locations_with_grav$settlement_tot_pop <- pop_sum$lspop2010
```

```{r}
#TEST
#Zoom into settlement just to see
#  	
# 43.26512
# -11.670252
aoi_zoom <- ext(43.18, 43.32, -11.75,-11.6) #slightly bigger area

landscan_zoomed <- crop(landscan, aoi_zoom)
plot(landscan_zoomed)
max(landscan_zoomed)
```

This gets the nearest cell with population density 75

```{r}
landscan_crop <- project(landscan_crop, "EPSG:3857")

r_crs <- crs(landscan_crop)


# Initialise columns to NA
locations_with_grav$nearest_pop75_lon <- NA_real_
locations_with_grav$nearest_pop75_lat <- NA_real_
locations_with_grav$nearest_pop75_dist_km <- NA_real_

# Try to find the nearest 75 pop cell by progressively increasing search radius
threshold <- 75
start_radius <- 5000    # 5 km
step_radius  <- 5000    # increase by 5 km
max_radius   <- 200000
for (i in seq_len(nrow(locations_with_grav))) {

  # --------------------------------------------------------------
  # Create point from lon/lat
  # --------------------------------------------------------------
  pt_ll <- vect(
    data.frame(
      x = locations_with_grav$longitude[i],
      y = locations_with_grav$latitude[i]
    ),
    geom = c("x", "y"),
    crs = "EPSG:4326"
  )

  # Project point to raster CRS (meters)
  pt_m <- project(pt_ll, r_crs)

  found  <- FALSE
  radius <- start_radius

  # --------------------------------------------------------------
  # Expand search radius until a cell is found
  # --------------------------------------------------------------
  while (!found && radius <= max_radius) {

    # Buffer in meters
    buf <- buffer(pt_m, width = radius)

    # Crop raster to buffer
    sub <- crop(landscan_crop, buf)

    if (!is.null(sub)) {

      # Binary mask of cells above threshold
      m <- sub > threshold
      # Safe check: are there any TRUE cells?
      if (!is.null(m) && ncell(m) > 0 && any(m[], na.rm = TRUE)) {
        #plot(m)
        # Coordinates of qualifying cells
        true_cells <- which(values(m))
        cell_xy <- xyFromCell(m, true_cells)

        # Euclidean distances (meters)
        d <- sqrt(
          (cell_xy[,1] - crds(pt_m)[1,1])^2 +
          (cell_xy[,2] - crds(pt_m)[1,2])^2
        )

        # Nearest qualifying cell
        j <- which.min(d)

        # Create nearest cell point (metric CRS)
        nearest_pt_m <- vect(
          data.frame(x = cell_xy[j,1], y = cell_xy[j,2]),
          geom = c("x", "y"),
          crs = r_crs
        )

        # Convert back to lon/lat for storage
        nearest_pt_ll <- project(nearest_pt_m, "EPSG:4326")
        coords <- crds(nearest_pt_ll)

        # Store results
        locations_with_grav$nearest_pop75_lon[i]     <- coords[1,1]
        locations_with_grav$nearest_pop75_lat[i]     <- coords[1,2]
        locations_with_grav$nearest_pop75_dist_km[i] <- d[j] / 1000

        found <- TRUE
      }
    }

    # Explicit cleanup to avoid memory fragmentation
    rm(sub, m)
    gc()

    radius <- radius + step_radius
  }
}
```

```{r}
#Find population in a 2km radius of the closest 75 pop density cell
df <- sf::st_drop_geometry(locations_with_grav)

points <- terra::vect(
  df,
  geom = c("nearest_pop75_lon", "nearest_pop75_lat"),
  crs = "EPSG:4326"
)

points_m   <- project(points, crs(landscan_m))


buffers <- buffer(points_m, width = 2000)

# Extract population
pop_sum <- extract(
  landscan_m,
  buffers,
  fun = sum,
  na.rm = TRUE
)


locations_with_grav$nearest_settlement_pop <- pop_sum$lspop2010

```

```{r}
locations_with_grav$nearest_pop_cost_distance <- mapply(
  cost_between_points,
  locations_with_grav$lon,
  locations_with_grav$lat,
  locations_with_grav$nearest_pop75_lon,
  locations_with_grav$nearest_pop75_lat, 
  MoreArgs = list(transition_graph = transition_graph, friction_raster = friction_raster)
)

#Set a minimum value of 0.005 (as some are 0)
locations_with_grav$nearest_pop_cost_distance <- pmax(locations_with_grav$nearest_pop_cost_distance, 0.005)

locations_with_grav$nearest_pop75_grav = locations_with_grav$nearest_settlement_pop / (locations_with_grav$nearest_pop_cost_distance ^ 2)
```

```{r}
#Save the final product
readr::write_rds(locations_with_grav,here::here("city_data","locations_with_grav.rds"))
```

#end
