---
title: "pisc_model_testing"
format: html
editor: visual
---
---
title: "tot_biomass_model_testing"
format: html
editor: visual
---

# Overview

This document explores the relationship between human gravity and fish biomass

# R Set-up

*Important note: Comprehensive instructions, including all prerequisites and detailed guidelines for running this analysis and relevant code, are available in the README file of my [GitHub repository](...) for this project. Please refer to the README ... to ensure proper setup and usage.*

```{r}
options(scipen = 999)
```

```{r}
#| label: renv-restore
# renv::restore()
```


Use the renv::restore() to load up the necessary package versions along with the necessary dependencies.

```{r}
#| label: r set-up
#| echo: false
#| include: false

#Load require libraries
library(tidyverse)
library(sf)
library(arm)
library(here)
library(dplyr)
library(tidyr)
library(ggplot2)
library(ggfortify) # extends ggplot2 for some more ease with stats
library(MASS) # fits regression models etc.
library(MuMIn) # multi-model inference (AIC etc.)
library(AICcmodavg)
library(DHARMa)
library(tweedie)
library(glmmTMB)
library(statmod)
```

## Load data

```{r}
#| label: load-data

fish_2009 <- readr::read_rds(here::here("processed_data", "clean_fish_2009.rds"))
location_2009 <- readr::read_rds(here::here("processed_data", "clean_location_2009.rds"))
gravity_2009 <- readr::read_rds(here::here("city_data", "locations_with_grav_combined.rds"))
gravity_2009_2 <- readr::read_rds(here::here("city_data", "locations_with_grav.rds"))
old_grav_2009 <- read.csv(here::here("processed_data", "locations_with_gravity.csv"))
chla_2009 <- read.csv(here::here("processed_data", "locations_with_chla_2009.csv"))
sst_2009 <- read.csv(here::here("processed_data", "locations_with_sst_2009.csv"))
dhw_2009 <- readr::read_rds(here::here("processed_data", "locations_with_dhw_2009.rds"))
rugosity_2009 <- readr::read_rds(here::here("processed_data", "clean_dive_details_2009.rds"))
```

# Piscivore biomass

Including zeros:

```{r}
piscivore_data <- fish_2009 %>%
  group_by(site, station, ts_no, date) %>%
  summarise(
    transect_pisc_biomass = sum(
      ifelse(trophic_group == "piscivores", tot_wt_g, 0),
      na.rm = TRUE
    ),
    country = first(country),
    .groups = "drop"
  ) %>%
  group_by(site) %>%
  filter(n() >= 3) %>%
  ungroup() %>%
  filter(
    country != "kenya",
    !site %in% c("fernauvloso", "paradise")
  ) %>%
  mutate(
    country = as.factor(country),
    site = as.factor(site)
  )
```

```{r}
summary(piscivore_data)

zeros <- mean(piscivore_data$transect_pisc_biomass == 0, na.rm = TRUE)
cat("Proportion of zeros:", zeros, "\n")

( pisc_hist <- ggplot(piscivore_data, aes(x = transect_pisc_biomass)) +
  geom_histogram(bins = 50) +
  labs(
    x = "Piscivore biomass per transect",
    y = "Frequency"
  ) +
  theme_bw() )
```


log-transformation:

```{r}
piscivore_data <- piscivore_data %>%
  mutate(
    log_pisc_biomass = log(transect_pisc_biomass + 0.01),
    sqrt_pisc_biomass = sqrt(transect_pisc_biomass + 0.01))


(log_biomass_hist <-ggplot(piscivore_data, aes(x = log_pisc_biomass)) +
  geom_histogram(bins = 30) +
  labs(
    x = "Log-transformed biomass",
    y = "Frequency"
  ) +
  theme_bw() )

(sqrt_biomass_hist <-ggplot(piscivore_data, aes(x = sqrt_pisc_biomass)) +
  geom_histogram(bins = 30) +
  labs(
    x = "Sqrt-transformed biomass",
    y = "Frequency"
  ) +
  theme_bw() )
```
Despite having a few outliers, the log transform is better here.

### Calculate gravity means per site

The gravity values were calculated for each station - as these were the reef sampling locations that were geo-referenced in the data. This code aggregates human‐gravity metrics to the site level by calculating mean values of the stations (excluding missing data) for each variable within sites. Note that the gravity metrics are the updated metrics from 2009, and the old gravity refers to the market gravity extracted from \[site\]. All the metrics are then combined into a table for ease of use.

```{r}
#| label: calculate-env-var-means

# Calculate mean human gravity per site
gravity_2009_sites <- gravity_2009 %>%
  st_drop_geometry() %>%
  group_by(site) %>%
  summarise(
    market_gravity = mean(market_grav, na.rm = TRUE),
    settlement_tot_pop = mean(settlement_tot_pop, na.rm = TRUE),
    nearest_pop_grav = mean(nearest_pop75_grav, na.rm = TRUE),
    .groups = "drop"
  )

# Calculate mean human gravity per site
old_grav_2009_sites <- old_grav_2009 %>% 
  group_by(site) %>% 
  summarise(
    old_gravity = mean(Grav_tot),
    .groups = "drop"
  )

# Combine gravity metrics
gravity_metrics_2009 <- gravity_2009_sites %>%
  left_join(old_grav_2009_sites, by = "site")

# Model data:
grav_model_data <- piscivore_data %>%
  left_join(gravity_metrics_2009, by = "site")
```

# Human gravity

## Correlation matrix
```{r}
grav_vars <- grav_model_data %>% 
  dplyr::select(market_gravity, settlement_tot_pop, nearest_pop_grav)

# Correlation matrix
cor_matrix <- cor(grav_vars, use = "complete.obs")
cor_matrix

# Visualise
library(corrplot)
corrplot(cor_matrix, method = "color", type = "lower")

```


```{r}
grav_model_data <- grav_model_data %>%
  mutate(
    log_market_gravity = log(market_gravity),
    log_nearest_pop_grav = log(nearest_pop_grav),
    log_settlement_tot_pop = log(settlement_tot_pop),

    log_market_gravity_sc = as.numeric(scale(log_market_gravity)),
    log_nearest_pop_grav_sc = as.numeric(scale(log_nearest_pop_grav)),
    log_settlement_tot_pop_sc = as.numeric(scale(log_settlement_tot_pop))
  )

```

```{r}
# Gaussian (we use glm so it's likelihood-based)
pisc_glm_gaussian <- glmmTMB::glmmTMB(transect_pisc_biomass ~ log_nearest_pop_grav_sc + (1|site),
                 family = gaussian(link = "identity"),
                 data = grav_model_data)
summary(pisc_glm_gaussian)

# Gaussian (we use glm so it's likelihood-based)
log_pisc_glm_gaussian <- glmmTMB::glmmTMB(log_pisc_biomass ~ log_nearest_pop_grav_sc + (1|site),
                 family = gaussian(link = "identity"),
                 data = grav_model_data)
summary(log_pisc_glm_gaussian)
```

```{r}
pisc_glm_tweedie <- glmmTMB::glmmTMB(transect_pisc_biomass ~ log_nearest_pop_grav_sc + (1|site),
                 family = glmmTMB::tweedie(link = "log"),
                 data = grav_model_data)
summary(pisc_glm_tweedie)

grav_model_data$nearest_pop_grav_sc <- scale(grav_model_data$nearest_pop_grav)

pisc_glm_tweedie_2 <- glmmTMB::glmmTMB(transect_pisc_biomass ~ nearest_pop_grav_sc + (1|site),
                 family = glmmTMB::tweedie(link = "log"),
                 data = grav_model_data)
summary(pisc_glm_tweedie_2)
```

```{r}
library(splines)

pisc_glm_tweedie_splin <- glmmTMB::glmmTMB(
  transect_pisc_biomass ~ ns(log_nearest_pop_grav_sc, df = 3) + (1|site),
  family = glmmTMB::tweedie(link = "log"),
  data = grav_model_data
)
summary(pisc_glm_tweedie_splin)
```
```{r}
library(mgcv)

grav_model_data$site <- factor(grav_model_data$site)

gam_model <- gam(
  transect_pisc_biomass ~ s(log_nearest_pop_grav_sc, k = 3) + s(site, bs = "re"),
  family = tw(link = "log"),
  method = "REML",
  data = grav_model_data
)

summary(gam_model)
```

```{r}
plot(gam_model, pages = 1)
plot(gam_model, pages = 1, shade = TRUE)

plot(gam_model, select = 1, shade = TRUE)

plot(log(transect_pisc_biomass + 1) ~ log_nearest_pop_grav_sc, data=grav_model_data)

remove.packages("dplyr")
install.packages("dplyr")
install.packages("gratia")

```





```{r}
library(DHARMa)

# Gaussian model
res_gaussian <- simulateResiduals(log_pisc_glm_gaussian, n = 1000)

# Tweedie model
res_tweedie <- simulateResiduals(pisc_glm_tweedie, n = 1000)
res_tweedie_2 <- simulateResiduals(pisc_glm_tweedie_2, n = 1000)


plot(res_gaussian)
plot(res_tweedie)
plot(res_tweedie_2)
```

Hmmm it seems a little bit non-linear but maybe the tweedie model is ok?



```{r}
mg_glmer <- glmer(transect_pisc_biomass ~ log_market_gravity_sc + (1|site),  family = Gamma(link = "log"),
  data = grav_model_data)

summary(mg_glmer)

mg_glmer2 <- glmer(transect_pisc_biomass ~ log_market_gravity_sc + (1|site),  family = gaussian(link = "log"),
  data = grav_model_data)

summary(mg_glmer2)

st_glmer <- glmer(transect_biomass ~ log_settlement_tot_pop_sc+ (1|site),  family = Gamma(link = "log"),
  data = grav_model_data)

summary(st_glmer)

sg_glmer <- glmer(transect_biomass ~ log_nearest_pop_grav_sc + (1|site),  family = Gamma(link = "log"),
  data = grav_model_data)

summary(sg_glmer)

```

```{r}
grav_metric_models <- list(mg_glmer, st_glmer, sg_glmer)
aictab(grav_metric_models)

anova(mg_glmer, st_glmer, sg_glmer)
```

```{r}
ggplot(grav_model_data, aes(x = log_nearest_pop_grav_sc, y = transect_biomass)) +
  geom_point(alpha = 0.5) +
  geom_smooth(method = "glm", method.args = list(family = Gamma(link = "log"))) +
  scale_y_log10() +
  theme_bw() +
  labs(
    x = "Scaled log population gravity",
    y = "Transect biomass"
  )
```

# Chlorophyll a 

```{r}
# Calculate mean chlorophyll per site
chla_2009_sites <- chla_2009 %>%
  group_by(site) %>%
  summarise(
    mean_annual_chla = mean(chla_annual_mean, na.rm = TRUE),
    .groups = "drop"
  )

# Join chlorophyll to biomass model data
chla_model_data <- ts_biomass_data %>%
  left_join(chla_2009_sites, by = "site")
```

```{r}
# Overview of the data
summary(chla_model_data)

# Histogram 
( chla_hist <- ggplot(chla_model_data, aes(x = mean_annual_chla)) +
  geom_histogram(bins = 30) +
  labs(
    x = expression("Mean Annual Chlorophyll-a"),
    y = "Frequency"
  ) +
  theme_bw() )

# Boxcox 
boxcox(lm(mean_annual_chla ~ 1, data = chla_model_data),
       lambda = seq(-2, 2, 0.1))
```


```{r}

# Scale the data 
chla_model_data <- chla_model_data %>%
  mutate(
    log_chla = log(mean_annual_chla),
    log_chla_sc = as.numeric(scale(log_chla))
  )

chla_glmer <- glmer(
  transect_biomass ~ log_chla_sc + (1 | site),
  family = Gamma(link = "log"),
  data = chla_model_data
)

summary(chla_glmer)

```

```{r}
ggplot(chla_model_data, aes(x = log_chla_sc, y = transect_biomass)) +
  geom_point(alpha = 0.5) +
  geom_smooth(method = "glm",
              method.args = list(family = Gamma(link = "log")),
              se = TRUE) +
  scale_y_log10() +
  labs(
    x = "Chlorophyll-a",
    y = "Transect biomass (log scale)"
  ) +
  theme_bw()

```


# SST 

```{r}
# Calculate mean chlorophyll per site
sst_2009_sites <- sst_2009 %>%
  group_by(site) %>%
  summarise(
    mean_annual_sst = mean(sst_annual_mean, na.rm = TRUE),
    .groups = "drop"
  )

# Join chlorophyll to biomass model data
sst_model_data <- ts_biomass_data %>%
  left_join(sst_2009_sites, by = "site")

```

```{r}
# Overview of the data
summary(sst_model_data)

# Histogram 
( sst_hist <- ggplot(sst_model_data, aes(x = mean_annual_sst)) +
  geom_histogram(bins = 30) +
  labs(
    x = expression("Total biomass in a transect"),
    y = "Frequency"
  ) +
  theme_bw() )
```


```{r}

# Scale the data 
sst_model_data <- sst_model_data %>%
  mutate(
    sst_sc = as.numeric(scale(mean_annual_sst))
  )

sst_glmer <- glmer(
  transect_biomass ~ sst_sc + (1 | site),
  family = Gamma(link = "log"),
  data = sst_model_data
)

summary(sst_glmer)

```

```{r}
ggplot(sst_model_data, aes(x = sst_sc, y = transect_biomass)) +
  geom_point(alpha = 0.5) +
  geom_smooth(method = "glm",
              method.args = list(family = Gamma(link = "log")),
              se = TRUE) +
  scale_y_log10() +
  labs(
    x = "Sea Surface Temperature",
    y = "Transect biomass (log scale)"
  ) +
  theme_bw()

```


# DHW

```{r}

# Calculate mean dhw per site
dhw_2009_sites <- dhw_2009 %>% 
  group_by(site) %>% 
  summarise(
    mean_max_dhw = mean(max_dhw),
    .groups = "drop"
  )

# # Join chlorophyll to biomass model data
dhw_model_data <- ts_biomass_data %>%
  left_join(dhw_2009_sites, by = "site")
```

```{r}

# Overview of the data
summary(dhw_model_data)

# Histogram 
( dhw_hist <- ggplot(dhw_model_data, aes(x = mean_max_dhw)) +
  geom_histogram(bins = 30) +
  labs(
    x = expression("Total biomass in a transect"),
    y = "Frequency"
  ) +
  theme_bw() )
```


```{r}
dhw_model_data <- dhw_model_data %>%
  mutate(
    dhw_log = log(mean_max_dhw + 0.01),
    dhw_log_sc = scale(dhw_log)
  )

dhw_glmer <- glmer(
  transect_biomass ~ dhw_log_sc + (1 | site),
  family = Gamma(link = "log"),
  data = dhw_model_data
)

summary(dhw_glmer)
```

```{r}
ggplot(dhw_model_data, aes(x = dhw_log_sc, y = transect_biomass)) +
  geom_point(alpha = 0.5) +
  geom_smooth(method = "glm",
              method.args = list(family = Gamma(link = "log")),
              se = TRUE) +
  scale_y_log10() +
  labs(
    x = "Cumulative Thermal Stress (degree-heating weeks)",
    y = "Transect biomass (log scale)"
  ) +
  theme_bw()

```

```{r}
temp_metric_models <- list(sst_glmer, dhw_glmer)
aictab(temp_metric_models)
```
# Rugosity

```{r}
# Select rugosity per site
rugosity_2009_sites <- rugosity_2009 %>%
  group_by(site) %>%
  summarise(rugosity = mean(rugosity), .groups = "drop")

# Join chlorophyll to biomass model data
rugosity_model_data <- ts_biomass_data %>%
  left_join(rugosity_2009_sites, by = "site")
```

```{r}
# Overview of the data
summary(rugosity_model_data)

# Histogram 
( rugosity_hist <- ggplot(rugosity_model_data, aes(x = rugosity)) +
  geom_histogram(bins = 30) +
  labs(
    x = expression("Total biomass in a transect"),
    y = "Frequency"
  ) +
  theme_bw() )

```
```{r}

# Scale the data 
sst_model_data <- rugosity_model_data %>%
  mutate(
    rugosity_sc = as.numeric(scale(rugosity))
  )

rugosity_glmer <- glmer(
  transect_biomass ~ rugosity + (1 | site),
  family = Gamma(link = "log"),
  data = rugosity_model_data
)

summary(rugosity_glmer)

rugosity_glmer_sc <- glmer(
  transect_biomass ~ rugosity_sc + (1 | site),
  family = Gamma(link = "log"),
  data = rugosity_model_data
)

summary(rugosity_glmer_sc)

rugosity_glmer_cat <- glmer(
  transect_biomass ~ factor(rugosity) + (1 | site),
  family = Gamma(link = "log"),
  data = rugosity_model_data
)

summary(rugosity_glmer_cat)


```
```{r}

ggplot(rugosity_model_data, aes(x = rugosity, y = transect_biomass)) +
  geom_point(alpha = 0.5) +
  geom_smooth(
    method = "glm",
    method.args = list(family = Gamma(link = "log")),
    se = TRUE
  ) +
  scale_y_log10() +
  labs(
    x = "Rugosity",
    y = "Transect biomass (log scale)"
  ) +
  theme_bw()

```



```{r}
rugosity_models <- list(rugosity_glmer, rugosity_glmer_cat)
aictab(rugosity_models)
```
Keep the continuous one - its ordinal!


# Multivariate model

```{r}
full_model_data <- ts_biomass_data %>%
  left_join(gravity_2009_sites, by = "site") %>%
  left_join(chla_2009_sites, by = "site") %>%
  left_join(sst_2009_sites, by = "site") %>%
  left_join(rugosity_2009_sites, by = "site")


full_model_data <- full_model_data %>%
  mutate(
    # gravity
    log_market_gravity_sc = scale(log(market_gravity)),
    log_nearest_pop_grav_sc = scale(log(nearest_pop_grav)),

    # chla
    log_chla_sc = scale(log(mean_annual_chla)),

    # sst
    sst_sc = scale(mean_annual_sst),
    
    #rugosity
     rugosity_sc = scale(rugosity)
  )
```


## Correlation matrix
```{r}
vars <- full_model_data %>%
  dplyr::select(rugosity_sc, log_chla_sc, sst_sc, log_nearest_pop_grav_sc)

cor(vars, use = "complete.obs")

```
```{r}

m0 <- glmer(
  transect_biomass ~ 1 + (1 | site),
  family = Gamma(link = "log"),
  data = full_model_data
)

m1 <- glmer(
  transect_biomass ~ rugosity + log_nearest_pop_grav_sc + (1 | site),
  family = Gamma(link = "log"),
  data = full_model_data
)

m2 <- glmer(
  transect_biomass ~ rugosity + sst_sc +(1 | site),
  family = Gamma(link = "log"),
  data = full_model_data
)

m3 <- glmer(
  transect_biomass ~ rugosity + log_chla_sc + (1 | site),
  family = Gamma(link = "log"),
  data = full_model_data
)

m4 <- glmer(
  transect_biomass ~ rugosity + log_nearest_pop_grav_sc + sst_sc +(1 | site),
  family = Gamma(link = "log"),
  data = full_model_data
)

m5 <- glmer(
  transect_biomass ~ rugosity + log_nearest_pop_grav_sc + log_chla_sc + (1 | site),
  family = Gamma(link = "log"),
  data = full_model_data
)

m6 <- glmer(
  transect_biomass ~ rugosity + log_nearest_pop_grav_sc + sst_sc + log_chla_sc + (1 | site),
  family = Gamma(link = "log"),
  data = full_model_data
)

aictab(list(m0, m1, m2, m3, m4, m5, m6))


```
