---
title: "spatial_analyses"
format: html
editor: visual
---

# 

# Overview

This document explores the 2009 dataset. This data ...

# R Set-up

*Important note: Comprehensive instructions, including all prerequisites and detailed guidelines for running this analysis and relevant code, are available in the README file of my [GitHub repository](...) for this project. Please refer to the README ... to ensure proper setup and usage.*

Load the necessary packages (these can be found in ...) \## Packages

```{r}
#| label: r set-up
#| echo: false
#| include: false

#Load require libraries
library(tidyverse)
library(dplyr)
library(tidyr)
library(sf)
library(ggplot2)
library(rnaturalearth)

```

## Set working space

Use the renv::restore() to load up the necessary package versions along with the necessary dependencies.

```{r}
#| label: renv-restore
renv::restore()
```

## Load data

```{r}
#| label: load-data

fish_data_2009 <- readr::read_rds(here::here("processed_data", "clean_fish_2009.rds"))
location_2009 <- readr::read_rds(here::here("processed_data", "clean_location_2009.rds"))
sst_2009 <- read.csv(here::here("processed_data", "locations_with_sst_2009.csv"))
chla_2009 <- read.csv(here::here("processed_data", "locations_with_chla_2009.csv"))
h_gravity_2009 <- read.csv(here::here("processed_data", "locations_with_gravity.csv"))
```

# Spatial analyses

## Biomass

### Preparing data

```{r}
# Add lat/lon data for stations
biomass_location_data <- location_2009 %>%
  dplyr::select(station, latitude, longitude)

fish_2009 <- fish_data_2009 %>%
  left_join(biomass_location_data, by = "station")

# Create site lat/lon as mean of station lat/lon
biomass_location_data <- fish_2009 %>%
  group_by(site) %>%
  mutate(
    site_lon = mean(longitude, na.rm = TRUE),
    site_lat = mean(latitude,  na.rm = TRUE),
    biomass = mean(tot_wt_g, na.rm = TRUE),
  ) %>%
  ungroup()
```

### Plot Biomass

```{r}

# World map
world <- ne_countries(scale = "medium", returnclass = "sf")

plot <- ggplot() +
  geom_sf(data = world, fill = "grey95", colour = "grey70") +
  geom_point(
    data = biomass_location_data,
    aes(x = site_lon, y = site_lat, colour = biomass),
    size = 1.5
  ) +
  scale_colour_gradient(
    low = "#56B1F7",
    high = "red3",
    na.value = "grey78",
    name = "Biomass"
  ) +
  coord_sf(
    xlim = c(38.5, 55),
    ylim = c(-19, -1),
    expand = FALSE
  ) +
  theme_classic() +   theme(aspect.ratio = 1)   #wider map

plot

```

## Chlor a

### Plot

```{r}

# World map
world <- ne_countries(scale = "medium", returnclass = "sf")

plot <- ggplot() +
  geom_sf(data = world, fill = "grey95", colour = "grey70") +
  geom_point(
    data = chla_2009,
    aes(x = longitude, y = latitude, colour = chla_annual_mean),
    size = 1.5
  ) +
  scale_colour_gradient(
    low = "lightgreen",
    high = "darkgreen",
    na.value = "grey78",
    name = "Chlorophyll"
  ) +
  coord_sf(
    xlim = c(38.5, 55),
    ylim = c(-19, -1),
    expand = FALSE
  ) +
  theme_classic() +   theme(aspect.ratio = 1)   #square map

plot

```

## SST

### Plot

```{r}

# World map
world <- ne_countries(scale = "medium", returnclass = "sf")

plot <- ggplot() +
  geom_sf(data = world, fill = "grey95", colour = "grey70") +
  geom_point(
    data = sst_2009,
    aes(x = longitude, y = latitude, colour = sst_annual_mean),
    size = 1.5
  ) +
  scale_colour_gradient(
    low = "salmon",
    high = "darkred",
    na.value = "grey78",
    name = "SST"
  ) +
  coord_sf(
    xlim = c(38.5, 45),
    ylim = c(-14, -1),
    expand = FALSE
  ) +
  theme_classic() +   theme(aspect.ratio = 1)   #square map

plot

```

## Human gravity

### Plot

```{r}

# Log data for plotting 
h_gravity_2009$log_grav_tot <- log(h_gravity_2009$Grav_tot)


# World map
world <- ne_countries(scale = "medium", returnclass = "sf")

plot <- ggplot() +
  geom_sf(data = world, fill = "grey95", colour = "grey70") +
  geom_point(
    data = h_gravity_2009,
    aes(x = longitude, y = latitude, colour = log_grav_tot),
    size = 1.5
  ) +
  scale_colour_gradient(
    low = "slategray1",
    high = "slateblue4",
    na.value = "grey78",
    name = "Human gravity"
  ) +
  coord_sf(
    xlim = c(38.5, 55),
    ylim = c(-19, -1),
    expand = FALSE
  ) +
  theme_classic() +   theme(aspect.ratio = 1)   #square map

plot
```

## Reef geomorphologies

### Plot

```{r}

# Data
fish_with_geomorph <- location_2009[, c("station", "reef_geomorphology")]
fish_2009 <- biomass_location_data %>%
  left_join(location_subset, by = "station")


# World map
world <- ne_countries(scale = "medium", returnclass = "sf")

plot <- ggplot() +
  geom_sf(data = world, fill = "grey95", colour = "grey70") +
  geom_point(
    data = fish_2009,
    aes(x = site_lon, y = site_lat, colour = biomass, shape= reef_geomorphology),
    size = 1.5
  ) +
  scale_colour_gradient(
    low = "#56B1F7",
    high = "red3",
    na.value = "grey78",
    name = "Biomass"
  ) +
  coord_sf(
    xlim = c(38.5, 45),
    ylim = c(-14, -1),
    expand = FALSE
  ) +
  theme_classic() +   theme(aspect.ratio = 0.7)   #wider map

plot
```

# Find clusters in the data (unfinished...)

## stations

```{r}
fish_2009_2 <- fish_data_2009 %>%
  left_join(
    location_2009 %>%
      dplyr::select(station, latitude, longitude) %>%
      distinct(),
    by = "station"
  )

station_means <- fish_2009_2 %>%
  group_by(station, latitude, longitude) %>%
  summarise(
    biomass_station = mean(tot_wt_g, na.rm = TRUE),
    .groups = "drop"
  )
```

```{r}
# convert to s[atial object
library(sf)

station_sf <- st_as_sf(
  station_means,
  coords = c("longitude", "latitude"),
  crs = 4326
)

```

```{r}
# Define spatial neighbours

library(spdep)

coords <- st_coordinates(station_sf)

# define neighbours within 100 km
nb_station <- dnearneigh(
  coords,
  d1 = 0,
  d2 = 100000  # metres
)

lw_station <- nb2listw(nb_station, style = "W", zero.policy = TRUE)

```


