---
title: "human_gravity_metric"
format: html
editor: visual
---

# Set up

```{r}
#|label: load-libraries
#|echo: false

#Load libraries 
library(terra)      # modern raster/vector handling
library(sf)         # vector data
library(gdistance)  # cost-distance modelling
library(dplyr)
library(units)
library(raster)
library(rnaturalearth)
library(rnaturalearthdata)

```

```{r}
options(scipen = 999) # keeps everything in readable (non-scientific) format
```

# Make friction map

## Get land cover raster

```{r}

# Load land cover legend
legend <- read.csv(here::here("land_cover_data", "ESACCI-LC-Legend.csv"), row.names = NULL,
 stringsAsFactors = FALSE,
check.names = FALSE)

# Load and rasterise land cover tif file
land_cover <- rast(here::here("land_cover_data", "ESACCI-LC-L4-LCCS-Map-300m-P5Y-2010-v1.6.1.tif"))
crs(land_cover) <- "EPSG:4326"


# Define area of interest (AOI)
#aoi <- ext(37, 47, 1, -15) # actual coordinates 
aoi <- ext(30, 55, -21,1) #slightly bigger area

# Crop the land cover raster to AOI
land_cover_crop <- crop(land_cover, aoi)

# Plot cropped land cover raster 
plot(land_cover_crop)

# Behrmann is a projection from the earth spherical surface to a flat surface 
# This line specifies how coords should be transformed (manually for consistency)
behrmann <- "+proj=cea +lat_ts=30 +lon_0=0 +datum=WGS84 +units=m +no_defs"

# Transforms land cover raster to behrmann form
land_cover_behr <- project(
  land_cover_crop,
  behrmann,
  method = "near",
  res = 1000 # resolution = 1km
)

```

## Get roads

```{r}
# Load world roads shapefile
roads <- st_read(here::here("road_data", "trans-road-l.shp"))

# Crop roads shapefile to region of interest
roads <- st_crop(st_transform(roads, 4326), aoi)

# transform to behrmann form
roads <- st_transform(roads, behrmann)

# removes all non-line features from roads (e.g. removes polygons)
roads <- roads[st_geometry_type(roads) %in% c("LINESTRING", "MULTILINESTRING"), ]

# Plot 
plot(st_geometry(roads), col = "black", lwd = 0.3,
     main = "VMAP0 Roads and Tracks")

# Convert roads to a spatial vector suitable for use with terra operations
roads_sp <- as(roads, "Spatial")
roads_v  <- terra::vect(roads_sp)

#Make roads 'thicker' in order to make it more continuous (set thickness to resolution which is 1km)
roads_buf <- terra::buffer(roads_v, width = res(land_cover_behr)[1] )

#build road raster
roads_raster <- terra::rasterize(
  roads_buf,
  land_cover_behr,
  field = 1,        # assign value 1 to road cells
  touches = TRUE    # include cells touched by lines
)
plot(roads_raster, col = "red")

```

## Get shoreline shapefile

```{r}

# Load shoreline shapefile
shoreline <- st_read(here::here("shoreline_data", "GSHHS_i_L1.shp"))

# Convert to behrmann form
shoreline <- st_transform(shoreline, behrmann)

# Rasterise shapefile
water_raster <- rasterize(vect(shoreline), land_cover_behr, field = 1)

# Plot 
plot(water_raster)
```

## Creating friction map

```{r}

# Create reference speed table
speed_table <- data.frame(
  land_cover_code = legend$nb_lab,  # example codes
  speed_kmh = legend$speed
)

# set speeds on land cover
speed <- classify(land_cover_behr, speed_table)

# set speeds for roads and water
speed[roads_raster == 1]  <- 60
speed[water_raster != 1] <- 20

# set speed for an NA cell
speed[is.na(speed)] <- 1.4

# create friction raster (ie time taken to travel a cell)
cell_km <- res(speed)[1] / 1000
friction <- (cell_km / speed) # Time in hours to cross a cell

# Plot 
plot(speed)
plot(friction)
```

```{r}
# rasterize friction surface (values = hours per km)
friction_raster <- raster(friction)

# make raster Behrmann-compatible (projected, metres)
crs(friction_raster) <- behrmann

# convert friction from hours per km -> hours per metre
friction_raster_hpm <- friction_raster / 1000

# create transition graph (conductance = metres per hour)
transition_graph <- transition(
  friction_raster_hpm,
  transitionFunction = function(x) 1 / mean(x),
  directions = 8
)

# apply geographic correction so diagonal distances are √2 × cellsize (pythagoras)
transition_graph <- geoCorrection(transition_graph, type = "c")

# save to file
saveRDS(transition_graph, file = here::here("land_cover_data", "transition_graph.rds"))

# load from file
transition_graph <- readRDS(here::here("land_cover_data", "transition_graph.rds"))
```

```{r}
# Function to calculate travel cost btwn two lat/lon pairs
cost_between_points <- function(lon1, lat1, lon2, lat2,
                                transition_graph, friction_raster) {

  p1 <- st_as_sf(data.frame(lon = lon1, lat = lat1),
                 coords = c("lon", "lat"), crs = 4326)
  p2 <- st_as_sf(data.frame(lon = lon2, lat = lat2),
                 coords = c("lon", "lat"), crs = 4326)

  p1p <- st_transform(p1, st_crs(crs(friction_raster)))
  p2p <- st_transform(p2, st_crs(crs(friction_raster)))

  from_xy <- st_coordinates(p1p)
  to_xy   <- st_coordinates(p2p)

  as.numeric(costDistance(transition_graph, from_xy, to_xy))
}
```

```{r}
## TEST THE COST FUNCTION ##
library(raster)
library(gdistance)
library(sf)

# choose a central 10x10 window
center_cell <- cellFromRowCol(
  friction_raster,
  row = round(nrow(friction_raster) / 2),
  col = round(ncol(friction_raster) / 2)
)

rc <- rowColFromCell(friction_raster, center_cell)

rows <- (rc[1] - 4):(rc[1] + 5)
cols <- (rc[2] - 4):(rc[2] + 5)

cells_10x10 <- cellFromRowColCombine(friction_raster, rows, cols)

friction_10x10 <- crop(
  friction_raster,
  extentFromCells(friction_raster, cells_10x10)
)

friction_10x10_hpm <- friction_10x10 / 1000

transition_10x10 <- transition(
  friction_10x10_hpm,
  transitionFunction = function(x) 1 / mean(x),  # metres per hour
  directions = 8
)

transition_10x10 <- geoCorrection(transition_10x10, type = "c")

cell_a <- 1
cell_b <- 13

xy_a <- xyFromCell(friction_10x10, cell_a)
xy_b <- xyFromCell(friction_10x10, cell_b)

p_a_ll <- st_transform(
  st_as_sf(data.frame(x = xy_a[1], y = xy_a[2]),
           coords = c("x", "y"),
           crs = crs(friction_10x10)),
  4326
)

p_b_ll <- st_transform(
  st_as_sf(data.frame(x = xy_b[1], y = xy_b[2]),
           coords = c("x", "y"),
           crs = crs(friction_10x10)),
  4326
)

lon1 <- st_coordinates(p_a_ll)[1]
lat1 <- st_coordinates(p_a_ll)[2]

lon2 <- st_coordinates(p_b_ll)[1]
lat2 <- st_coordinates(p_b_ll)[2]

time_hours <- cost_between_points(
  lon1, lat1,
  lon2, lat2,
  transition_10x10,
  friction_10x10_hpm
)

message(time_hours)

plot(friction_10x10)
points(xy_a[1], xy_a[2], col = "red", pch = 16)
points(xy_b[1], xy_b[2], col = "blue", pch = 16)

```

```{r}
#load world city data
world_cities <- read.csv(here::here("city_data", "world_cities.csv"))

#Remove cities that are too small
cities_to_remove <- c("Mkokotoni","Mtwara", "Koani", "Lindi", "Pemba", "Antsiranana", "Mocambique", "Mamoutzou", "Mahajanga")

world_cities <- world_cities[!world_cities$CITY_NAME %in% cities_to_remove, ]

# function to convert xy coords to lat/lon
xy_to_lonlat <- function(x, y, crs_in = 3857) {
  st_coordinates(
    st_transform(
      st_as_sf(data.frame(x, y), coords = c("x", "y"), crs = crs_in),
      4326
    )
  )
}

city_coords <- xy_to_lonlat(world_cities$X, world_cities$Y)

# get lat lon for cities (previously only had xy coords)
world_cities$lon <- city_coords[, 1]
world_cities$lat <- city_coords[, 2]


# convert cities to sf
cities_sf <- st_as_sf(
  world_cities,
  coords = c("lon", "lat"),
  crs = 4326,
  remove = FALSE
)


filter_cities_bbox <- function(df, lon_min, lon_max, lat_min, lat_max) {
  df[
    df$lon >= lon_min & df$lon <= lon_max &
    df$lat >= lat_min & df$lat <= lat_max,
  ]
}

#filter cities to our wanted region
cities_sf <- filter_cities_bbox(
  cities_sf,
  lon_min = 30, lon_max = 50,
  lat_min = -20, lat_max = 0
)


# get world map for plotting
world <- ne_countries(scale = "medium", returnclass = "sf")

# plot
plot(st_geometry(world), xlim = c(45, 47), ylim = c(-20, 0), col = "grey90", border = "grey60")
points(st_coordinates(cities_sf), pch = 20, cex = 0.6, col = "red")
```

```{r}
# load reef locations data
location_2009 <- readr::read_rds(here::here("processed_data", "clean_location_2009.rds"))


locations_sf <- st_as_sf(
  location_2009,
  coords = c("longitude", "latitude"),
  crs = 4326,
  remove = FALSE
)

locations_2010 <- locations_sf %>% dplyr::filter(date < as.Date("2013-01-01"))
locations_2015 <- locations_sf %>% dplyr::filter(date >= as.Date("2013-01-01"), date < as.Date("2017-01-01"))

assign_nearest_city <- function(locations_by_date){
  ## For each reef location, find the nearest city:
  
  locations_p <- st_transform(locations_by_date, 3857)
  cities_p   <- st_transform(cities_sf, 3857)
  
  #Find nearest city and update the columbs
  nearest_city_idx <- st_nearest_feature(locations_p, cities_p)
  locations_by_date$nearest_city <- cities_sf$CITY_NAME[nearest_city_idx]
  locations_by_date$nearest_city_lat <- cities_sf$lat[nearest_city_idx]
  locations_by_date$nearest_city_lon <- cities_sf$lon[nearest_city_idx]
  locations_by_date$nearest_city_pop <- 0
  locations_by_date$nearest_city_cost_distance <- 0
  
  # Use graph to calculate the travel cost to that nearest city
  locations_by_date$nearest_city_cost_distance <- mapply(
    cost_between_points,
    locations_by_date$lon,
    locations_by_date$lat,
    locations_by_date$nearest_city_lon,
    locations_by_date$nearest_city_lat, 
    MoreArgs = list(transition_graph = transition_graph, friction_raster = friction_raster)
  )
  
  locations_by_date
}

locations_2010_with_city <- assign_nearest_city(locations_2010)
locations_2015_with_city <- assign_nearest_city(locations_2015)

readr::write_rds(locations_2010_with_city,here::here("city_data","locations_with_city_2010.rds"))
readr::write_rds(locations_2015_with_city,here::here("city_data","locations_with_city_2015.rds"))


#write.csv(locations_sf, here::here("city_data","locations_sf.csv"))

```

defined as provincial capital cities, major population centres, landmark cities, national capitals, and ports - metropolitan area population

```{r}
# Load locations with closest cities
locations_with_city_2010 <- readr::read_rds(here::here("city_data", "locations_with_city_2010.rds"))
locations_with_city_2015 <- readr::read_rds(here::here("city_data", "locations_with_city_2015.rds"))


# Manually set 2010 pops - make sure that you justify 250,000 pop cut off
cities = c("Zanzibar","Tanga","Dar es Salaam","Mombasa","Kismaayo", "Moroni", "Nampula", "Toamasina")
pops_2010 =  c(460000,    260000,  3870000,        940000,    296000,   54000    , 539000,   246000)
pops_2015 =  c(569000,    300000,  5116000,       1103000,    375000,   70000    , 676000,   328000)

#Use pop data for 2010
locations_2010_with_city$nearest_city_pop <- pops_2010[match(locations_2010_with_city$nearest_city, cities)]

locations_2010_with_city$market_grav = locations_2010_with_city$nearest_city_pop / (locations_2010_with_city$nearest_city_cost_distance ** 2)

#Use pop data for 2015
locations_2015_with_city$nearest_city_pop <- pops_2015[match(locations_2015_with_city$nearest_city, cities)]

locations_2015_with_city$market_grav = locations_2015_with_city$nearest_city_pop / (locations_2015_with_city$nearest_city_cost_distance ** 2)

```

# Settlement Population
This is getting the sum of the population in a 25km radius

```{r}
# ── Settlement population metric ──────────────────────────────
# 25km radius — reflects typical maximum SSF fishing
# range in the WIO. 

# Load rasters
landscan_2010 <- rast(here::here("city_data", "landscan-global-2010.tif"))
landscan_2015 <- rast(here::here("city_data", "landscan-global-2015.tif"))

# Crop to AOI
landscan_crop_2010 <- crop(landscan_2010, aoi)
landscan_crop_2015 <- crop(landscan_2015, aoi)

# Convert sf -> SpatVector and reproject
points_2010 <- terra::project(
  terra::vect(locations_with_city_2010),
  crs(landscan_crop_2010)
)
points_2015 <- terra::project(
  terra::vect(locations_with_city_2015),
  crs(landscan_crop_2015)
)

# Function accepts buffer_m for flexible radius
calc_settlement_total_pop <- function(landscan_crop, points,
                                      locations_with_city,
                                      buffer_m = 25000) {
  buffers <- terra::buffer(points, width = buffer_m)

  pop_sum_df <- terra::extract(landscan_crop, buffers,
                               fun = \(x) sum(x, na.rm = TRUE))

  if (ncol(pop_sum_df) < 2)
    stop("extract() did not return an aggregated column.")

  pop_values <- pop_sum_df[[2]]

  if (length(pop_values) != nrow(locations_with_city))
    stop("Length mismatch: extract returned ", length(pop_values),
         " values but locations has ", nrow(locations_with_city), " rows.")

  locations_with_city$settlement_tot_pop <- pop_values
  return(locations_with_city)
}

# 25km radius
locations_with_settlement_2010 <- calc_settlement_total_pop(
  landscan_crop_2010, points_2010,
  locations_with_city_2010, buffer_m = 25000
)

locations_with_settlement_2015 <- calc_settlement_total_pop(
  landscan_crop_2015, points_2015,
  locations_with_city_2015, buffer_m = 25000
)

```

```{r}
#TEST
#Zoom into settlement just to see
#  	
# 43.26512
# -11.670252
aoi_zoom <- ext(43.18, 43.32, -11.75,-11.6) #slightly bigger area

landscan_zoomed <- crop(landscan_2010, aoi_zoom)
plot(landscan_zoomed)
max(landscan_zoomed)
```

## Settlement gravity metric 
Finds nearest LandScan cell with density ≥ 75 people/km²,extracts population within 
2km of that cell, and computes gravity as population / travel_cost_distance².
Threshold of 75 people/km² validated by Cinner et al. (2016) sensitivity analysis 
as the optimal peri-urban settlement threshold for reef fish biomass prediction.

```{r}
landscan_crop_2010 <- project(landscan_crop_2010, "EPSG:3857")
landscan_crop_2015 <- project(landscan_crop_2015, "EPSG:3857")

get_nearest75_cell <- function(locations_with_settlement, landscan_crop){
  r_crs <- crs(landscan_crop)
  
  
  # Initialise columns to NA
  locations_with_settlement$nearest_pop75_lon <- NA_real_
  locations_with_settlement$nearest_pop75_lat <- NA_real_
  locations_with_settlement$nearest_pop75_dist_km <- NA_real_
  
  # Try to find the nearest 75 pop cell by progressively increasing search radius
  threshold <- 75
  start_radius <- 5000    # 5 km
  step_radius  <- 5000    # increase by 5 km
  max_radius   <- 200000
  for (i in seq_len(nrow(locations_with_settlement))) {
  
    # --------------------------------------------------------------
    # Create point from lon/lat
    # --------------------------------------------------------------
    pt_ll <- vect(
      data.frame(
        x = locations_with_settlement$longitude[i],
        y = locations_with_settlement$latitude[i]
      ),
      geom = c("x", "y"),
      crs = "EPSG:4326"
    )
  
    # Project point to raster CRS (meters)
    pt_m <- project(pt_ll, r_crs)
  
    found  <- FALSE
    radius <- start_radius
  
    # --------------------------------------------------------------
    # Expand search radius until a cell is found
    # --------------------------------------------------------------
    while (!found && radius <= max_radius) {
  
      # Buffer in meters
      buf <- buffer(pt_m, width = radius)
  
      # Crop raster to buffer
      sub <- crop(landscan_crop, buf)
  
      if (!is.null(sub)) {
  
        # Binary mask of cells above threshold
        m <- sub > threshold
        # Safe check: are there any TRUE cells?
        if (!is.null(m) && ncell(m) > 0 && any(m[], na.rm = TRUE)) {
          #plot(m)
          # Coordinates of qualifying cells
          true_cells <- which(values(m))
          cell_xy <- xyFromCell(m, true_cells)
  
          # Euclidean distances (meters)
          d <- sqrt(
            (cell_xy[,1] - crds(pt_m)[1,1])^2 +
            (cell_xy[,2] - crds(pt_m)[1,2])^2
          )
  
          # Nearest qualifying cell
          j <- which.min(d)
  
          # Create nearest cell point (metric CRS)
          nearest_pt_m <- vect(
            data.frame(x = cell_xy[j,1], y = cell_xy[j,2]),
            geom = c("x", "y"),
            crs = r_crs
          )
  
          # Convert back to lon/lat for storage
          nearest_pt_ll <- project(nearest_pt_m, "EPSG:4326")
          coords <- crds(nearest_pt_ll)
  
          # Store results
          locations_with_settlement$nearest_pop75_lon[i]     <- coords[1,1]
          locations_with_settlement$nearest_pop75_lat[i]     <- coords[1,2]
          locations_with_settlement$nearest_pop75_dist_km[i] <- d[j] / 1000
  
          found <- TRUE
        }
      }
  
      # Explicit cleanup to avoid memory fragmentation
      rm(sub, m)
      gc()
  
      radius <- radius + step_radius
    }
  }
  
  return(locations_with_settlement)
}

locations_with_grav_2010 <- get_nearest75_cell(locations_with_settlement_2010, landscan_crop_2010)
locations_with_grav_2015 <- get_nearest75_cell(locations_with_settlement_2015, landscan_crop_2015)
```

```{r}
#Find population in a 2km radius of the closest 75 pop density cell

calc_closest_settlement_pop <- function(locations_with_grav, landscan_crop){
  landscan_m <- project(landscan_crop, "EPSG:54009")   # World Mollweide
  
  df <- sf::st_drop_geometry(locations_with_grav)
  
  points <- terra::vect(
    df,
    geom = c("nearest_pop75_lon", "nearest_pop75_lat"),
    crs = "EPSG:4326"
  )
  
  points_m   <- project(points, crs(landscan_m))
  
  
  buffers <- buffer(points_m, width = 2000)
  
  # Extract population
  pop_sum <- extract(
    landscan_m,
    buffers,
    fun = \(x) sum(x, na.rm = TRUE)
  )
  
locations_with_grav$nearest_settlement_pop <- pop_sum[[2]]
return(locations_with_grav)
}

locations_with_all_2010 <- calc_closest_settlement_pop(locations_with_grav_2010, landscan_crop_2010)
locations_with_all_2015 <- calc_closest_settlement_pop(locations_with_grav_2015, landscan_crop_2015)

```

```{r}

 calc_settlement_grav <- function(locations_with_grav){
  tmp <- mapply(cost_between_points,
                locations_with_grav$longitude,
                locations_with_grav$latitude,
                locations_with_grav$nearest_pop75_lon,
                locations_with_grav$nearest_pop75_lat,
                MoreArgs = list(transition_graph = transition_graph,
                                friction_raster  = friction_raster),
                SIMPLIFY = TRUE, USE.NAMES = FALSE)

  if (length(tmp) != nrow(locations_with_grav)) stop("distance vector length mismatch: ", length(tmp))
  locations_with_grav$nearest_pop_cost_distance <- tmp

  # Set a minimum value of 0.005 (as some are 0)
  locations_with_grav$nearest_pop_cost_distance <- pmax(locations_with_grav$nearest_pop_cost_distance, 0.005)

  locations_with_grav$nearest_pop75_grav = locations_with_grav$nearest_settlement_pop / (locations_with_grav$nearest_pop_cost_distance ^ 2)

  return(locations_with_grav)
}

```

```{r}
readr::write_rds(locations_with_grav_2010_final,here::here("city_data","locations_with_grav_2010_final.rds"))
readr::write_rds(locations_with_grav_2015_final,here::here("city_data","locations_with_grav_2015_final.rds"))

#combine different years back into one
locations_with_grav_combined <- dplyr::bind_rows(locations_with_grav_2010_final, locations_with_grav_2015_final)

#Save the final product
readr::write_rds(locations_with_grav_combined,here::here("city_data","locations_with_grav_combined.rds"))
```

#end
